---
title: "Causality and Potential Outcomes"
subtitle: "BUSS975: Causal Inference in Financial Research"
author: "Professor Ji-Woong Chung"
institute: "Korea University"
format:
  beamer:
    aspectratio: 169
    fontsize: 11pt
    slide-level: 2
    toc: false
    pdf-engine: pdflatex
    pdf-engine-opts:
      - "-interaction=nonstopmode"
    keep-tex: true
    include-in-header:
      text: |
        \setbeamertemplate{navigation symbols}{}
        \setbeamertemplate{footline}[frame number]
        \setbeamertemplate{itemize items}[circle]
        \setbeamertemplate{itemize subitem}[triangle]
        \setbeamercolor{frametitle}{fg=black}
        \setbeamerfont{frametitle}{series=\bfseries}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{booktabs}
        % shrink code input (fancyvrb) and code output (verbatim)
        \usepackage{fancyvrb}
        \fvset{fontsize=\scriptsize}
        \makeatletter
        \def\verbatim@font{\scriptsize\ttfamily}
        \makeatother
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

## Outline

1. What causality is --- and isn't
2. The potential outcomes framework
3. Selection bias: the Perfect Regulator
4. Randomization and independence
5. SUTVA
6. The regression bridge
7. Randomization inference

# 1. What Causality Is --- and Isn't

## Motivation

- As researchers, we want to make **causal** statements:
  - What is the effect of a corporate tax change on firms' leverage?
  - What is the effect of CEO equity ownership on risk-taking?
  - What is the effect of a disclosure mandate on the cost of capital?
- We are rarely satisfied with saying variables are merely "associated."
- But nearly all of our data is **observational**: nobody randomized leverage, ownership, or regulation for us.
- This course is about when --- and how --- observational data can answer causal questions.

## Three goals of empirical analysis

1. **Description** --- how things are or were.
   - Has industry concentration increased over time?
2. **Prediction** --- guessing an unknown value, without intervening.
   - What will this firm's earnings be next quarter?
3. **Causality** --- how *changing* one variable would change another, all else equal.
   - How would a disclosure mandate change firms' cost of capital?

\vspace{0.5em}

- All three are legitimate. But they require different tools, and confusing them is the original sin of empirical work.

## Fallacy 1: prediction is not causation

- Momentum: past returns *predict* future returns.
  - Does buying past winners *cause* them to keep winning? No --- the predictive relationship says nothing about intervention.
- A model can predict superbly while being causally empty.
  - This is the promise (and the limit) of much of machine learning.
- **Causal inference asks a different question**: what would happen if we *changed* something?
  - Prediction asks what $Y$ **is**, among the units where $X = x$.
  - Causation asks what $Y$ **would be**, if we set $X = x$.
<!-- - These are two different mathematical objects --- not two emphases on the same one.-->

<!--
## Conditioning versus intervening

- Two operations that look alike and are not:

$$
\underbrace{E[Y \mid X = x]}_{\text{conditioning}}
\qquad \text{versus} \qquad
\underbrace{E[Y \mid \mathrm{do}(X = x)]}_{\text{intervening}}
$$

- **Conditioning** --- *filter the data*. "Among the units that happen to have $X = x$, what is the average $Y$?"
  - The population **changes**: we keep a subpopulation, and it may be special in ways we cannot see.
- **Intervening** --- *reach in and set*. "If we assigned $X = x$ to every unit, what would the average $Y$ be?"
  - The population **stays fixed**: only $X$ moves.
- Prediction needs only the first. Causal inference needs the second.
- $\mathrm{do}(\cdot)$ is Pearl's notation; we formalize it next lecture, with DAGs.

## Ratings predict default; ratings do not prevent it

- Let $Y = 1$ if a firm defaults within five years, and let $X$ be its credit rating.
- **Conditioning**: $E[Y \mid X = \text{AAA}]$ is very low --- AAA issuers rarely default.
  - Ratings are superb *predictors*. Selling this number is a profitable business.
- **Intervening**: $E[Y \mid \mathrm{do}(X = \text{AAA})]$ --- a regulator relabels every junk issuer AAA.
  - Default rates barely move. The label is not the cash flows.
- Conditioning **selects** firms that were already creditworthy; intervening **changes the label** and leaves creditworthiness untouched.
- The causal effect is not exactly zero --- ratings do move borrowing costs, covenant triggers, and mandate-constrained demand --- but nothing like the AAA-to-junk gap.
- Ask of any empirical result: *is this a filter, or an intervention?*
-->

## Fallacy 2: temporal ordering is not causation

- *Post hoc ergo propter hoc*: "after this, therefore because of this."
- The rooster crows, then the sun rises.
- Finance version: a struggling firm fires its CEO, then performance recovers.
  - Did the new CEO cause the recovery?
  - Or is this **mean reversion** --- firms replace CEOs at the trough?
- Timing evidence is useful (we will use it in event studies and DiD), but timing alone proves nothing.

## Fallacy 3: no correlation does not mean no causation

- A sailor moves the rudder constantly, yet the boat sails straight.
  - Correlation between rudder and course: $\approx 0$. Causal effect: enormous.
  - Why? The rudder *responds optimally* to the wind.
- Finance is full of optimizing agents doing exactly this:
  - Firms hedge **because** they are exposed: hedging may show zero correlation with distress even if it strongly prevents distress.
  - The Fed adjusts rates to offset shocks: policy can look uncorrelated with output.
- **Lesson**: when treatment is chosen optimally, causal effects can be *invisible* in correlations.

## The common thread: the assignment mechanism

- In all three fallacies, what breaks the link between correlation and causation is **how treatment got assigned**:
  - Who chose to be treated? Why? Based on what?
- The central message of this lecture (and this course):

\begin{center}
\emph{Correlations are properties of the data.\\
Causal effects are properties of the data \textbf{plus} the assignment mechanism.}
\end{center}

- To formalize this, we need a language for "what would have happened otherwise."

## Defining causality: counterfactuals

- Our (narrow) definition: *ceteris paribus*, how does $D$ affect $Y$?
- To know the effect of $D$ on unit $i$, we need to compare:
  - $Y_i$ **with** the treatment, and
  - $Y_i$ **without** the treatment,
  - with everything else the same.
- One of these is always hypothetical --- a **counterfactual**.
- **The Fundamental Problem of Causal Inference** (Holland 1986): we can never observe both states for the same unit at the same time.

# 2. The Potential Outcomes Framework

## States of the world

- Imagine each unit exists in two potential states of the world, identical in every respect except one: treatment.
- Binary treatment indicator:
$$
D_i = \begin{cases} 1 & \text{if unit } i \text{ is treated} \\ 0 & \text{otherwise} \end{cases}
$$
- Two **potential outcomes** for every unit:
  - $Y_i^1$: the outcome unit $i$ *would* have if treated,
  - $Y_i^0$: the outcome unit $i$ *would* have if untreated.
- Both are defined for every unit --- treated or not. That is the trick.

## The switching equation

- What we *observe* is only one of the two:
$$
Y_i = D_i\, Y_i^1 + (1 - D_i)\, Y_i^0
$$
- Treatment "switches on" which potential outcome is realized.
- Notice what this equation does **not** contain: no error term, no functional form, no linearity. It is a definition, not a model.

## Individual treatment effects

- The causal effect of the treatment on unit $i$:
$$
\delta_i = Y_i^1 - Y_i^0
$$
-  $\delta_i$ carries an $i$ subscript.
  - Effects are allowed to **differ across units** from the start.
- The Fundamental Problem, restated: $\delta_i$ is never observed for any unit, because one of its two ingredients is always missing.
  - Causal inference is a **missing data problem** --- but the missing value never existed to begin with.

## A running example: capital injections

- Setting: a banking crisis. The government can inject capital into banks (think TARP, 2008).
- Outcome $Y$: bank lending growth (%) over the following year.
- $Y_i^1$: bank $i$'s lending growth *with* an injection.
- $Y_i^0$: bank $i$'s lending growth *without* one.
- Question: what is the causal effect of capital injections on lending?

## Ten banks, two potential outcomes

| Bank | $Y^1$ | $Y^0$ | $\delta$ |
|:----:|:-----:|:-----:|:--------:|
| 1    | 9     | 2     | 7        |
| 2    | 6     | 1     | 5        |
| 3    | 7     | 3     | 4        |
| 4    | 5     | 2     | 3        |
| 5    | 6     | 4     | 2        |
| 6    | 7     | 6     | 1        |
| 7    | 7     | 7     | 0        |
| 8    | 7     | 8     | $-1$     |
| 9    | 7     | 9     | $-2$     |
| 10   | 9     | 8     | 1        |

- Weak banks (low $Y^0$) gain a lot; healthy banks gain little --- or are even hurt.

## Average treatment effects

- **Average Treatment Effect (ATE)**:
$$
ATE = E[\delta_i] = E[Y_i^1] - E[Y_i^0]
$$
- In our table: $ATE = \frac{7+5+4+3+2+1+0-1-2+1}{10} = 2.0$
- On average, injections raise lending growth by 2 percentage points.
- But note the heterogeneity: effects range from $+7$ to $-2$.
  - An average can hide winners and losers. 

## ATT and ATU: different questions, different answers

- **Average Treatment effect on the Treated**:
$$
ATT = E[\delta_i \mid D_i = 1]
$$
- **Average Treatment effect on the Untreated**:
$$
ATU = E[\delta_i \mid D_i = 0]
$$
- These are **different policy questions**:
  - ATT: "Did TARP help the banks that received it?"
  - ATE: "Would injections help a randomly chosen bank?"
  - ATU: "Would extending the program to the remaining banks help them?"
- When effects are heterogeneous and assignment is not random, $ATT \neq ATE \neq ATU$ --- and each estimator you will meet in this course targets a *particular* one of these.

## What we actually observe

| Bank | $D$ | $Y^1$ | $Y^0$ | $\delta$ |
|:----:|:---:|:-----:|:-----:|:--------:|
| 1    | 1   | 9     | ?     | ?        |
| 2    | 1   | 6     | ?     | ?        |
| 3    | 1   | 7     | ?     | ?        |
| 4    | 1   | 5     | ?     | ?        |
| 5    | 1   | 6     | ?     | ?        |
| 6    | 0   | ?     | 6     | ?        |
| 7    | 0   | ?     | 7     | ?        |
| 8    | 0   | ?     | 8     | ?        |
| 9    | 0   | ?     | 9     | ?        |
| 10   | 0   | ?     | 8     | ?        |

- Half of the table is missing --- forever. Everything we do from here on is a strategy for filling in the question marks *on average*.

# 3. Selection Bias: The Perfect Regulator

## The naive comparison

- With observational data, the natural move is to compare treated and untreated:
$$
SDO = E[Y_i^1 \mid D_i = 1] - E[Y_i^0 \mid D_i = 0]
$$
- **Simple Difference in Outcomes** (SDO): mean outcome of the treated minus mean outcome of the untreated.
- Every cross-sectional regression of $Y$ on $D$ is, at heart, an SDO.
- Question: when does SDO recover the ATE?

## The Perfect Regulator

- Suppose the regulator has perfect foresight: it knows every bank's $\delta_i$ and injects capital into the five banks that **gain the most** (banks 1--5).
- This is not a villain --- it is an *optimizing* regulator, allocating scarce capital where it helps most.
  - Exactly what a good policymaker (or a good CEO, or a good doctor) should do.
- Treated: banks 1--5. Untreated: banks 6--10.
- Now compute the SDO from the observable data.

## The Perfect Regulator's data

- Mean outcome among treated: $E[Y^1 \mid D=1] = \frac{9+6+7+5+6}{5} = 6.6$
- Mean outcome among untreated: $E[Y^0 \mid D=0] = \frac{6+7+8+9+8}{5} = 7.6$
$$
SDO = 6.6 - 7.6 = -1.0
$$
- The truth: $ATE = +2.0$, and $ATT = \frac{7+5+4+3+2}{5} = +4.2$.
- The naive comparison gets the **sign wrong**: injections look *harmful* precisely because they were allocated *well*.
- This is the TARP debate in one table: "bailed-out banks lent less than healthy banks" is an SDO, not a causal effect.

## The decomposition

- The SDO can always be decomposed --- this is an identity, not an assumption ($\pi$ = share of units treated):
$$
\begin{aligned}
\underbrace{E[Y^1 \mid D{=}1] - E[Y^0 \mid D{=}0]}_{\text{SDO}}
  = ATE &+ \underbrace{E[Y^0 \mid D{=}1] - E[Y^0 \mid D{=}0]}_{\text{selection bias}} \\
  &+ \underbrace{(1-\pi)\,(ATT - ATU)}_{\text{heterogeneous TE bias}}
\end{aligned}
$$

- **Selection bias**: treated and untreated differ in their *untreated* outcomes.
- **Heterogeneous treatment effect bias**: the treated gain differently than the untreated would.
- Memorize this equation. Every research design in this course is an attempt to kill these two terms.

## Proving the decomposition

Let $\pi = P(D{=}1)$. 

**Step 1.** Add and subtract $E[Y^0 \mid D{=}1]$ --- the *counterfactual* mean for the treated:
$$
\begin{aligned}
SDO &= E[Y^1 \mid D{=}1] - E[Y^0 \mid D{=}0] \\
    &= \underbrace{E[Y^1 \mid D{=}1] - E[Y^0 \mid D{=}1]}_{ATT} \;+\;
       \underbrace{E[Y^0 \mid D{=}1] - E[Y^0 \mid D{=}0]}_{\text{selection bias}}
\end{aligned}
$$

**Step 2.** By the law of total expectation, $ATE = \pi\,ATT + (1-\pi)\,ATU$. Subtract $ATE$ from $ATT$:
$$
ATT - ATE \;=\; (1-\pi)\,ATT - (1-\pi)\,ATU \;=\; (1-\pi)(ATT - ATU)
$$

Substituting $ATT = ATE + (1-\pi)(ATT-ATU)$ into Step 1 gives the three-term decomposition. $\blacksquare$

## The decomposition, in our example

- Selection bias:
\small
$$
E[Y^0 \mid D{=}1] - E[Y^0 \mid D{=}0] = \frac{2+1+3+2+4}{5} - \frac{6+7+8+9+8}{5} = 2.4 - 7.6 = -5.2
$$
\normalsize
  - Treated banks were *much weaker* to begin with.
- Heterogeneous TE bias, with $\pi = 0.5$:
$$
(1 - 0.5)(ATT - ATU) = 0.5 \times (4.2 - (-0.2)) = +2.2
$$
- Check: $\;2.0 - 5.2 + 2.2 = -1.0 = SDO$ \checkmark
- Both bias terms are large; they partially offset, and the net result is a sign flip.

## Selection on gains is endemic in finance

- The Perfect Doctor (Cunningham's version) treats patients who benefit most. Finance is full of Perfect Doctors:
  - Regulators bail out the banks where intervention matters most.
  - Boards grant equity to the CEOs they expect to respond.
  - Firms go public when going public helps them most.
  - Banks lend to the borrowers they screen as creditworthy.
- Agents **optimize**. Optimization means treatment is assigned on expected gains --- the worst case for naive comparisons.
- This is why finance cannot simply borrow "big data" logic: more observations do not fix a biased assignment mechanism.

## Simulation: triage vs. coin flip

```{r}
#| echo: true
set.seed(975)
n    <- 100000
y0   <- rnorm(n, 7, 2)     # lending growth, no injection
gain <- rnorm(n, 2, 1)     # true individual gain (so ATE = 2)
y1   <- y0 + gain

D1 <- as.integer(y0 < quantile(y0, 0.30))  # triage: weakest 30%
D2 <- rbinom(n, 1, 0.30)                   # coin flip: random 30%

sdo <- function(D) mean(y1[D == 1]) - mean(y0[D == 0])
round(c(ATE = mean(gain), triage = sdo(D1), coin = sdo(D2)), 2)
```

## What the simulation tells us

- The true effect is $+2$ in both worlds --- same banks, same gains, only the **assignment rule** differs.
- Triage (treat the weakest): SDO is strongly *negative*. An observer concludes bailouts destroy lending.
- Coin flip: SDO $\approx$ ATE. Same data-generating process, correct answer.
- With $n = 100{,}000$: the bias does not shrink with sample size. **Selection bias is not a small-sample problem.**

# 4. Randomization and Independence

## The independence assumption

- What made the coin flip work? Treatment assignment was **independent of potential outcomes**:
$$
(Y^1, Y^0) \indep D
$$
- In words: knowing a bank's treatment status tells you *nothing* about its potential outcomes --- neither its baseline $Y^0$ nor its gain $\delta$.
- Under independence:
  - $E[Y^0 \mid D{=}1] = E[Y^0 \mid D{=}0]$ $\Rightarrow$ **selection bias $= 0$**
  - $E[\delta \mid D{=}1] = E[\delta \mid D{=}0]$ $\Rightarrow$ $ATT = ATU = ATE$ $\Rightarrow$ **heterogeneity bias $= 0$**
- Both terms die, and $SDO = ATE$. Randomization is the assignment mechanism that *guarantees* independence.

<!--
- Equivalently: randomization is what makes **conditioning equal intervening**,
$$
E[Y \mid D = d] \;=\; E[Y \mid \mathrm{do}(D = d)].
$$
-->

## What randomization balances

- Physical randomization balances treated and control groups on:
  - observable characteristics (size, leverage, ratings, ...),
  - **unobservable** characteristics (management quality, risk appetite, ...),
  - potential outcomes themselves, and
  - treatment gains $\delta_i$.
- No control variables needed. No model needed. The balance is created by the *mechanism*, not by the econometrician.
- Practice: in any experiment (or natural experiment), the first table is a **balance table** --- compare pre-treatment covariates across groups. Balance on observables is the testable *symptom* of a sound mechanism.

## Case: eBay and the value of paid search

- **The setting.** In paid search advertising, a firm bids so that its ad appears when a user searches a keyword. eBay spent heavily on this.
- Two kinds of keyword:
  - **Brand** keywords contain the firm's own name --- "eBay typewriter."
  - **Non-brand** keywords do not --- "used typewriter."
- **How the industry measured returns.** *Attribution*: if a user clicks the ad and then buys, credit the sale to the ad. By this arithmetic, paid search looked extraordinarily profitable.
- **The problem.** Somebody typing "eBay typewriter" was already going to eBay. Take the ad away and they click the organic eBay link one line below.
- Attribution **conditions on clicking**. It never intervenes. It is an SDO, not an $ATE$.

## eBay: what the experiments found

- Blake, Nosko, and Tadelis (2015, *Econometrica*) ran large-scale field experiments at eBay, switching paid search off in randomly selected markets.
- **Brand keywords**: **no measurable short-run benefit.** Traffic substituted almost entirely to the free organic link. eBay was paying for clicks it would have received anyway.
- **Non-brand keywords**: ads *did* work --- but only for **new and infrequent** users.
  - Frequent users were unaffected, yet accounted for most of the advertising spend.
  - Net of costs, **average returns were negative**.
- Experimental returns were "a fraction of conventional non-experimental estimates."
- **Two lessons.** Attribution measures selection, not causation. And the average effect concealed enormous heterogeneity --- precisely the variation in $\delta_i$ we defined earlier.

## A genuine finance RCT: Karlan and Zinman (2009)

- *"Observing Unobservables: Identifying Information Asymmetries with a Consumer Credit Field Experiment,"* **Econometrica** 77.
- **Setting.** A major South African consumer lender mailed loan offers to **58,000** former clients.
- **Three independent randomizations:**
  1. The **offer rate** printed on the direct-mail solicitation.
  2. A **contract rate**, revealed only *after* the client accepted the offer --- sometimes a surprise discount.
  3. A **dynamic repayment incentive**, also a surprise: preferential pricing on future loans for borrowers who stayed in good standing.
- The surprises carry the design. Anything revealed *after* acceptance cannot have influenced **who** accepted.

## Karlan and Zinman: separating two unobservables

- Credit markets have two classic private-information problems, and observational data cannot tell them apart:
  - **Adverse selection** (hidden *type*): high rates attract intrinsically worse borrowers.
  - **Moral hazard** (hidden *action*): high rates weaken the incentive to repay.
- Both predict the same correlation --- higher rate, more default. No amount of loan data separates them.
- The experiment does:
  - Vary the **offer** rate $\Rightarrow$ changes *who selects in* $\Rightarrow$ identifies **selection**.
  - Hold the offer rate fixed and vary the **contract** rate $\Rightarrow$ same borrowers, different incentives $\Rightarrow$ identifies **moral hazard**.
- **Findings**: strong evidence of moral hazard, weaker evidence of hidden information; roughly **13--21%** of default attributable to moral hazard.
- Randomization used not to estimate one number, but to **decompose a mechanism**.

## Other field experiments in finance

- **Cole, Giné, and Vickery (2017, *RFS*)** --- rainfall insurance in India. Randomly insured farmers shifted into **higher-return, rainfall-sensitive cash crops**, *before* rainfall was realized.
- **Bertrand et al. (2010, *QJE*)** --- direct-mail loan offers in South Africa, randomizing price *and* content. A **photo of an attractive woman** raised demand as much as a **25% rate cut**.
- **Duflo and Saez (2003, *QJE*)** --- \$20 to attend a benefits fair raised enrollment, and raised it too for **untreated colleagues in treated departments**.
- That last one is a clean randomization that still violates SUTVA (Part 5).


## Where finance experiments live

- Notice what these studies have in common: the researcher controls **an offer**.
  - A mailer, a price quote, an insurance policy, an invitation.
- That is why field experiments cluster in **household finance, credit, and insurance** --- markets with many small units and a lender or insurer willing to randomize.


## Why finance rarely gets to randomize

- Nobody will randomize leverage, M&A, CEO pay, bankruptcy, or bank regulation:
  - stakes too high, units too big, ethics, general-equilibrium effects.
- So corporate finance and asset pricing live almost entirely in observational data.
- **The rest of this course is a toolkit for *earning* independence without a coin flip:**
  - Unconfoundedness / matching: independence *conditional on observables*.
  - RDD: independence *near a threshold*.
  - IV: independence *of an instrument*.
  - DiD / panel: independence *of timing, in changes*.
- Every method = an argument that some comparison mimics the coin flip. Your job as a referee is to ask: *whose coin, and who flipped it?*

# 5. SUTVA

## The fine print under the potential outcomes

- Writing $Y_i^1, Y_i^0$ at all presumes the **Stable Unit Treatment Value Assumption**:

1. **No interference**: unit $i$'s potential outcomes do not depend on *other units'* treatment status.
2. **No hidden versions of treatment**: "treated" means the same thing for every unit --- one treatment, one dose.

- If SUTVA fails, $Y_i^1$ is not even well-defined: bank $i$'s outcome under treatment depends on *who else* was treated.

## Interference is the norm in finance

- Units in finance **compete and interact**. Examples:
  - **Peer effects in capital structure** (Leary and Roberts 2014): firm $i$'s leverage responds to peers' leverage. Treating one firm treats its rivals.
  - **Bailouts and competition**: a capital injection into bank $i$ lets it bid for deposits and borrowers --- *taking them from control banks*.
  - **Banking DiDs**: treated banks' customers migrate to control banks; the control group's outcome moves *because of* the treatment.
- If treatment hurts controls, the treated-control gap **overstates** the true effect; if it helps them (spillovers), it understates.

## "The control group is treated too"

- In a competitive industry, there may be no untreated state of the world to observe:
  - Treatment effects estimated off treated-vs-control gaps are **relative** (partial-equilibrium) effects.
  - The **aggregate** (general-equilibrium) effect can be smaller, zero, or of opposite sign.
- Example: a subsidy to treated firms raises their market share. Industry output may be unchanged --- pure reallocation. The DiD shows a large "effect."
- Always ask: *what happens to the market, not just the treated units?*
- We will return to this in DiD (choosing clean controls) and in the reflection problem (peer effects).

## Hidden versions of treatment

- SUTVA also requires the treatment to be *one thing*.
- Finance treatments are rarely uniform in dose:
  - "Having a credit rating" spans AAA to CCC.
  - "Being bailed out" ranged from \$1bn to \$45bn in TARP.
  - "Adopting an anti-takeover provision" bundles very different provisions.
- If doses differ and units select their dose, the single $\delta_i$ hides a dose-response function --- and the estimand becomes an uninterpretable mixture.

<!--
Discuss estimator vs. estimand vs. parameter here:
Estimator: a function of sample (random) (sum of Y_i for D=1)
Estimand: a feature of population (E[Y|D=1])
Parameter: a feature of population we're interested in (ATE - we can't observe even in the population). Requires identifying assumptions

Estimator -> Estimand: Inference
Estimand -> Parameter: Identification
-->

## What to do about varying dose

- **Define** the treatment precisely. What exactly is the "1" in $D_i = 1$?
- **Report** the *distribution* of dose, not just a count of treated units.
- If doses vary widely, do **not** collapse them into a binary. Instead, one of:
  - use the dose itself as the treatment and estimate a **dose-response function**, $E[Y \mid \text{dose}]$;
  - restrict the sample to a narrow dose band, so that "treated" means one thing;
  - interact the treatment dummy with dose, and report the effect at several dose levels.
- Whichever you choose, state which estimand it delivers:

\begin{center}
\emph{``the effect of a \$1bn injection'' is not ``the effect of being bailed out.''}
\end{center}

# 6. The Regression Bridge

## Same problem, older dialect

- You already know a language for causality --- the regression language:
$$
y = \beta_0 + \beta_1 x + u
$$
- Causal interpretation of $\beta_1$ requires **conditional mean independence**:
$$
E[u \mid x] = E[u] = 0
$$
- The potential outcomes framework and the CMI framework describe the **same problem** in two dialects. Let's build the dictionary.

## The dictionary

| Potential outcomes dialect | Regression dialect |
|:---------------------------------------------|:--------------------------------|
| $(Y^1, Y^0) \indep D$ | $E[u \mid x] = E[u]$ (CMI) |
| Selection bias | $\mathrm{cov}(x, u) \neq 0$ (OVB, simultaneity) |
| Heterogeneous $\delta_i$ | Random coefficients $b_i$ |
| ATE | Average partial effect (APE) |
| SDO | OLS coefficient of $Y$ on $D$ |

- The PO dialect is sharper about *estimands* (ATE vs. ATT) and *assignment*.
- The regression dialect is sharper about *direction of bias* --- and it is the language referees speak. You need both.

## Omitted variable bias

- You estimate $y = \beta_0 + \beta_1 x + u$, but the true model is
$$
y = \beta_0 + \beta_1 x + \beta_2 z + v
$$
- Then
$$
\hat{\beta}_1 \xrightarrow{p} \beta_1 + \underbrace{\frac{\mathrm{cov}(x, z)}{\mathrm{var}(x)}}_{\delta_{xz}} \beta_2
$$
- The bias is the *effect of the omitted variable* ($\beta_2$) times *its regression on the included one* ($\delta_{xz}$).
- Consistent only if $\mathrm{cov}(x,z) = 0$ or $\beta_2 = 0$: the omitted variable must matter **and** be correlated with $x$.

## Signing the bias

$$
\text{bias} = \delta_{xz} \, \beta_2
$$

| | $\mathrm{cov}(x,z) > 0$ | $\mathrm{cov}(x,z) < 0$ |
|:---------------|:-----------:|:-----------:|
| $\beta_2 > 0$ | $+$ | $-$ |
| $\beta_2 < 0$ | $-$ | $+$ |

- Classic example: $\ln(wage)$ on education, omitting ability.
  - Ability raises wages ($\beta_2 > 0$) and correlates positively with education $\Rightarrow$ upward bias.
- Finance example: firm value on governance, omitting CEO talent.
  - Talented CEOs may sort into well-governed firms and raise value $\Rightarrow$ governance coefficient too big.
- With **multiple** omitted variables (or multiple included regressors), signs interact and the algebra stops being your friend. Sign one bias at a time, or don't sign at all.

<!--

## From switching equation to regression

- Start from the switching equation and factor out $Y_i^0$:
$$
Y_i \;=\; D_i Y_i^1 + (1-D_i) Y_i^0
    \;=\; Y_i^0 + D_i \underbrace{\left(Y_i^1 - Y_i^0\right)}_{\delta_i}
    \;=\; Y_i^0 + D_i \delta_i
$$
- Now split each random quantity into its mean plus a deviation, writing $\delta = E[\delta_i] = ATE$:
$$
Y_i^0 = E[Y^0] + \left(Y_i^0 - E[Y^0]\right), \qquad \delta_i = \delta + (\delta_i - \delta)
$$
- Substitute both into $Y_i = Y_i^0 + D_i\delta_i$ and collect terms:
$$
Y_i \;=\; E[Y^0] \;+\; \delta D_i \;+\; \left(Y_i^0 - E[Y^0]\right) + D_i(\delta_i - \delta)
$$
- This **is** a regression of $Y$ on a constant and $D$ --- exactly, with no approximation and no assumptions. We only rewrote the switching equation.

-->

## The omitted variable is $Y^0$

- The switching equation factors as $Y_i = Y_i^0 + D_i\delta_i$. Split each term into its mean plus a deviation, writing $\delta = E[\delta_i] = ATE$:
$$
Y_i \;=\; \underbrace{E[Y^0]}_{\beta_0}
      \;+\; \underbrace{\delta}_{\beta_1} D_i
      \;+\; \underbrace{\left(Y_i^0 - E[Y^0]\right)}_{\text{baseline}}
      \;+\; \underbrace{D_i(\delta_i - \delta)}_{\text{gain}}
$$
- That **is** a regression of $Y$ on a constant and $D$ --- exactly, with no approximation and no assumptions. The last two terms are the error $u_i$.
- So the "omitted variable" is the **baseline potential outcome** $Y_i^0$ (together with the individual gain $\delta_i$).
- OVB --- in causal settings, it *is* selection bias, written in the older dialect.
- Careful: $Y_i^0$ is **observed** for the untreated. It is counterfactual only for the treated --- missing exactly where we would need it. And no better data collection can fix that.


## Two ways treatment can be selected

- $D$ can be correlated with **either** piece of the error:
$$
u_i = \underbrace{\left(Y_i^0 - E[Y^0]\right)}_{\text{baseline}} \;+\; \underbrace{D_i(\delta_i - \delta)}_{\text{gain}}
$$
- **Selection on levels** --- treatment goes to units with unusual $Y_i^0$: how they would have done *anyway*, untreated.
  - Bailing out the weakest banks. Training the least employable workers.
- **Selection on gains** --- treatment goes to units with unusual $\delta_i$: how much they would *benefit*.
  - The Perfect Regulator. Any optimizing agent allocating a scarce treatment.
- The two are distinct, and **either one alone** breaks the naive comparison.
- Note our own running example sorts on **gains**, not levels. Defining selection bias as "$\mathrm{cov}(D,Y^0) \neq 0$" would misclassify it.

## Both biases, from one lemma

\small
- **Lemma.** For *binary* $D$ with $\pi = P(D{=}1)$, and any random variable $W$:
$$
\mathrm{cov}(D, W) = \pi(1-\pi)\Big(E[W \mid D{=}1] - E[W \mid D{=}0]\Big)
$$
  - *Proof.* $E[DW] = \pi E[W|D{=}1]$ since $D$ is binary, and $E[W] = \pi E[W|D{=}1] + (1-\pi)E[W|D{=}0]$. Substitute into $\mathrm{cov}(D,W) = E[DW] - E[D]E[W]$ and collect. $\blacksquare$
- **Take $W = Y^0$:**
$$
\mathrm{cov}(D, Y^0) = \pi(1-\pi)\Big(E[Y^0|D{=}1] - E[Y^0|D{=}0]\Big)
$$
- **Take $W = D_i(\delta_i - \delta)$.** It is $0$ whenever $D=0$, so $E[W|D{=}0] = 0$, while $E[W|D{=}1] = ATT - ATE$:
$$
\mathrm{cov}\big(D,\, D_i(\delta_i-\delta)\big) = \pi(1-\pi)\big(ATT - ATE\big)
$$
- Since $\mathrm{var}(D) = \pi(1-\pi)$, dividing each by $\mathrm{var}(D)$ **cancels that factor** --- leaving the selection bias term and $ATT - ATE$.
\normalsize

## The two dialects give the same equation

- Since $\hat\beta_1 \xrightarrow{p} \delta + \mathrm{cov}(D,u)/\mathrm{var}(D)$, and using $ATT - ATE = (1-\pi)(ATT-ATU)$ from earlier:
$$
\hat\beta_1 \;\xrightarrow{p}\;
\underbrace{ATE}_{\delta}
\;+\; \underbrace{\Big(E[Y^0|D{=}1]-E[Y^0|D{=}0]\Big)}_{\text{selection bias}}
\;+\; \underbrace{(1-\pi)(ATT-ATU)}_{\text{heterogeneity bias}}
$$
- The right-hand side is exactly the $SDO$ decomposition we proved in Part 3.
- The regression route and the potential-outcomes route are not analogies. **They are the same three terms.**
- So $\hat\beta_1$ is inconsistent for the $ATE$ --- but perfectly *consistent* for the $SDO$. Once again: OLS delivers the estimand you asked for, not the one you wanted.


## Simultaneity: when $y$ fires back

- Suppose the regressor responds to the outcome:
$$
y = \beta_0 + \beta_1 x + u, \qquad x = \alpha_0 + \alpha_1 y + \varepsilon
$$
- Solve for $x$:
$$
x = \frac{\alpha_0 + \alpha_1 \beta_0}{1 - \alpha_1 \beta_1} + \frac{\alpha_1 u + \varepsilon}{1 - \alpha_1 \beta_1}
$$
- $x$ is mechanically correlated with $u$: CMI fails, OLS is biased --- **reverse causality**.
- Finance examples: leverage and performance; board independence and performance; analyst coverage and returns. In equilibrium, almost every firm characteristic responds to almost every outcome.
- Note: lagging $x$ does **not** fix this if $x$ is persistent and anticipates $y$. It hides the problem; it does not solve it.

## Random coefficients: the formal bridge

- Let every firm have its own line:
$$
y_i = a_i + b_i x_i, \qquad a_i = \alpha + c_i, \quad b_i = \beta + d_i
$$
- With binary $x$, this **is** the potential outcomes model: $b_i = \delta_i$, $a_i = Y_i^0$.
- Rewrite: $y_i = \alpha + \beta x_i + \underbrace{(c_i + d_i x_i)}_{u_i}$
- OLS identifies the **average partial effect** $\beta = E[b_i]$ (the ATE!) only if
$$
E[c_i \mid x_i] = 0 \;\;(\text{no selection on levels}), \qquad E[d_i \mid x_i] = 0 \;\;(\text{no selection on gains})
$$
- These are precisely "selection bias $=0$" and "ATT $=$ ATU." The two dialects have converged.

## Bad controls: a preview

- Adding controls is not free. If a control $x_2$ is itself **affected by the treatment**:
$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u, \qquad x_2 = \gamma_0 + \gamma_1 x_1 + \varepsilon
$$
- Then conditioning on $x_2$ blocks part of the effect you want --- and can *introduce* selection bias where none existed.
- Rule of thumb: controls must be **pre-treatment** (determined before $D$).
- Why conditioning on a post-treatment variable creates bias is subtle --- it is the centerpiece of the next lecture (DAGs and colliders).

# 7. Randomization Inference

## Two null hypotheses

- Standard (Neyman) null: the **average** effect is zero, $E[\delta_i] = 0$.
  - Inference relies on asymptotics: large $n$, estimated standard errors.
- Fisher's **sharp null**: the effect is zero *for every unit*,
$$
H_0: \delta_i = 0 \;\; \text{for all } i
$$
- Under the sharp null, both potential outcomes are known for every unit: $Y_i^1 = Y_i^0 = Y_i$ (the observed value).
- That makes *exact* inference possible --- no asymptotics, no standard errors, any sample size.

## The permutation logic

1. Under the sharp null, the observed outcomes would have been the same under **any** treatment assignment.
2. So: re-randomize $D$ on paper, in every way the actual randomization could have come out.
3. Recompute the test statistic (e.g., the SDO) for each hypothetical assignment.
4. The exact p-value is the share of assignments giving a statistic at least as extreme as the observed one.
- The randomness comes from the **assignment mechanism itself** --- the same mechanism that justified the causal comparison justifies the inference.

## Randomization inference on ten banks

- Suppose the ten banks were randomized: five drawn by lot got injections. Observed SDO $= 1.6$.

```{r}
#| echo: true
y <- c(2, 6, 7, 2, 6, 6, 7, 7, 7, 8)     # observed lending growth
treated <- c(2, 3, 5, 8, 9)              # banks drawn by lot
sdo_obs <- mean(y[treated]) - mean(y[-treated])

perms <- combn(10, 5)                     # all 252 possible assignments
sdo_all <- apply(perms, 2, function(tr) mean(y[tr]) - mean(y[-tr]))
p_exact <- mean(abs(sdo_all) >= abs(sdo_obs))
round(c(SDO = sdo_obs, exact_p = p_exact), 3)
```

- With $n = 10$, a difference of 1.6pp is *not* statistically distinguishable from luck-of-the-draw --- and we know this **exactly**, not asymptotically.

<!--

## Reading the code [1]: the observed statistic

- `treated <- c(2, 3, 5, 8, 9)` holds **positions**, not a 0/1 indicator: banks \#2, \#3, \#5, \#8, \#9 won the lottery.
- Two indexing idioms do the work:
  - `y[treated]` selects those five $\Rightarrow$ $6,7,6,7,7$, mean $6.6$.
  - `y[-treated]` **drops** them, giving the complement $\Rightarrow$ $2,2,6,7,8$, mean $5.0$.
- In R a negative index means *"exclude these"* --- not "count from the end," as in Python.
  - This works only because `treated` holds positive integers. For a logical vector you would need `y[!treated]`.
- So $SDO_{\text{obs}} = 6.6 - 5.0 = 1.6$.

## Reading the code [2]: the reference distribution

- `combn(10, 5)` lists every way to choose 5 units from 10: $\binom{10}{5} = 252$.
  - It returns a **matrix, 5 rows $\times$ 252 columns**; each *column* is one hypothetical assignment.
  - Despite the variable name, these are **combinations** --- order within a group is irrelevant.
- `apply(perms, 2, f)` runs `f` over **columns** (`MARGIN = 2`; `1` would mean rows), recomputing the same difference in means for all 252 assignments.
- **This is where the statistics live.** Reusing the same `y` under a different assignment is licensed only by the **sharp null**:
  - if $\delta_i = 0$ for every $i$, then $Y_i^1 = Y_i^0 = Y_i$, so each bank's outcome is the same no matter who was treated.
  - Under the weaker null "$ATE = 0$" this fails: units with offsetting $\delta_i \neq 0$ *would* have had different outcomes, so `y` cannot be held fixed.

## Reading the code [3]: the exact $p$-value

- `mean(abs(sdo_all) >= abs(sdo_obs))` does three things at once:
  1. `>=` produces a length-252 **logical** vector --- `TRUE` where that assignment was at least as extreme;
  2. taking `abs` of *both* sides makes the test **two-sided**;
  3. `mean()` coerces `TRUE`$\to 1$, so it returns the **proportion** --- here $100/252 = 0.397$.
- Structure worth noticing in the 252 values:
  - the distribution is **centered exactly at zero** and symmetric --- as the null requires;
  - it takes only **15 distinct values**, because ten small integers produce many ties. Discreteness is intrinsic to exact tests;
  - the smallest attainable two-sided $p$ is $2/252 \approx 0.008$. Ten units *can* reach significance --- this draw simply is not close.

## The assumption hiding in `combn(10, 5)`

- The reference set must match the **actual randomization design**.
- `combn(10, 5)` presumes a completely randomized experiment with a **fixed** five treated --- exactly five drawn by lot.
- Had each bank instead been **independently coin-flipped**, the correct reference set would be all $2^{10} = 1024$ assignments, including those with 3 or 7 treated --- and the $p$-value would differ.
- Get this wrong and the $p$-value is not "approximately right." It answers a different question.

\begin{center}
\emph{The permutation scheme must mirror the physical randomization.}
\end{center}

-->

## Where you will see this again

- Randomization inference is not a curiosity:
  - **Synthetic control**: placebo tests permute the treatment across donor units --- that *is* randomization inference.
  - **Small-N settings**: one treated state, one regulatory event, a handful of treated firms --- common in finance, hopeless for asymptotics.
  - Cluster-robust inference with few clusters: permutation tests as the honest fallback.
- The deeper point: *inference should mirror the assignment mechanism.* Where the mechanism is uncertain, so is the p-value.

# Wrapping Up

## Summary [Part 1]

- Causality is defined by **counterfactuals**; the fundamental problem is that one potential outcome is always missing.
- Estimands differ: **ATE, ATT, ATU** answer different policy questions. Know which one your design identifies.
- The naive comparison decomposes as
$$
SDO = ATE + \text{selection bias} + (1-\pi)(ATT - ATU)
$$
- Optimizing agents generate selection **on gains** --- the Perfect Regulator gets the sign wrong *because* it allocates well.
- More data does not fix a biased assignment mechanism.

## Summary [Part 2]

- **Randomization** kills both bias terms by making $(Y^1, Y^0) \indep D$. Balance tables are its testable symptom.
- **SUTVA** fails easily in finance: competition, peer effects, and heterogeneous doses. Ask what happens to the market, not just the treated.
- The **regression dialect** translates one-for-one: CMI $\Leftrightarrow$ independence; OVB $\Leftrightarrow$ selection bias; random coefficients $\Leftrightarrow$ heterogeneous $\delta_i$.
- **Randomization inference** gives exact p-values from the assignment mechanism --- remember it when you meet synthetic control.
- Next lecture: DAGs --- a graphical language for deciding *what to control for, and what not to*.

## References

\footnotesize

- Cunningham, S. *Causal Inference: The Remix*, Ch. 4. 
- Holland, P. (1986). Statistics and Causal Inference. *JASA* 81, 945--960.
- Blake, T., C. Nosko, and S. Tadelis (2015). Consumer Heterogeneity and Paid Search Effectiveness: A Large-Scale Field Experiment. *Econometrica* 83, 155--174.
- Karlan, D. and J. Zinman (2009). Observing Unobservables: Identifying Information Asymmetries with a Consumer Credit Field Experiment. *Econometrica* 77, 1993--2008.
- Bertrand, M., D. Karlan, S. Mullainathan, E. Shafir, and J. Zinman (2010). What's Advertising Content Worth? *Quarterly Journal of Economics* 125, 263--306.
- Cole, S., X. Giné, and J. Vickery (2017). How Does Risk Management Influence Production Decisions? *Review of Financial Studies* 30, 1935--1970.
- Duflo, E. and E. Saez (2003). The Role of Information and Social Interactions in Retirement Plan Decisions. *Quarterly Journal of Economics* 118, 815--842.
- Leary, M. and M. Roberts (2014). Do Peer Firms Affect Corporate Financial Policy? *Journal of Finance* 69, 139--178.
- Angrist, J. and J.-S. Pischke (2009). *Mostly Harmless Econometrics*. Princeton UP.

