---
title: "Unconfoundedness I: Assumptions and Matching"
subtitle: "BUSS975: Causal Inference in Financial Research"
author: "Professor Ji-Woong Chung"
institute: "Korea University"
format:
  beamer:
    aspectratio: 169
    fontsize: 11pt
    slide-level: 2
    toc: false
    pdf-engine: pdflatex
    pdf-engine-opts:
      - "-interaction=nonstopmode"
    keep-tex: true
    include-in-header:
      text: |
        \setbeamertemplate{navigation symbols}{}
        \setbeamertemplate{footline}[frame number]
        \setbeamertemplate{itemize items}[circle]
        \setbeamertemplate{itemize subitem}[triangle]
        \setbeamercolor{frametitle}{fg=black}
        \setbeamerfont{frametitle}{series=\bfseries}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{booktabs}
        \usepackage{tikz}
        \usetikzlibrary{arrows.meta,positioning,calc}
        \tikzset{
          dnode/.style={circle,draw,thick,minimum size=9mm,inner sep=1pt,font=\small},
          unode/.style={circle,draw,thick,dashed,minimum size=9mm,inner sep=1pt,font=\small},
          bnode/.style={rectangle,draw,thick,fill=gray!25,minimum size=9mm,inner sep=3pt,font=\small},
          de/.style={-{Stealth[length=2.2mm]},thick},
          dde/.style={-{Stealth[length=2.2mm]},thick,dashed}
        }
        % shrink code input (fancyvrb) and code output (verbatim)
        \usepackage{fancyvrb}
        \fvset{fontsize=\scriptsize}
        \makeatletter
        \def\verbatim@font{\scriptsize\ttfamily}
        \makeatother
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

## Outline

1. Introduction: what actually resolves bias
2. Confounders, covariates, and colliders
3. Identification assumptions
4. Subclassification
5. Exact matching
6. Nearest-neighbour matching
7. Inference with matching
8. Bias correction

\vspace{0.6em}

Lecture 3b takes the same assumption and does something different with it: collapse $X$ to a single number, weight instead of match, and compare all of it to regression.

# 1. Introduction

## Where we left off

- Lecture 2 gave us a rule for **which** variables to condition on: block every backdoor path, condition on no descendant of $D$.
- It was silent on two things:
  - What if the variables that would close the backdoors are **unobserved**? (Usually the case.)
  - Given a valid set, **how** do we actually condition --- and does it matter?
- This lecture takes the optimistic branch: **suppose the confounders are observed**. What then?
- The methods --- subclassification, matching, propensity scores, regression --- are all answers to "how." They differ less than their names suggest.

<!--
## Two rivers of causal inference

Cunningham opens the chapter with a piece of intellectual history, because it explains why this material is contested.

\begin{center}
\begin{tabular}{p{5.4cm}p{5.4cm}}
\toprule
\textbf{The statistics tradition} & \textbf{The labor economics tradition} \\
Rubin, Imbens, Abadie & Card, Angrist, Krueger \\
\midrule
Model the \emph{assignment mechanism}; adjust for covariates & Find an \emph{external} source of variation \\
Unconfoundedness, matching, propensity scores & IV, RDD, DiD \\
``Which covariates make treatment as-if random?'' & ``What shock assigned treatment for us?'' \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- The Princeton school was, in Cunningham's words, "pretty skeptical that unconfoundedness based methods could on their own successfully eliminate selection bias in observational work."
- Empirical finance drinks from both rivers. You need to know why each camp distrusts the other.
-->


## Treatment assignment mechanisms resolve bias, not models

Recall Lecture 1's decomposition. It is an **identity** --- true in every dataset, experimental or not:

$$
\underbrace{E[Y|D{=}1] - E[Y|D{=}0]}_{SDO}
\;=\; ATE
\;+\; \underbrace{E[Y^0|D{=}1] - E[Y^0|D{=}0]}_{\text{selection on levels}}
\;+\; \underbrace{(1-\pi)(ATT - ATU)}_{\text{selection on gains}}
$$

- Nothing here is an assumption. The identity holds whatever you do.
- **A regression cannot delete the last two terms.** No functional form, no set of fixed effects, no clustering choice touches them.
- What deletes them is a **claim about how treatment was assigned**.

## What randomization actually did

- Under $(Y^1,Y^0) \indep D$, we were permitted to *deduce*:
$$
E[Y^0|D{=}1] = E[Y^0|D{=}0], \qquad E[Y^1|D{=}1] = E[Y^1|D{=}0]
$$
- The first kills selection on levels. Both together kill $ATT - ATU$. What is left is the $ATE$.
- The deductions came from the **assignment mechanism**, not from any estimator. We then *inserted* them into a statistical model.
<!--- Cunningham's line, worth memorising:-->

\vspace{0.3em}
\begin{center}
\emph{The statistical model alone cannot resolve bias --- the statistical model matched with the assignment mechanism can.}
\end{center}

## Why this framing matters

- Randomization is an **assignment mechanism**, not a statistical model. A regression is a **statistical model**, not an assignment mechanism.
- Confusing the two produces the most common error in empirical finance: believing that a long enough control list is itself an identification strategy.
- Unconfoundedness is this lecture's assignment mechanism. It says treatment was assigned **as if at random within strata of $X$**.
- Everything that follows --- subclassification, matching, propensity scores, regression --- is a statistical model that becomes credible *only* if you can defend that mechanism.
- So the order of business is: defend the mechanism first, choose the estimator second. 

## The bet we are making

\begin{center}
\begin{tikzpicture}
\node[dnode] (D) at (0,0)   {$D$};
\node[dnode] (Y) at (4.6,0) {$Y$};
\node[dnode] (X) at (2.3,1.5) {$X$};
\draw[de] (D) -- (Y);
\draw[de] (X) -- (D);
\draw[de] (X) -- (Y);
\end{tikzpicture}
\end{center}

- Everything rests on one claim: **there is no dashed circle**. No unobserved $U$ pointing at both $D$ and $Y$.
- That claim is **not testable**. No diagnostic, balance table, or specification test can establish it.
- It is defended with institutional knowledge, not statistics --- which is why this lecture spends as much time on *what could go wrong* as on estimators.
- The literature calls it **unconfoundedness**, **conditional independence (CIA)**, or **selection on observables**. Same thing.

# 2. Confounders, Covariates, and Colliders

## A worked DAG: first class on the Titanic

- **Question**: did a first-class ticket ($D$) raise your chance of surviving ($Y$)?
- The raw comparison is stark: $62.5\%$ of first-class passengers survived, versus $27.1\%$ of everyone else --- a gap of $35.4$ percentage points.
- But the *"women and children first"* rule was actually enforced, and first-class cabins were closer to the boat deck.
- So **sex** and **age** drive both who bought a first-class ticket and who got into a lifeboat. Textbook confounding.
- We use this because the data are real, complete, and small enough to compute by hand --- and because the confounders are not in dispute.

## The Titanic DAG

\begin{center}
\begin{tikzpicture}
\node[dnode] (D) at (0,0)     {$D$};
\node[dnode] (M) at (2.2,0)   {$M$};
\node[dnode] (Y) at (4.4,0)   {$Y$};
\node[dnode] (W) at (1.1,1.7) {$W$};
\node[dnode] (C) at (3.3,1.7) {$C$};
\node[dnode] (F) at (2.2,-1.9){$F$};
\node[dnode] (S) at (6.6,1.2) {$S$};
\draw[de] (D) -- (M); \draw[de] (M) -- (Y);
\draw[de] (D) to[bend right=40] (Y);
\draw[de] (W) -- (D); \draw[de] (W) -- (Y);
\draw[de] (C) -- (D); \draw[de] (C) -- (Y);
\draw[de] (D) -- (F); \draw[de] (Y) -- (F);
\draw[de] (S) -- (Y);
\end{tikzpicture}
\end{center}

\vspace{-0.3em}

\footnotesize
$D$ first class \quad $Y$ survival \quad $W$ sex \quad $C$ age \quad $M$ lifeboat \quad $F$ fame \quad $S$ swimming

\normalsize

- Backdoor paths: $D \leftarrow W \rightarrow Y$ and $D \leftarrow C \rightarrow Y$. Both **open**. Condition on $\{W,C\}$ and both close.
- $F$ is a **collider**: $D \rightarrow F \leftarrow Y$. Already closed; conditioning on it *opens* it.
- $M$ is a **mediator**: first class sat nearer the boat deck. It is *part of* the effect, not a confounder of it.

## Four kinds of variable

\begin{center}
\begin{tabular}{llp{6.5cm}}
\toprule
Type & Example & What conditioning does \\
\midrule
\textbf{Confounder}       & $W$ sex, $C$ age & \textbf{Required.} Closes a backdoor path. \\
\textbf{Collider}         & $F$ fame         & \textbf{Forbidden.} Opens a closed path. \\
\textbf{Neutral / precision} & $S$ swimming  & Optional. No bias either way; shrinks the standard error. \\
\textbf{Mediator}         & $M$ lifeboat    & \textbf{Forbidden} for the total effect. Blocks the causal path. \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- Only one of these four rows is about bias *reduction*. Two are about bias *creation*.
- Note $S$: it predicts the outcome but not treatment. Free precision --- the honest case for controls that are not confounders.
- $M$ is the subtle one. Conditioning on it asks: *did first class help, holding fixed the main way it helped?* Near zero --- and the wrong question.
- The categories are properties of the **graph**, not the variable. Press coverage *before* boarding would not be a collider.

## The same DAG in corporate finance

Effect of a **bank credit line** ($D$) on **firm investment** ($Y$):

\begin{center}
\begin{tabular}{lll}
\toprule
Titanic & Finance analogue & Rule \\
\midrule
$W$, $C$ sex, age  & Firm size, industry, credit rating & Condition on \\
$F$ fame           & \emph{Subsequent} analyst coverage & Never condition on \\
$S$ swimming       & Pre-period export-market demand    & Optional, improves precision \\
$M$ lifeboat       & Amount actually drawn on the line  & Never condition on \\
--- & Unobserved managerial quality & The thing that sinks you \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- Analyst coverage is the finance collider you will meet: it responds *both* to the financing event and to firm performance.
- The drawn amount is the mediator a referee will *ask* you for --- "surely control for how much they actually borrowed." That removes the channel you are trying to measure.
- The last row has no Titanic counterpart, and that is the whole problem. On the Titanic we know the confounders are sex and age. In finance we rarely do.

## Covariate selection without a DAG

In practice you rarely have expert consensus on the full graph. Three working rules:

1. **Include variables you are confident are confounders.** Prior knowledge about the assignment mechanism is the input, not the data.
2. **Include pre-treatment variables that strongly predict the outcome.** These are either confounders or precision-improvers like $S$. Crucially, they are *probably not colliders* --- a variable measured before treatment cannot be caused by it.
3. **Exclude anything that might be an outcome.** If you are unsure, exclude everything measured after treatment.

\vspace{0.3em}

- Rule 2 does most of the work in practice, and rule 3 is the one most often violated in finance.
- "Pre-treatment" is doing heavy lifting in rule 2. It is a timing argument, and timing is checkable.

<!--
## Placebo analysis

The one *empirical* check on a covariate strategy:

- Run your specification on a period **before treatment occurred**, where the effect is known to be zero by construction.
- If your covariate set produces a "significant effect" of the program when the program did not yet exist, the strategy is not credible.
- This does not validate unconfoundedness. Passing is necessary, not sufficient --- it can only detect a strategy that is *already* broken.
- Related finance practice: pre-trend plots, lagged-treatment coefficients, and outcomes that the treatment could not plausibly move.

\vspace{0.3em}
\begin{center}
\emph{A placebo test can refute your design. It can never confirm it.}
\end{center}

## A placebo test in action

**Setting.** Firms get a bank credit line ($D$) at $t=0$; outcome is investment. You observe
size $S$ but not quality $Q$ or a transitory shock $Z$. You control for $S$ only.

**The test.** Regress the *pre-treatment* outcome $Y_{-1}$ on $D$ and $S$. Treatment did not
exist yet, so the true coefficient is **zero by construction**.

\vspace{0.3em}

Two worlds, identical in every draw except **what drives selection into treatment**:

\begin{center}
\begin{tabular}{lll}
\toprule
 & Selection driven by & Does it move $Y_{-1}$? \\
\midrule
\textbf{Persistent} confounder & $Q$ --- firm quality, always present & \textbf{Yes} \\
\textbf{Transitory} confounder & $Z$ --- a shock arriving at $t=0$   & \textbf{No} \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.3em}

- Both violate unconfoundedness. Both bias the estimate. Only one leaves a trace in the past.

## Running the placebo

```{r}
#| echo: true
n <- 5000
sim <- function(persistent) {
  set.seed(975)                                 # same draws in both worlds
  Q <- rnorm(n); S <- rnorm(n); Z <- rnorm(n)   # Q, Z UNOBSERVED; S observed
  drv <- if (persistent) Q else Z               # what drives selection into D
  D   <- as.integer(0.9 * drv + 0.9 * S + rnorm(n) > 0)
  Ypre  <- Q + S + rnorm(n)                     # BEFORE treatment exists
  Ypost <- Q + S + 1.5 * Z + 2 * D + rnorm(n)   # true effect = 2
  pl <- coef(summary(lm(Ypre  ~ D + S)))["D", ] # the PLACEBO regression
  po <- coef(summary(lm(Ypost ~ D + S)))["D", ] # the real regression
  round(c(placebo = pl[[1]], plac_t = pl[[3]], estimate = po[[1]]), 2)
}
rbind(persistent = sim(TRUE), transitory = sim(FALSE))
```

\vspace{0.2em}

- Column 1--2: the placebo coefficient and its $t$. Column 3: the estimate you would publish. **Truth is 2.**

## Reading the placebo

\begin{center}
\begin{tabular}{lccc}
\toprule
 & Placebo coef. ($t$) & Verdict & Published estimate \\
\midrule
Persistent confounder & $1.10$ \; $(t = 26.5)$  & \textbf{FAILS} & $3.07$ \; (bias $+1.07$) \\
Transitory confounder & $-0.01$ \; $(t = -0.2)$ & \textbf{PASSES} & $3.61$ \; (bias $+1.61$) \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- Row 1 is the success story: a $t$ of $26.5$ on an effect that **cannot exist**. The design is caught before you publish it.
- Row 2 is the warning. The placebo is immaculate --- $t = -0.2$, exactly what a clean design looks like --- and the estimate is off by $1.61$.
- **The design that passed is more biased than the design that failed.** Passing bought you nothing.
- Why: a placebo can only detect confounders that were *already moving the outcome* before treatment. A shock arriving *with* treatment is invisible to it, and those are common in finance --- a credit-supply shock, a rating change, a new CEO.

## What to take from this

- **Run it anyway.** It is nearly free, and row 1 is a real failure mode that it really does catch.
- **Report it honestly.** "The placebo passes" is evidence of *absence of one kind of problem*, not evidence your design is sound.
- Never write "the placebo test confirms our identifying assumption." It cannot. Write: *"we find no evidence of differential pre-treatment trends."*
- The corresponding finance practices --- pre-trend plots, lagged-treatment coefficients, unaffected outcomes --- all share this same asymmetry, and all get over-claimed in the same way.

\vspace{0.3em}
\begin{center}
\emph{Failing is informative. Passing is merely not-yet-refuted.}
\end{center}
-->


# 3. Identification Assumptions

## Selecting the target parameter

Before estimating anything, decide *whose* effect you want.

$$
ATE = E[Y^1 - Y^0], \quad
ATT = E[Y^1 - Y^0 \mid D{=}1], \quad
ATU = E[Y^1 - Y^0 \mid D{=}0]
$$

- Cunningham's example: a polio vaccine. The three parameters answer three different policy questions.
  - $ATE$: what if we vaccinate **everyone**?
  - $ATT$: how much did the vaccine help **those who took it**?
  - $ATU$: what would happen if we reached **the currently unvaccinated**?
- These coincide only under constant treatment effects. With heterogeneity they differ, sometimes a lot.
- **The estimand is a choice you make about the policy question.**

## Which parameter does the question need?

\begin{center}
\begin{tabular}{p{5.6cm}l}
\toprule
Finance question & Parameter \\
\midrule
Should the regulator extend this mandate to all firms? & $ATE$ \\
Did the firms that adopted IFRS benefit from it? & $ATT$ \\
Would forcing non-adopters to adopt help them? & $ATU$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- Note the third row: it is the *expansion* question, and it is almost never the one your data can answer well.
- Papers routinely estimate the $ATT$ and then write policy conclusions that require the $ATE$ or $ATU$. Watch for this in referee reports --- including your own.

## Unconfoundedness is a version of independence

- **Strong unconfoundedness:**
$$
(Y^1, Y^0) \;\indep\; D \;\mid\; X
$$
- Compare Lecture 1's randomization assumption, $(Y^1,Y^0) \indep D$, with no conditioning.
- In words: among units with the *same* $X$, treatment is as good as randomly assigned. It is **conditional randomization**.
- This is the assignment mechanism from §1, Introduction, stratum by stratum:
$$
\begin{aligned}
E[Y^1 \mid X{=}x] &= E[Y^1 \mid D{=}1, X{=}x] = E[Y^1 \mid D{=}0, X{=}x] \\[0.7em]
E[Y^0 \mid X{=}x] &= E[Y^0 \mid D{=}1, X{=}x] = E[Y^0 \mid D{=}0, X{=}x]
\end{aligned}
$$
- Strong because it's conditional independence with respect to **both** potential outcomes.

## What the assumption buys: plug-in

- The fundamental problem of causal inference is that $Y^1$ and $Y^0$ are never both observed.
- Unconfoundedness lets us **replace a missing counterfactual with an observed outcome** from a comparable unit:

$$
\underbrace{E[Y^0 \mid D{=}1, X{=}x]}_{\text{never observed}}
\;=\;
\underbrace{E[Y^0 \mid D{=}0, X{=}x]}_{\text{observed}}
\;=\;
\underbrace{E[Y \mid D{=}0, X{=}x]}_{\text{in your data}}
$$

- The first equality is the **assumption**. The second is the switching equation --- pure definition.
- Every method in this lecture is a different way of executing that substitution.
- So switching estimators cannot rescue a design: all of them lean on the first equality, and none of them can test it. **Identification** credibility is common to all.
- They *do* differ downstream --- functional form, extrapolation into thin regions, which strata get weight. Those are **estimation** differences; Lecture 3b, §3, What Matching Cannot Do, returns to them.

## Unconfoundedness is controversial

- Read literally, the assumption says: firms with identical observables effectively **flip a coin** over a major financial decision.
- Every model of optimizing behaviour says otherwise. Firms choose treatment *because* they expect to benefit --- Lecture 1's selection on gains. <!--, and what the literature calls **essential heterogeneity**.-->
- This is the deepest objection, and it comes from economics, not statistics. <!--The Princeton river was built on it. It may be at least partly for this reason that the Princeton group tended to favor methods where treatments were assigned to units from the outside, as opposed to treatments that were conditionally random.-->
- It is not a reason never to use these methods. It is a reason to state, explicitly, why *your* setting escapes it.
<!-- Unconfoundness means "no confounders." If you're confident that there is no confounder (think of it as omitted variables) in your research design, you proceed. Otherwise, don't. -->

## Guido W. Imbens's three defences

1. **Statistical motivation.** It is the natural starting point. <!-- the benchmark you must argue *away* from, and the one a reader will assume you considered.-->
2. **It is an empirical question.** With enough institutional knowledge, the relevant $X$ may genuinely be measurable. Whether it is depends on the setting, not on econometrics.
3. **Objectives differ.** The decision-maker optimizes over something *other than your outcome*. A firm choosing a credit line optimizes liquidity and covenant slack --- not the investment measure you happen to study.

\vspace{0.3em}

- Defence 3 is the strongest available, and worth naming explicitly when you use it.
- It also tells you when the design *fails*: when the agent is optimizing over precisely your outcome variable.

## Road map: two routes through this chapter

Everything ahead assumes unconfoundedness. The methods split on **what they do when a treated unit has no comparable control.**

\begin{center}
\begin{tabular}{lp{4.3cm}p{4.3cm}}
\toprule
 & \textbf{Route 1: interpolation} & \textbf{Route 2: extrapolation} \\
\midrule
Rests on       & Common support & Functional form \\
Fills the gap by & It does not --- it refuses & Projecting the fitted model into it \\
Methods        & Subclassification, matching, propensity scores & Regression \\
Covered in     & §4--§8, and Lecture 3b §1 & Lecture 3b §2 \\
\bottomrule
\end{tabular}
\end{center}

<!--
\vspace{0.3em}

- Route 1 buys **honesty**: it cannot compare units with no counterpart, so it stops.
- Route 2 buys **coverage**: it always returns a number --- including where the data are silent.
- The routes meet in §8, Bias Correction, and in doubly-robust estimators, which use a model only to patch small gaps.

## Interpolation versus extrapolation

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.3
set.seed(975)
par(mar = c(4, 4.2, 1.0, 1), cex = 0.85)
mu0 <- function(x) 2 + 0.5 * x + 0.25 * x^2
nc <- 260; nt <- 180
Xc <- runif(nc, 0, 6);  Yc <- mu0(Xc) + rnorm(nc, 0, 1.6)
Xt <- runif(nt, 2, 10)
fit <- lm(Yc ~ Xc); g <- seq(0, 10, 0.05)
pred <- coef(fit)[1] + coef(fit)[2] * g
plot(NULL, xlim = c(0, 10), ylim = c(0, 46), xlab = "covariate  X",
     ylab = expression(Y^0), main = "", yaxt = "n")
axis(2, at = seq(0, 40, 10))
rect(2, -5, 6, 38, col = "#4D9DE022", border = NA)
segments(c(2, 6), -5, c(2, 6), 38, col = "#4D9DE0", lty = 3)
points(Xc, Yc, pch = 16, col = "#00000040", cex = 0.55)
segments(Xt, 40.5, Xt, 42.5, col = "#B2182B60", lwd = 0.8)
text(5, 44.6, "where the treated units are", col = "#B2182B", cex = 0.82)
lines(g, mu0(g), lwd = 2.4, col = "black")
lines(g[g <= 6], pred[g <= 6], lwd = 2.4, col = "#E1A100")
lines(g[g >= 6], pred[g >= 6], lwd = 2.4, col = "#E1A100", lty = 2)
arrows(9, coef(fit)[1] + coef(fit)[2] * 9, 9, mu0(9),
       code = 3, length = 0.045, col = "#B2182B", lwd = 1.4)
text(9, 30.5, "bias", col = "#B2182B", cex = 0.82, adj = c(0.5, 0))
text(4, 35, "common support", col = "#2A6F9E", cex = 0.85)
text(8.2, 5, "no controls here", col = "#B2182B", cex = 0.85)
legend("topleft", c(expression(paste("truth  ", mu^0, (X))),
                    "linear fit on controls", "controls", "extrapolated"),
       col = c("black", "#E1A100", "#00000060", "#E1A100"),
       lty = c(1, 1, NA, 2), pch = c(NA, NA, 16, NA), lwd = c(2.4, 2.4, NA, 2.4),
       bty = "n", cex = 0.78)
```

\vspace{-0.4em}

\footnotesize
Inside the band, both groups exist: the counterfactual is **interpolated** from nearby controls. Past $X = 6$ there are no controls at all --- the dashed line is the model talking, not the data, and it misses the truth by $8.2$.
-->


## Common support

- Unconfoundedness alone is not enough. You also need units to compare.
- **Strong common support** (needed for the $ATE$):
$$
0 < \Pr(D=1 \mid X = x) < 1 \quad \text{for all } x
$$
  At every value of $X$, both treated and untreated units must exist.
<!-- - Cunningham's jigsaw analogy: **unconfoundedness is the blueprint; common support is whether the pieces are actually in the box.**
- A perfect blueprint is useless if a piece is missing --- and a complete box is useless if the blueprint is wrong. -->

## The two assumptions differ in kind

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Unconfoundedness} & \textbf{Common support} \\
\midrule
Untestable & \textbf{Verifiable in your data} \\
Defended with institutional knowledge & Checked with a histogram \\
Failure $\Rightarrow$ bias & Failure $\Rightarrow$ extrapolation \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- You can have one without the other, and the failures look nothing alike.
- Common support is the assumption you have **no excuse** for not checking --- and the one most papers skip.
- This distinction is the key to the NSW debate at the end of this lecture.

## Weighting treatment effects under unconfoundedness

Put the two assumptions together and the $ATE$ becomes computable. 

$$
\begin{aligned}
ATE(x) &= E\big[\,Y^1 \mid X{=}x\,\big] - E\big[\,Y^0 \mid X{=}x\,\big] \\
       &= \underbrace{E\big[\,Y^1 \mid D{=}1, X{=}x\,\big] - E\big[\,Y^0 \mid D{=}0, X{=}x\,\big]}_{\text{unconfoundedness --- used on \emph{both} terms}} \\
       &= \underbrace{E\big[\,Y \mid D{=}1, X{=}x\,\big] - E\big[\,Y \mid D{=}0, X{=}x\,\big]}_{\text{switching equation}} \\[0.3em]
ATE    &= \underbrace{\sum_x \Big( E[Y \mid D{=}1, X{=}x] - E[Y \mid D{=}0, X{=}x] \Big)\Pr(X{=}x)}_{\text{iterated expectations; common support}}
\end{aligned}
$$

- $ATE(x)$ is the effect **within stratum $x$** --- *conditional average treatment effect* (CATE). The last line averages those with **population** weights.
- **Unconfoundedness is used twice**, once for $Y^1$ and once for $Y^0$. That is precisely why the $ATE$ needs the *strong* version.


## A three-stratum example

Suppose the effect on mortality differs by age group, and the groups are equally sized:

\begin{center}
\begin{tabular}{lcc}
\toprule
Stratum & Effect in stratum (pp) & $\Pr(X = x)$ \\
\midrule
Young       & $80$  & $1/3$ \\
Middle-aged & $55$  & $1/3$ \\
Old         & $100$ & $1/3$ \\
\bottomrule
\end{tabular}
\end{center}

$$
\begin{aligned}
  \widehat{\mathit{ATE}} &= \frac{N_y}{N} \times SDO_{y} +
  \frac{N_m}{N} \times SDO_{m} + \frac{N_o}{N} \times SDO_{o} \\
  &= \frac{1}{3} \times 80 + \frac{1}{3} \times 55 + \frac{1}{3} \times 100 \\
  &= 78.33
\end{aligned}
$$


- The $ATE$ is a number **no stratum experiences**. It describes a population, not a unit.


## Estimating the ATT under weaker assumptions

- A weaker assumption suffices if you only want the $ATT$:
$$
Y^0 \;\indep\; D \;\mid\; X
$$
- Only the **untreated** potential outcome must be independent of treatment.
- It allows firms to *anticipate their own gains* $Y^1$ and select on them, while assuming they do not select on how they would have done anyway.
  - In finance that ordering is often defensible. A firm knows its expected benefit from a credit line better than it knows its counterfactual path without one.
- **Weak common support**: Controls must exist wherever treated units do --- but not conversely.
$$
  \Pr(D = 0 \mid X) > 0 \quad \text{for all } X \text{ such that }
  \Pr(D = 1 \mid X) > 0 
$$

## Why the asymmetry makes sense

$$
\begin{aligned}
ATT(x) &= E[Y^1 \mid D{=}1, X{=}x] - E[Y^0 \mid D{=}1, X{=}x] \\[0.3em]
       &= \underbrace{E[Y \mid D{=}1, X{=}x]}_{\text{switching equation}} - E[Y^0 \mid D{=}1, X{=}x] \\[0.3em]
       &= E[Y \mid D{=}1, X{=}x] - \underbrace{E[Y \mid D{=}0, X{=}x]}_{\text{weak unconfoundedness}} \\[0.4em]
ATT    &= \underbrace{\sum_x \Big( E[Y \mid D{=}1, X{=}x] - E[Y \mid D{=}0, X{=}x] \Big)\Pr(X{=}x \mid D{=}1)}_{\text{common support, wherever treated units live}}
\end{aligned}
$$

- Weighting by $\Pr(X \mid D{=}1)$ --- covariates **among the treated** --- is what makes it the $ATT$: strata with no treated units get zero weight.
- **Rule**: if you can only defend the weak version, your estimand is the $ATT$. Say so.
<!-- - This is the single most common honest retreat in observational finance, and it costs you nothing to state clearly.-->

# 4. Subclassification

## Cochran's method

- The oldest estimator here, and the most transparent. Three steps:
  1. **Stratify** on $X$.
  2. Compute the treated-minus-control difference **within** each stratum.
  3. **Average** the strata, weighting by whatever the estimand requires.
- No functional form is assumed. Within a stratum, units are identical on $X$ by construction.
- The weights define the estimand:

\begin{center}
\begin{tabular}{lll}
\toprule
Estimand & Weight stratum $k$ by & Interpretation \\
\midrule
$ATE$ & $n_k / N$                      & population share \\
$ATT$ & $n_{k,\text{treated}} / N_{\text{treated}}$   & treated share \\
$ATU$ & $n_{k,\text{untreated}} / N_{\text{untreated}}$ & control share \\
\bottomrule
\end{tabular}
\end{center}

## The Titanic: the raw comparison

- Treated: 325 first-class passengers, $62.5\%$ survived.
- Control: 1,876 others, $27.1\%$ survived.
$$
SDO = 62.5 - 27.1 = 35.4 \text{ percentage points}
$$
- Taken at face value: a first-class ticket nearly *tripled* your survival odds.
- But the first-class cabins held a very different mix of people. Sex and age are confounders, and we can see them directly.
- Stratify on $\{$sex, age$\}$: four strata, and every one of them is occupied.

## The four strata

\begin{center}
\begin{tabular}{lrrrrr}
\toprule
Stratum & $n$ & first class & survival, 1st & survival, other & difference \\
\midrule
Adult male   & 1,667 & 175 & $32.6\%$  & $18.8\%$ & $13.7$ pp \\
Adult female &   425 & 144 & $97.2\%$  & $62.6\%$ & $34.6$ pp \\
Male child   &    64 &   5 & $100.0\%$ & $40.7\%$ & $59.3$ pp \\
Female child &    45 &   1 & $100.0\%$ & $61.4\%$ & $38.6$ pp \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- Every within-stratum difference is **smaller than the $35.4$ pp raw gap.** Confounding was inflating the estimate.
- Look at the composition: $76\%$ of all passengers were adult males --- the stratum with the *smallest* effect --- but only $54\%$ of first class.
- Note the last stratum: **one** first-class female child. Remember her; she returns shortly.

## Three estimands, three answers

Same four differences. Only the weights change.

\begin{center}
\begin{tabular}{lrrrr}
\toprule
Weight on & Adult male & Adult female & Male child & Female child \\
\midrule
$ATE$ (population share)  & $0.757$ & $0.193$ & $0.029$ & $0.020$ \\
$ATT$ (treated share)     & $0.538$ & $0.443$ & $0.015$ & $0.003$ \\
$ATU$ (control share)     & $0.795$ & $0.150$ & $0.031$ & $0.023$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.3em}

$$
\widehat{ATE} = 19.6 \text{ pp}, \qquad
\widehat{ATT} = 23.8 \text{ pp}, \qquad
\widehat{ATU} = 18.9 \text{ pp}
$$

- The raw $35.4$ pp falls to roughly **20 pp** once composition is held fixed. Nearly half the headline gap was confounding.
- $ATT > ATE$ because first class was disproportionately adult *female* --- the stratum with a large effect.

<!--
## Subclassification in six lines

```{r}
#| echo: true
d  <- transform(as.data.frame(Titanic), D = Class == "1st")
tt <- xtabs(Freq ~ Sex + Age + D + Survived, d)
n  <- tt[, , , "Yes"] + tt[, , , "No"]      # cell sizes
r  <- tt[, , , "Yes"] / n                   # survival rates
del <- r[, , "TRUE"] - r[, , "FALSE"]       # within-stratum differences
round(c(ATE = weighted.mean(del, n[, , "TRUE"] + n[, , "FALSE"]),
        ATT = weighted.mean(del, n[, , "TRUE"]),
        ATU = weighted.mean(del, n[, , "FALSE"])), 3)
```

\vspace{0.3em}

- The entire method is `weighted.mean` called three times with three weight vectors.
- If an estimator's output changes when you change *only* the weights, the estimand was a choice --- not a finding.
-->


## The curse of dimensionality

- Now delete one passenger: the single first-class female child.
- That stratum still exists --- 44 girls travelled in other classes --- but it now has **no treated unit**. The within-stratum difference is undefined.

\begin{center}
\begin{tabular}{lc}
\toprule
Parameter & Still identified? \\
\midrule
$ATE$ & \textbf{No} --- needs all four differences \\
$ATU$ & \textbf{No} --- female children get weight $0.023$ \\
$ATT$ & \textbf{Yes} --- $23.7$ pp from the three occupied strata \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

- One passenger out of 2,201 destroys two of the three parameters. The $ATT$ survives because it never needed that cell.
- This is common support failing.

## Why it gets worse fast

- One binary covariate: 2 strata. Ten binary covariates: $2^{10} = 1{,}024$. One continuous covariate: infinitely many.
- Cells multiply exponentially; your sample does not. Empty cells become the norm, not the exception.
- Non-parametric methods --- subclassification, exact matching, reweighting --- **break down honestly**: they simply cannot compute the estimate.
- Regression **does not break down**. It extrapolates across the empty cells using functional form, and reports a number with a standard error.

<!--
## Simulating the collapse

$n = 1{,}000$ firms, $k$ independent binary covariates, treatment assigned by a coin flip.

```{r}
#| echo: true
set.seed(975)
n <- 1000
for (k in c(2, 4, 6, 8, 10)) {
  X    <- matrix(rbinom(n * k, 1, 0.5), n, k)
  D    <- rbinom(n, 1, 0.5)
  cell <- apply(X, 1, paste, collapse = "")
  tt   <- table(cell, D)
  cat(sprintf("k=%2d: %5d strata | %4d occupied | %4d usable (%3.0f%%)\n",
      k, 2^k, nrow(tt), sum(tt[,1] > 0 & tt[,2] > 0),
      100 * sum(tt[,1] > 0 & tt[,2] > 0) / nrow(tt)))
}
```

## Reading the collapse

- Through $k = 6$ everything is fine: every stratum is occupied and usable.
- At $k = 8$, only $71\%$ of occupied strata contain both a treated and a control firm.
- At $k = 10$, just **$146$ of $1{,}024$** strata are usable --- $23\%$. The other $77\%$ of the design is unavailable.
- And this is the *easy* case: covariates independent, treatment randomized, no continuous variables. Real data is worse on all three counts.
- Two escape routes, and the rest of the lecture is about them:
  - **Reduce the dimension.** Collapse $X$ into a single number --- the propensity score.
  - **Impose functional form.** Let regression extrapolate across empty cells --- at the cost of assuming the shape.
-->


# 5. Exact Matching

## Matching as explicit imputation

- Every causal inference is an imputation problem: the counterfactual is missing and must be filled in.
- Most methods impute **implicitly**, inside a formula. Matching does it **explicitly**, unit by unit --- and that transparency is its main virtue.
- Start from the definition of the $ATT$. Exactly **one** term is unavailable:

$$
ATT \;=\; \underbrace{E[Y^1 \mid D{=}1]}_{\text{observed: } Y = Y^1 \text{ when } D=1}
\;-\; \underbrace{E[Y^0 \mid D{=}1]}_{\textbf{missing}}
$$

- The first term is just the treated sample mean. Everything matching does is build a stand-in for the second.

## Replacing the missing term

Condition on $X$, and the assumption does its work:

$$
\begin{aligned}
E[Y^0 \mid D{=}1, X{=}x]
  &= \underbrace{E[Y^0 \mid D{=}0, X{=}x]}_{\text{weak unconfoundedness}} \\[0.2em]
  &= \underbrace{E[Y \mid D{=}0, X{=}x]}_{\text{switching equation}} \\[0.4em]
E[Y^0 \mid D{=}1]
  &= \underbrace{\sum_x E[Y \mid D{=}0, X{=}x]\,\Pr(X{=}x \mid D{=}1)}_{\text{common support: this must exist wherever treated units live}}
\end{aligned}
$$

- The missing quantity is now written entirely in terms of things you can estimate from control units.
- **Matching is simply a non-parametric estimate of $E[Y \mid D{=}0, X{=}X_i]$** --- take the controls that look like $i$, and average them.

## The estimator, and what to do about ties

Let $\mathcal{J}(i)$ be the set of control units matched to treated unit $i$, and $M_i = |\mathcal{J}(i)|$.

$$
\widehat{ATT} \;=\; \frac{1}{N_T}\sum_{i\,:\,D_i = 1}
\Bigg( Y_i - \underbrace{\frac{1}{M_i}\sum_{j \in \mathcal{J}(i)} Y_j}_{\widehat{Y^0_i}} \Bigg)
$$

- If two controls sit at exactly the same distance from $i$, use both --- breaking the tie arbitrarily injects the analyst's coin flip into the estimate.
<!-- - $M_i$ also varies when a caliper binds: a treated unit with no control inside the caliper contributes nothing, and $N_T$ must then count only the units you actually matched.
- You can *point at the specific firms* doing the work for each observation. No regression output permits that. -->

## The same idea for the ATE

For a control unit $j$, the missing half is $Y^1$. Let $\mathcal{I}(j)$ be the treated units matched to it, $M_j = |\mathcal{I}(j)|$.

$$
\widehat{ATE} = \frac{1}{N}\Bigg[
\underbrace{\sum_{i\,:\,D_i=1}\Big( Y_i - \tfrac{1}{M_i}\!\!\sum_{j \in \mathcal{J}(i)}\!\! Y_j \Big)}_{\text{treated: observed} \,-\, \text{imputed}}
\;+\;
\underbrace{\sum_{j\,:\,D_j=0}\Big( \tfrac{1}{M_j}\!\!\sum_{i \in \mathcal{I}(j)}\!\! Y_i - Y_j \Big)}_{\text{control: imputed} \,-\, \text{observed}}
\Bigg]
$$

- Every unit contributes one *estimated* treatment effect; the $ATE$ averages all $N$ of them.
- Equivalently $\widehat{ATE} = \frac{N_T}{N}\widehat{ATT} + \frac{N_C}{N}\widehat{ATU}$ --- the weights from §4, Subclassification, back again.


## Exact matching and its limit

- With discrete $X$ and full common support, exact matching and subclassification give the **same answer**. They are the same estimator, organized differently.
- The difference is bookkeeping: subclassification aggregates to stratum effects first; matching imputes unit by unit.
- Same fatal problem, too: with a continuous covariate, **exact matches occur with probability zero.** 
- Two responses, and they define the next two sections:
  - Give up on exact equality --- match to the *nearest* unit (§6, Nearest-Neighbour Matching), then correct that error (§8, Bias Correction).
  - Give up on matching $X$ itself --- match on a *summary* of $X$ (Lecture 3b, §1, Propensity Scores).

# 6. Nearest-Neighbour Matching

## One confounder

- Drop the demand for exact equality. Match each treated unit to the control that minimizes $|X_i - X_j|$.
- The gap you accept is called the **matching discrepancy**, and §8, Bias Correction, is entirely about what it costs you.
- Choices you must make and report:
  - **With or without replacement.** With replacement always takes the closest match, so it **lowers bias** and raises **variance.**
  <!-- reuse → fewer independent control draws → less noise cancellation when averaging → larger variance of the control mean → larger variance of treatement effect.  -->
  - **How many neighbours, $M$.** More neighbours: less variance, more bias.
  - **Caliper**: refuse any match beyond a maximum distance. Better to drop a treated unit than to match it to something unlike it.

<!--
## Does replacement really cost variance?

Simulated $ATT$, one covariate, $N_T = 100$--$150$, 500 replications. True effect $= 2$.

\begin{center}
\small
\begin{tabular}{lcccccc}
\toprule
& \multicolumn{2}{c}{\textbf{with replacement}} & \multicolumn{2}{c}{\textbf{without}} & \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Control pool & bias & SD & bias & SD & $\sum_j K_j^2$ \\
\midrule
Abundant, good overlap      & $-0.01$ & $0.148$ & $-0.01$ & $0.145$ & $107$ vs $100$ \\
Scarce, good overlap        & $0.00$  & $0.179$ & $0.01$  & $0.158$ & $200$ vs $100$ \\
Treated clustered, thin there & $0.00$ & $0.172$ & $0.00$  & $0.140$ & $184$ vs $100$ \\
\midrule
Scarce \emph{and} poor overlap & $0.07$ & $0.176$ & $\mathbf{1.12}$ & $\mathbf{0.374}$ & $408$ vs $150$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- **Mechanism.** Control $j$ used $K_j$ times enters $K_j$ times; the control-term variance scales with $\sum_j K_j^2$, smallest when reuse is spread evenly.
- Rows 1--3 are the textbook trade-off, and the cost is **small** ($2$--$23\%$ more SD). Row 4 **reverses** it: with controls scarce, without-replacement takes bad matches and loses on *both*.
- **Default to with replacement**; $K_j$ is also why you need the Abadie--Imbens variance estimator.
-->


## Multiple dimensions and the distance metric

With several covariates, "closest" requires a definition --- and the choice changes the answer.

- **Euclidean**: $\sqrt{\sum_k (X_{ik}-X_{jk})^2}$. Scale-dependent: a variable in won swamps one measured as a ratio.
- **Normalized Euclidean**: divide each covariate by its standard deviation. Fixes scale, ignores correlation.
- **Mahalanobis**: $\sqrt{(X_i - X_j)' \Sigma^{-1}(X_i-X_j)}$. Standardizes scale **and** accounts for correlation between covariates. The default choice.

\vspace{0.3em}

- Why correlation matters: if size and leverage are highly correlated, Euclidean distance double-counts what is essentially one dimension.
<!-- - Dropping a caliper on top of any of these is a **common-support restriction**, and it changes your estimand: you are now describing only the treated units that had a counterpart.-->

# 7. Inference with Matching

## Why the usual standard errors fail

- Matched observations are **dependent by construction**: the same control may serve several treated units, and the matches were chosen *using the data*.
- Conventional standard errors treat the matched sample as if it were drawn at random. It was not, and they understate uncertainty.
- Abadie and Imbens (2008) show conventional standard errors is invalid for matching estimators 
<!-- - Bootstrapping does **not** fix it. Abadie and Imbens (2008) show the bootstrap is invalid for matching estimators --- the estimator is not smooth enough in the data. -->
- Use the **Abadie--Imbens variance estimator**, which prices in how many times each control was reused.
- Practical note: check what your software reports *by default*. Several packages report the naive standard error unless told otherwise.

## The Abadie--Imbens variance estimator

Let $\hat\tau_i = Y_i - \frac{1}{M}\sum_{m=1}^{M} Y_{j_m(i)}$ be unit $i$'s **own** effect, so $\widehat{ATT} = \frac{1}{N_T}\sum_{i:D_i=1}\hat\tau_i$ is their average. $K_j$ counts how often control $j$ is used.

$$
\hat\sigma^2_{ATT} \;=\;
\underbrace{\frac{1}{N_T}\sum_{i\,:\,D_i=1}\big(\hat\tau_i - \widehat{ATT}\big)^2}_{\text{(a) spread of the unit-level effects}}
\;+\;
\underbrace{\frac{1}{N_T}\sum_{j\,:\,D_j=0}\frac{K_j(K_j-1)}{M^2}\,\hat\sigma^2(X_j)}_{\text{(b) penalty for reusing controls}}
$$

$$
\mathrm{SE}\big(\widehat{ATT}\big) \;=\; \sqrt{\hat\sigma^2_{ATT} \,\big/\, N_T}
$$

- So term (a) is simply the **sample variance of the $\hat\tau_i$**. Alone, it is the without-replacement formula: with $K_j \in \{0,1\}$ term (b) vanishes.
- $\hat\sigma^2(X_j) = \mathrm{var}(Y \mid X_j, D{=}0)$, the outcome variance among controls. <!-- Estimate it by matching *controls to controls*: $\tfrac{1}{2}(Y_j - Y_\ell)^2$. -->
- Here $M$ is **fixed** --- the clean $K_j(K_j{-}1)$ split needs it. 
<!-- With ties, software reweights matches by $1/M_i$. --> 

<!-- 
## Reading the correction term

Rewrite the estimator as a weighted sum. Control $j$ carries weight $K_j/M$:

$$
\widehat{ATT} \;=\; \frac{1}{N_T}\sum_{i\,:\,D_i=1} Y_i \;-\; \frac{1}{N_T}\sum_{j\,:\,D_j=0} \frac{K_j}{M}\,Y_j
$$

So $j$'s noise contributes $\big(K_j/M\big)^2\sigma^2(X_j)$. Split the square:
$\;K_j^2 = \underbrace{K_j}_{\text{inside term (a)}} + \underbrace{K_j(K_j-1)}_{\text{term (b)}}$

- Term (a) treats the $\hat\tau_i$ as independent, so it counts each appearance's **own** variance once --- the diagonal.
- Term (b) is the **off-diagonal**: $K_j(K_j{-}1)$ pairs of *distinct appearances of the same number*, whose covariance is $\sigma^2(X_j)$ --- not the zero a naive SE assumes.
- Only **controls** get reused --- hence the sum over $D_j{=}0$; each treated unit appears once. $M^2$ is that weight, squared.
- $K_j \leq 1$ $\Rightarrow$ no pairs $\Rightarrow$ no correction. That is the without-replacement case.

## Where does $\hat\sigma^2(X_j)$ come from?

**Matching controls to controls**: for each control $j$, find the *other control* $\ell$ nearest in $X$ --- not a treated unit, a second control.

- One observation cannot give you a variance. But if $X_\ell \approx X_j$, then $Y_j$ and $Y_\ell$ are two draws from the same conditional distribution:
$$
Y_j - Y_\ell \;=\; \underbrace{\mu^0(X_j) - \mu^0(X_\ell)}_{\approx\, 0 \text{ if the match is close}} \;+\; \varepsilon_j - \varepsilon_\ell
\qquad\Longrightarrow\qquad
E\big[(Y_j - Y_\ell)^2\big] \;\approx\; 2\,\sigma^2(X_j)
$$

- Hence the $\tfrac{1}{2}$. Drop it and you estimate $2\sigma^2$: simulated, $8.01$ when the truth is $4$.
- **One pair is a hopeless variance estimate** --- one degree of freedom. It need not be good: term (b) *sums* over all controls, and noisy-but-unbiased pieces average out.
- If the nearest control is *not* close, the mean difference fails to cancel and inflates the estimate. The error runs **upward**, so standard errors err conservative.


## Checking it against the truth

$N_T = 200$, $N_C = 400$, $M=1$ with replacement, 1,500 replications. "Truth" is the Monte Carlo standard deviation of $\widehat{ATT}$.

\begin{center}
\begin{tabular}{lcc}
\toprule
 & standard error & ratio to truth \\
\midrule
Monte Carlo truth                       & $0.1230$ & --- \\
Term (a) only \emph{(naive)}            & $0.1003$ & $0.82$ \\
Terms (a) $+$ (b), $\sqrt{\hat\sigma^2/N_T}$ & $\mathbf{0.1224}$ & $\mathbf{1.00}$ \\
\midrule
Forgetting to divide by $N_T$           & $1.7316$ & $14.07$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- The full estimator is essentially exact. Dropping term (b) **understates the standard error by 18\%** --- every $t$-statistic too big by a fifth.
- $\hat\sigma^2_{ATT}$ is a variance *per treated unit*, not the variance of the estimator. **You must still divide by $N_T$.** Miss that and you are off by a factor of $\sqrt{N_T}$.

--> 

## In Stata

Cunningham's training data: 10 trainees, 20 non-trainees, matched on `age` and `gpa`.

\footnotesize
```stata
use https://github.com/scunning1975/mixtape/raw/master/training_inexact.dta, clear

* Minimized euclidean distance with usual variance and robust
teffects nnmatch (earnings age gpa) (treat), atet nn(1) metric(eucl) vce(iid)
teffects nnmatch (earnings age gpa) (treat), atet nn(1) metric(eucl) generate(match1)

* Minimized Mahalanobis distance with usual variance and robust
teffects nnmatch (earnings age gpa) (treat), atet nn(1) metric(maha) vce(iid)
teffects nnmatch (earnings age gpa) (treat), atet nn(1) metric(maha) generate(match2)
```
\normalsize

- `(earnings age gpa)` is outcome-then-covariates; `(treat)` is the treatment. `atet` asks for the $ATT$, `nn(1)` for one neighbour.
- `vce(iid)` is the constant-variance standard error. **Drop it and you get Abadie--Imbens** --- that is Stata's default, so the second command of each pair is the one to report.
- `generate(match1)` writes each unit's matched partner back into the data. Always look at it.

<!--

## The same thing in R

\footnotesize
```r
d <- haven::read_dta("https://github.com/scunning1975/mixtape/raw/master/training_inexact.dta")
Y <- d$earnings; D <- d$treat; X <- as.matrix(d[, c("age", "gpa")])

nnmatch <- function(Z) {                             # Z: covariates in the chosen metric
  it <- which(D == 1); ic <- which(D == 0); nT <- length(it)
  dm  <- as.matrix(dist(Z))                          # all pairwise distances
  j   <- ic[max.col(-dm[it, ic], ties.method = "first")]    # each treated unit's match
  tau <- Y[it] - Y[j]; att <- mean(tau)              # unit-level effects, then the ATT
  K   <- as.integer(table(factor(j, levels = ic)))   # how often each control is reused
  cc  <- dm[ic, ic]; diag(cc) <- Inf                 # control-to-control matches
  s2  <- 0.5 * (Y[ic] - Y[ic][max.col(-cc, ties.method = "first")])^2   # sigma^2(X_j)
  se  <- function(s) sqrt((sum((tau - att)^2) + sum(K * (K - 1) * s)) / nT^2)
  c(ATT = att, se_iid = se(rep(mean(s2), length(ic))), se_AI = se(s2))
}
W <- solve(chol(cov(X)))          # whiten: Mahalanobis = Euclidean on X %*% W
rbind(euclidean = nnmatch(X), mahalanobis = nnmatch(X %*% W))
```
\normalsize

- `se_iid` forces one common $\hat\sigma^2$; `se_AI` lets it vary with $X_j$. The two differ only through term (b).
- Package route, identical to the last column: `Matching::Match(Y, Tr, X, M = 1, estimand = "ATT", Weight = 2, Var.calc = 1)`.
-->


## Two metrics, two answers

\begin{center}
\small
\begin{tabular}{lcccc}
\toprule
 & $\widehat{ATT}$ & \texttt{vce(iid)} & \textbf{Abadie--Imbens} & distinct controls used \\
\midrule
Euclidean    & $1607.5$ & $38.7$  & $38.7$  & 10 of 20 \\
Mahalanobis  & $1315.0$ & $325.5$ & $590.1$ & \textbf{3} of 20 \\
\midrule
\emph{Truth} & $1607.5$ & & & \\
\bottomrule
\end{tabular}
\end{center}

<!-- The data are built: $Y^0 = 4000 + 225\cdot\texttt{age}$, $Y^1 = 5000 + 250\cdot\texttt{age}$, one control's earnings knocked down by $1{,}750$. `gpa` enters nowhere. -->

- `age` has standard deviation $9.1$ and `gpa` only $0.57$, so raw Euclidean distance is **matching on age and ignoring gpa** --- the scale problem from §6, Nearest-Neighbour Matching. Every trainee gets a *different* control: $K_j \leq 1$, term (b) is zero, both standard errors coincide.
- Mahalanobis makes `gpa` count --- and `gpa` is noise. **Three controls end up serving all ten trainees**; one is used five times.
- Term (a) alone still reports $199.8$. The reuse correction triples it.


# 8. Bias Correction

## The matching discrepancy is bias, not noise

Matching was only *approximate*: $X_{j(i)} \approx X_i$, never $=$. What does the leftover gap cost?

- Abadie--Imbens (2006): More data does not help much.

<!-- - Abadie--Imbens (2006): with $k$ **continuous** covariates the matching discrepancy is $O_p(N^{-1/k})$, while the standard error is $O_p(N^{-1/2})$.

\begin{center}
\begin{tabular}{lccc}
\toprule
$k$, the number of continuous covariates & $1$ & $2$ & $3$ \\
\midrule
bias relative to the standard error & $N^{-1/2}$ & $N^{0}$ & $N^{1/6}$ \\
\bottomrule
\end{tabular}
\end{center}

- $k=1$: negligible. $k=2$: bias and noise stay the same order *forever*. $k\geq 3$: the bias **dominates**, and no amount of data fixes the $t$-statistic.
- More data is the natural remedy, and the rate says it barely works: with $k=4$, halving the discrepancy takes $16\times$ the sample. --> 
- Usually you cannot get more data anyway. Abadie--Imbens (2011) is the second-best answer: **model the bias and subtract it.**

## An outcome model for the missing $Y^0$

$$
\begin{aligned}
\mu^0(x) &= E\big[\,Y \mid X{=}x, D{=}0\,\big] \;=\; \underbrace{E\big[\,Y^0 \mid X{=}x\,\big]}_{\text{unconfoundedness}} \\
\mu^1(x) &= E\big[\,Y \mid X{=}x, D{=}1\,\big] \;=\; \underbrace{E\big[\,Y^1 \mid X{=}x\,\big]}_{\text{unconfoundedness}} \\[0.3em]
Y_i &= \mu^{D_i}(X_i) + \varepsilon_i, \qquad E\big[\,\varepsilon_i \mid X_i, D_i\,\big] = 0
\end{aligned}
$$

- The second equality on each line **is** unconfoundedness. It is what lets a regression fitted on controls speak about treated units.
<!-- - This also finally *defines* $\varepsilon_i$: the deviation of $Y_i$ from its own group's conditional mean. It is the same $\varepsilon$ that $\hat\sigma^2(X_j)$ estimated in §7, Inference with Matching. -->
- So far, all definition. The **assumption** arrives only when we choose a functional form for $\hat\mu^0$.

## Where the bias hides

Substitute $Y_i = \mu^{D_i}(X_i)+\varepsilon_i$ into $\widehat{ATT}=\frac{1}{N_T}\sum_{D_i=1}(Y_i - Y_{j(i)})$, subtract the true $ATT$, then add and subtract $\frac{1}{N_T}\sum_{D_i=1}\mu^0(X_i)$:

$$
\begin{aligned}
\widehat{ATT} - ATT \;=\;& \underbrace{\tfrac{1}{N_T}\textstyle\sum_{D_i=1}\big(\mu^1(X_i) - \mu^0(X_i) - ATT\big)}_{\text{(1) sampling variation in \emph{which} units got treated}} \\
+\;& \underbrace{\tfrac{1}{N_T}\textstyle\sum_{D_i=1}\big(\varepsilon_i - \varepsilon_{j(i)}\big)}_{\text{(2) mean-zero noise}} \\
+\;& \underbrace{\tfrac{1}{N_T}\textstyle\sum_{D_i=1}\big(\mu^0(X_i) - \mu^0(X_{j(i)})\big)}_{\text{(3) \textbf{matching bias}}}
\end{aligned}
$$

- Scaled by $\sqrt{N_T}$, rows (1) and (2) converge to mean-zero normals. They are noise, and they average away.
- Row (3) does not. Its summands are **signed**: if the nearest control is systematically older, poorer, smaller, then every term carries the same sign.

## Two things make the bias big

$$
\text{Matching bias} \;=\; E\big[\,\mu^0(X_i) - \mu^0(X_{j(i)}) \,\big|\, D=1\,\big]
$$

- Read it as a product of two quantities, and it is zero if **either** is:
  - **How bad the matches are** --- the size of $X_i - X_{j(i)}$. Perfect matching kills it.
  - **How steep $\mu^0$ is** --- the heterogeneity of $Y^0$ across strata of $X$. A flat $\mu^0$ kills it too.
- Match a 13-year-old to a 15-year-old and nothing is lost *unless* $Y^0$ differs between 13- and 15-year-olds.
- Note which potential outcome appears. For the $ATT$ it is $Y^0$ that is missing and imputed, so it is $\mu^0$ --- and only $\mu^0$ --- that has to be modelled.

## The Abadie--Imbens procedure, in five steps

1. **Match.** Nearest neighbour on your metric; form $\hat\delta_i = Y_i - Y_{j(i)}$ for each treated unit.
2. **Regress.** OLS on the **matched control units only**: $\;Y_{j} = \alpha + \beta X_{j} + e_j$.
3. **Predict.** $\hat\mu^0(X_i) = \hat\alpha + \hat\beta X_i$ and $\hat\mu^0(X_{j(i)}) = \hat\alpha + \hat\beta X_{j(i)}$.
4. **Subtract** the estimated bias:

$$
\widehat{ATT}^{BC} = \frac{1}{N_T}\sum_{D_i=1}\Big[\,\big(Y_i - Y_{j(i)}\big) - \big(\hat\mu^0(X_i) - \hat\mu^0(X_{j(i)})\big)\Big]
$$

5. **Report an Abadie--Imbens standard error** (§7, Inference with Matching). The correction changes the estimate, not the reuse problem.

- Step 2 never touches a treated outcome: $\hat\mu^0$ is fitted where $Y^0$ is *observed*, then transplanted. Unconfoundedness is what permits the transplant.
<!-- - Cunningham's note: Abadie--Imbens's theory uses all controls, but their estimation section --- and every implementation since --- uses the **matched** controls.-->

## Only the gap has to be right

- The estimator uses only the **difference** $\hat\mu^0(X_i) - \hat\mu^0(X_{j(i)})$. With one covariate and a linear fit that is simply $\hat\beta\,(X_i - X_{j(i)})$.
- So $\hat\mu^0$ must be accurate *over the matching gap*, not globally. Get the level wrong everywhere and the correction is unaffected --- $\hat\alpha$ cancels.
- **This is the whole reason bias correction is more credible than regression alone.** Regression extrapolates $\hat\mu^0$ across the entire covariate space; bias correction only interpolates it across a gap that matching already made small.
- Division of labour: **matching does the heavy lifting on comparability; regression cleans up the remainder.**
- Corollary: the worse your matches, the more work the regression does, and the more the method degenerates into regression.

## Ten workers

\small
\begin{center}
\begin{tabular}{llrrrrc}
\toprule
Unit & Worker & $Y^1$ & $Y^0$ / $\widehat{Y^0}$ & $\hat\delta_i$ & Age & $D$ \\
\midrule
1 & Andy  & \$200 & \$195 & \$5   & 23 & 1 \\
2 & Betty & \$250 & \$225 & \$25  & 27 & 1 \\
3 & Chad  & \$150 & \$140 & \$10  & 29 & 1 \\
4 & Doug  & \$300 & \$195 & \$105 & 22 & 1 \\
\midrule
5 & Edith & --- & \$225 & & 27 & 0 \\
6 & Fred  & --- & \$500 & & 32 & 0 \\
7 & Gina  & --- & \$200 & & 26 & 0 \\
8 & Hank  & --- & \$190 & & 26 & 0 \\
9 & Inez  & --- & \$180 & & 28 & 0 \\
10 & Janet & --- & \$140 & & 29 & 0 \\
\bottomrule
\end{tabular}
\end{center}
\normalsize

- Andy (23) and Doug (22) both match Gina and Hank, the 26-year-olds: $\widehat{Y^0} = \tfrac{200+190}{2} = 195$. Betty and Chad match exactly.
- $\widehat{ATT} = (5 + 25 + 10 + 105)/4 = \$36.25$. Fred and Inez are never used.
- Both inexact matches point the same way --- the match is *older* than the trainee. 

## Steps 2--4, worked through

**Step 2**, over the four *matched* controls --- Edith $(27,225)$, Gina $(26,200)$, Hank $(26,190)$, Janet $(29,140)$:
$\;\hat\mu^0(\texttt{age}) = 683.75 - 18.33\,\texttt{age}$.

\small
\begin{center}
\begin{tabular}{lccrrr}
\toprule
Worker & Age & Match age & $\hat\delta_i$ & $\hat\beta(X_i - X_{j(i)})$ & $\hat\delta_i^{BC}$ \\
\midrule
Andy  & 23 & 26 & \$5   & $\$55.0$ & $-\$50.0$ \\
Betty & 27 & 27 & \$25  & $\$0$    & $\$25.0$ \\
Chad  & 29 & 29 & \$10  & $\$0$    & $\$10.0$ \\
Doug  & 22 & 26 & \$105 & $\$73.3$ & $\$31.7$ \\
\bottomrule
\end{tabular}
\end{center}
\normalsize

$\widehat{ATT} = \$36.25 \;\longrightarrow\; \widehat{ATT}^{BC} = \$4.17$.

- Exact matches get **zero** correction by construction. Only the two bad matches move.
- $\hat\beta < 0$: among *these four* controls earnings fall with age. Refit on all six and Fred $(32, \$500)$ flips it to $\hat\beta = +43.7$, giving $\widehat{ATT}^{BC} = \$112.6$.
- Four points and one outlier. The "small" regression is doing an enormous amount of work --- that is the method's exposure.

<!--
## Which regression? Underfit, correct, overfit

```{r}
#| echo: true
mcspec <- function(nn, R = 300) {                 # DGP truth: ATT = 2
  set.seed(975); out <- matrix(NA, R, 4)
  for (r in 1:R) {
    X <- rnorm(nn); D <- rbinom(nn, 1, plogis(1.5 * X))
    Y <- 2 * D + 3 * X + 2 * X^2 + rnorm(nn)
    it <- which(D == 1); ic <- which(D == 0)
    j  <- ic[max.col(-abs(outer(X[it], X[ic], "-")), ties.method = "first")]
    m  <- unique(j)                                   # the MATCHED controls
    bc <- function(f) {
      fit <- lm(f, data.frame(X = X[m], Y = Y[m]))
      p   <- function(z) predict(fit, data.frame(X = z))
      mean(Y[it] - Y[j] - (p(X[it]) - p(X[j])))       # step 4
    }
    out[r, ] <- c(mean(Y[it] - Y[j]), bc(Y ~ X), bc(Y ~ X + I(X^2)),
                  bc(Y ~ X + I(X^2) + I(X^3) + I(X^4)))
  }
  setNames(colMeans(out), c("raw", "underfit", "correct", "overfit"))
}
round(rbind(n_500 = mcspec(500), n_2000 = mcspec(2000)), 3)
```

## Reading the specification results

\begin{center}
\small
\begin{tabular}{lcccc}
\toprule
 & Raw & Underfit & Correct & Overfit \\
 & (no correction) & $X$ & $X, X^2$ & $X,\dots,X^4$ \\
\midrule
$n=500$   & $2.349$ & $2.255$ & $\mathbf{1.995}$ & $2.002$ \\
\emph{s.d.} & $0.315$ & $0.271$ & $\mathbf{0.196}$ & $0.374$ \\
\midrule
$n=2{,}000$ & $2.154$ & $2.115$ & $\mathbf{1.993}$ & $1.994$ \\
\emph{s.d.} & $0.137$ & $0.122$ & $\mathbf{0.098}$ & $0.110$ \\
\bottomrule
\end{tabular}
\end{center}

- **Underfitting still helps** --- $2.154 \to 2.115$ --- but only a little. A wrong $\hat\mu^0$ removes only the part of the gap it can see.
- **Overfitting** gets the mean right and pays in variance: at $n=500$ the standard deviation nearly doubles, $0.196 \to 0.374$.
- Cunningham's advice, which the table supports: lean toward a few extra polynomials and interactions rather than too few --- but guard against overfitting. Ben-Michael, Feller and Rothstein (2021) use *ridge* for exactly this reason.
- Lin, Ding and Han (2023): read Abadie--Imbens (2011) as a precursor to double machine learning. A flexible, regularized step 2 is the modern version.
-->

## What the correction does and does not do

- It converts a *matching* problem into a *much easier regression* problem. It does not eliminate the regression.
- It is honest about its own assumption: you must name a functional form for $\hat\mu^0$, and the ten-worker example shows how much that choice can move the answer.
- It does **not** repair the matches. 
<!-- In the training data of §7, Inference with Matching, it moves Mahalanobis from $1315.0$ to $1607.5$, but three controls still carry all ten comparisons and the Abadie--Imbens standard error still says so.-->
- Practical rule: report the raw and the bias-corrected estimate side by side. A large gap between them is a **diagnostic** --- it says your matches were poor and your outcome model is now load-bearing.

<!--
## Bias correction in software

**Stata** --- add `biasadj()` with the covariates whose gap you want priced out:

```stata
teffects nnmatch (earnings age gpa) (treat), atet nn(1) ///
    metric(maha) biasadj(age gpa)
```

**R**

```r
Matching::Match(Y, Tr, X, M = 1, estimand = "ATT", Weight = 2,
                Var.calc = 1, BiasAdjust = TRUE)
```

- Both fit step 2 on the **matched** controls. On the ten workers, `Match(BiasAdjust = TRUE)` returns exactly $\$4.167$, and the Abadie--Imbens standard error falls from $20.3$ to $16.2$.
- `biasadj()` takes its own variable list --- it need not be the list you matched on, and adding squares and interactions there is exactly the flexibility the last slide argued for.
--> 

# Wrapping Up

## Summary [Part 1]: the assumption

- **Assignment mechanisms resolve bias, not models.** The SDO decomposition is an identity; only a claim about how treatment was assigned deletes its bias terms.
- **Covariates come in four kinds**: confounders (include), colliders (never), precision covariates (optional), mediators (never). Pre-treatment timing is your best protection.
- **Unconfoundedness**, $(Y^1,Y^0) \indep D \mid X$: randomization *within* strata of $X$. Untestable; defended with institutional knowledge, and Imbens's third defence is the strongest in finance.
- **Common support** is a *separate* assumption --- and unlike unconfoundedness, it is **checkable**. It is also asymmetric: the $ATT$ needs only $e(X) < 1$, the $ATE$ needs both ends.
- **Subclassification** makes the estimand a choice about weights. On the Titanic: $SDO = 35.4$ pp, but $ATE = 19.6$, $ATT = 23.8$, $ATU = 18.9$.
- The **curse of dimensionality** is structural, not a small-sample nuisance. One missing passenger killed two of three parameters.

## Summary [Part 2]: matching, and what it costs

- **Matching** is explicit imputation: replace the missing $Y^0_i$ with the observed outcome of a comparable control. Exact matching is the honest version, and it fails on continuous $X$.
- **Nearest-neighbour matching** buys feasibility with a **matching discrepancy**, and that discrepancy is *signed*, not noise --- treated units with high $X$ tend to be matched downward. It is bias.
- **The bias shrinks slower than the standard error**: $O_p(N^{-1/k})$ against $O_p(N^{-1/2})$. At $k = 2$ they are the same order; at $k \geq 3$ the bias dominates and the $t$-statistic is not valid.
- **Two separate repairs.** Abadie--Imbens *variance* fixes inference: matching with replacement makes the comparisons dependent, and the $K_j(K_j-1)$ term prices that in. Abadie--Imbens *bias correction* fixes the estimate, using a regression only to price the gap $\hat\mu(X_i) - \hat\mu(X_{j(i)})$ --- a far weaker demand than modelling the level of $Y$.
- **Scale matters.** Euclidean and Mahalanobis distance are different estimators, not different implementations, and on the training data they return $1607.5$ and $1315.0$.
- Next: the same assumption, executed by collapsing $X$ to **one number**.

## References

\scriptsize

- Cunningham, S. *Causal Inference: The Remix*, Ch. 5.
- Cochran, W. (1968). The Effectiveness of Adjustment by Subclassification in Removing Bias in Observational Studies. *Biometrics* 24, 295--313.
- Imbens, G. (2004). Nonparametric Estimation of Average Treatment Effects under Exogeneity. *Review of Economics and Statistics* 86, 4--29.
- Abadie, A. and G. Imbens (2006). Large Sample Properties of Matching Estimators for Average Treatment Effects. *Econometrica* 74, 235--267.
- Abadie, A. and G. Imbens (2008). On the Failure of the Bootstrap for Matching Estimators. *Econometrica* 76, 1537--1557.
- Abadie, A. and G. Imbens (2011). Bias-Corrected Matching Estimators for Average Treatment Effects. *Journal of Business \& Economic Statistics* 29, 1--11.
- Lin, Z., P. Ding and F. Han (2023). Estimation Based on Nearest Neighbor Matching: From Density Ratio to Average Treatment Effect. *Econometrica* 91, 2187--2217.
- Angrist, J. and J.-S. Pischke (2009). *Mostly Harmless Econometrics*, Ch. 3. Princeton University Press.
