---
title: "Unconfoundedness II: Propensity Scores and Regression"
subtitle: "BUSS975: Causal Inference in Financial Research"
author: "Professor Ji-Woong Chung"
institute: "Korea University"
format:
  beamer:
    aspectratio: 169
    fontsize: 11pt
    slide-level: 2
    toc: false
    pdf-engine: pdflatex
    pdf-engine-opts:
      - "-interaction=nonstopmode"
    keep-tex: true
    include-in-header:
      text: |
        \setbeamertemplate{navigation symbols}{}
        \setbeamertemplate{footline}[frame number]
        \setbeamertemplate{itemize items}[circle]
        \setbeamertemplate{itemize subitem}[triangle]
        \setbeamercolor{frametitle}{fg=black}
        \setbeamerfont{frametitle}{series=\bfseries}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{booktabs}
        \usepackage{tikz}
        \usetikzlibrary{arrows.meta,positioning,calc}
        \tikzset{
          dnode/.style={circle,draw,thick,minimum size=9mm,inner sep=1pt,font=\small},
          unode/.style={circle,draw,thick,dashed,minimum size=9mm,inner sep=1pt,font=\small},
          bnode/.style={rectangle,draw,thick,fill=gray!25,minimum size=9mm,inner sep=3pt,font=\small},
          de/.style={-{Stealth[length=2.2mm]},thick},
          dde/.style={-{Stealth[length=2.2mm]},thick,dashed}
        }
        % shrink code input (fancyvrb) and code output (verbatim)
        \usepackage{fancyvrb}
        \fvset{fontsize=\scriptsize}
        \makeatletter
        \def\verbatim@font{\scriptsize\ttfamily}
        \makeatother
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

## Outline

1. Propensity scores
2. Regression
3. What matching cannot do
4. The NSW demonstration


# 1. Propensity Scores

## Defining the propensity score

- **Propensity score**: $e(X) = \Pr(D = 1 \mid X)$ 
  - "the conditional probability of assignment to a particular treatment given a vector of observed covariates"

- **Balancing property** (Rosenbaum--Rubin 1983): $X \perp D \mid \Pr(D = 1 \mid X)$
  - $E[X \mid D=1, \Pr(D=1 \mid X)] = E[X \mid D=0, \Pr(D=1 \mid X)]$
  - Covariate distributions are similar between treatment and control conditional the propensity score.

- **Propensity score theorem** (Rosenbaum--Rubin 1983): if $(Y^1,Y^0) \indep D \mid X$, then
$$
(Y^1, Y^0) \;\indep\; D \;\mid\; e(X)
$$
  - Conditioning on a **single number** does what conditioning on all of $X$ does. Ten covariates collapse to one scalar --- the curse of dimensionality defeated, for *conditioning*.




<!--
## Why the score balances

By definition, $\Pr(D{=}1 \mid X,\, e(X)) = e(X)$. Now compute the same probability knowing *only* the score:

$$
\Pr(D{=}1 \mid e(X)) \;=\; E\big[\,\underbrace{\Pr(D{=}1 \mid X)}_{=\;e(X)} \;\big|\; e(X)\,\big]
\;=\; E\big[e(X) \mid e(X)\big] \;=\; e(X)
$$

- The two are equal: knowing $X$ *on top of* $e(X)$ does not change the chance of treatment. That is $D \indep X \mid e(X)$, and **no assumption was used**.
- Add unconfoundedness --- $\Pr(D{=}1 \mid Y^0,Y^1,X) = e(X)$ --- and the identical argument conditioning on $(Y^0,Y^1,e(X))$ again returns $e(X)$. The potential outcomes carry no extra information either.
- **In one sentence**: $e(X)$ is a *sufficient statistic for the assignment mechanism*. Once you know how biased the coin is, knowing which $X$ made it that biased says nothing more about how it landed.
- What is **not** claimed: units with the same score are not similar. Only their *distributions* match.
-->

<!--
## Three cells, one number

\begin{center}
\small
\begin{tabular}{llrrrr}
\toprule
Cell & $X$ = (size, leverage) & $N$ & $e(X)$ & treated & control \\
\midrule
A & Large, Low  & $200$ & $0.2$ & $40$  & $160$ \\
B & Small, High & $400$ & $0.2$ & $80$  & $320$ \\
C & Large, High & $400$ & $0.6$ & $240$ & $160$ \\
\midrule
  & Total       & $1{,}000$ & & $360$ & $640$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- Read the score as a frequency: *what fraction of firms like this are treated?* In cell A, $40$ of $200$, so $e = 0.2$.
- Cells **A and B share a score** but are *opposites* on $X$. The stratum $e = 0.2$ pools $600$ firms: $120$ treated, $480$ control.
- Of those $120$ treated, $40$ come from A --- $33.3\%$. Of the $480$ controls, $160$ come from A --- $33.3\%$. **Identical**, and not an approximation: both cells drew from the same $0.2$ coin, so pooling preserves their $1{:}2$ ratio on *both* sides.

## The balance table

\begin{center}
\small
\begin{tabular}{lrrr}
\toprule
 & Treated & Control & Gap (pp) \\
\midrule
\emph{Raw sample} ($n = 1{,}000$) & & & \\
\quad \% Large             & $77.8$ & $50.0$ & $27.8$ \\
\quad \% High leverage     & $88.9$ & $75.0$ & $13.9$ \\
\midrule
\emph{Within} $e = 0.2$ (cells A and B) & & & \\
\quad \% Large             & $33.3$ & $33.3$ & $\mathbf{0.0}$ \\
\quad \% High leverage     & $66.7$ & $66.7$ & $\mathbf{0.0}$ \\
\midrule
\emph{Within} $e = 0.6$ (cell C) & & & \\
\quad \% Large             & $100.0$ & $100.0$ & $\mathbf{0.0}$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- The raw sample is badly imbalanced on **both** covariates. Conditioning on one number --- never on size, never on leverage --- balances **both**, exactly.
- This is the whole appeal of the score, and it is also the check you run in your own data: split $\hat e$ into deciles and put this table in the paper.
-->

## Estimating the score

- You know the **true** score only when you ran the experiment yourself. Otherwise, it must be estimated: $\Pr(D_i {=} 1 \mid X_i) = F(\beta X_i)$, fitted by **logit or probit**.
- Judge the specification by **balance achieved**, not by fit. A score with a high pseudo-$R^2$ but poor balance is a bad score; the reverse is fine.
- Fit, check balance, add interactions and polynomials where balance fails, refit. A legitimate specification search, because it never looks at $Y$ --- the whole stage is **blind to the outcome**, which protects against fishing. Say so in the paper.
- **The score creates no support that $X$ lacks.** If a covariate cell is all-treated or all-control, collapsing $X$ to a scalar hides that; it does not fix it.


## Checking balance: two numbers, not a $t$-test

The balancing property is checkable. Report it covariate by covariate, before and after.

$$
\text{SMD} \;=\; \frac{\bar X_1 - \bar X_0}{\sqrt{\big(s_1^2 + s_0^2\big)/2}},
\qquad\qquad
\text{VR} \;=\; \frac{s_1^2}{s_0^2}
$$

- **Standardized mean difference.** $|\text{SMD}| < 0.1$ is good.
  - Keep the denominator **fixed at its raw-sample value** when recomputing after weighting, or the yardstick moves with the thing you are measuring.
- **Variance ratio.** Means can match while spreads do not. Rubin: stay inside $[0.5,\,2]$, ideally $[0.8,\,1.25]$.
- **Do not use $t$-tests.** Balance is a property of *this sample*, not a population hypothesis --- and $p$ confounds imbalance with sample size (Imai, King and Stuart 2008).

## A balance plot: means are not enough

```{r}
#| echo: false
#| fig-width: 9.6
#| fig-height: 3.25
set.seed(975)
n <- 5000
size <- rnorm(n); lev <- rnorm(n); roa <- rnorm(n)
age <- rnorm(n); vol <- rnorm(n); rnd <- rnorm(n)
X  <- cbind(Size=size, Leverage=lev, ROA=roa, Age=age, Volatility=vol, `R&D`=rnd)
D  <- rbinom(n, 1, plogis(0.9*size + 0.7*lev - 0.6*roa + 0.25*age + 0.7*vol^2 - 0.7))
eh <- predict(glm(D ~ size+lev+roa+age+vol, family = binomial), type = "response")
w  <- ifelse(D == 1, 1/eh, 1/(1-eh))
wm <- function(x,u) sum(u*x)/sum(u)
wv <- function(x,u){ m<-wm(x,u); sum(u)*sum(u*(x-m)^2)/(sum(u)^2-sum(u^2)) }
den <- apply(X,2,function(x) sqrt((var(x[D==1])+var(x[D==0]))/2))
smd <- function(u) sapply(1:ncol(X), function(k){ x<-X[,k]
         (wm(x[D==1],u[D==1])-wm(x[D==0],u[D==0]))/den[k] })
vr  <- function(u) sapply(1:ncol(X), function(k){ x<-X[,k]
         wv(x[D==1],u[D==1])/wv(x[D==0],u[D==0]) })
one <- rep(1,n); k <- ncol(X); y <- k:1; RED <- "#B2182B"
par(mfrow=c(1,2), mar=c(4,7,2.2,1), cex=0.8)
plot(NULL, xlim=c(0,0.78), ylim=c(0.5,k+0.5), yaxt="n", ylab="",
     xlab="| standardized difference |", main="Means")
axis(2, at=y, labels=colnames(X), las=1); abline(h=y, col="grey92")
abline(v=0.1, lty=2, col=RED)
points(abs(smd(one)), y, pch=1,  cex=1.3, col="grey35")
points(abs(smd(w)),   y, pch=19, cex=1.3, col=RED)
legend("right", c("raw","IPW weighted"), pch=c(1,19),
       col=c("grey35",RED), bty="n", cex=0.95)
plot(NULL, xlim=c(0.6,3.0), ylim=c(0.5,k+0.5), yaxt="n", ylab="",
     xlab="variance ratio  (treated / control)", main="Spreads")
axis(2, at=y, labels=colnames(X), las=1); abline(h=y, col="grey92")
abline(v=1, col="grey40"); abline(v=c(0.8,1.25), lty=2, col=RED)
points(vr(one), y, pch=1,  cex=1.3, col="grey35")
points(vr(w),   y, pch=19, cex=1.3, col=RED)
```

\vspace{-0.3em}
\footnotesize
Selection here is linear in `Size`, `Leverage`, `ROA`, `Age` --- and **quadratic in `Volatility`**. The score was fitted with linear terms only.

## Reading the balance plot

- **Left panel, the success.** Four covariates start badly imbalanced --- `Size` at $0.65$, `Leverage` and `ROA` near $0.49$. After weighting, every one is under $0.02$. This is the plot referees want, and it is the one most papers stop at.
- **Left panel, the trap.** `Volatility` looks *perfect* from the start: $\text{SMD} = -0.010$, and a $t$-test returns $p = 0.72$. On the evidence of means, nothing to see.
- **Right panel.** Its variance ratio is $\mathbf{2.27}$ --- treated firms have more than twice the spread. Extreme volatility drives selection in *both* tails.
- **Weighting makes it worse**, $2.27 \to \mathbf{2.83}$, because the fitted score never contained $\texttt{vol}^2$. Add that one term and the ratio falls to $\mathbf{1.03}$.
- The lesson of this section: judge the score by **balance achieved, not by fit** --- and balance means the whole distribution. Check squares and interactions, not just levels.


## Three things to do with the score

Once every unit has a $\hat e(X_i)$, there are three standard uses. 

1. **Diagnose.** Plot $\hat e$ separately for treated and control units. This is the common-support check, and it costs nothing.
2. **Weight.** Reweight the sample to build a pseudo-experiment --- inverse probability weighting.
3. **Match.** Pair treated to control units with similar scores, then difference --- propensity score matching.

<!--
\vspace{0.3em}

- A fourth, older use: **stratify** on the score. That is Lecture 3a, §4, Subclassification, with strata defined by deciles of $\hat e$ --- Rosenbaum and Rubin's original proposal, and still a reasonable one.
- Uses 1 and 4 are the safest. Use 2 is standard but fragile in the tails. Use 3 has a serious critique against it.
- The section closes with a fifth use that combines 2 with a regression: **doubly robust** estimation.
-->

## Visualizing overlap

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.0
set.seed(975)
n <- 4000
par(mfrow = c(1, 2), mar = c(4, 4, 2.2, 1), cex = 0.8)
for (a in c(0.8, 3.5)) {
  X  <- rnorm(n); D <- rbinom(n, 1, plogis(a * X))
  eh <- predict(glm(D ~ X, family = binomial), type = "response")
  br <- seq(0, 1, 0.05)
  h1 <- hist(eh[D == 1], breaks = br, plot = FALSE)
  h0 <- hist(eh[D == 0], breaks = br, plot = FALSE)
  plot(NULL, xlim = c(0, 1), ylim = c(0, max(h1$density, h0$density) * 1.28),
       xlab = "estimated propensity score", ylab = "density",
       main = if (a < 1) "Good overlap" else "Poor overlap")
  rect(br[-length(br)], 0, br[-1], h0$density, col = "#00000022", border = "grey40")
  rect(br[-length(br)], 0, br[-1], h1$density, col = "#B2182B55", border = "#B2182B")
  legend("top", c("control", "treated"), fill = c("#00000022", "#B2182B55"),
         bty = "n", horiz = TRUE, cex = 0.9)
}
```

\vspace{-0.4em}

\footnotesize
Right panel: a treated mass at $\hat e \approx 1$ with no controls, and a control mass at $\hat e \approx 0$ with no treated. Those units have **no counterfactual**. Any estimate covering them is extrapolation, not comparison.

## This plot belongs in every such paper

- Common support is verifiable, unlike unconfoundedness.
- What to do when it looks like the right panel:
  - **Trim** observations with extreme scores, or impose a caliper. Drop units with $\hat e \notin [0.1,\,0.9]$. A rule of thumb, but a defensible and widely accepted one (Crump, Hotz, Imbens and Mitnik, 2009).
  - **Report the change in estimand.** After trimming you describe a subpopulation, not the original one. Say how many observations went and how the treated sample shifted.
  - Consider retreating to the $ATT$, which only needs overlap in one direction.
- What *not* to do: report the estimate and stay silent. The regression will happily extrapolate into a region with no data and give you a standard error that reflects none of that.

## Inverse probability weighting

Create balanced pseudo-population by inverse 
probability weighting. Write $\hat e_i$ for $\hat e(X_i)$ and $\pi = \Pr(D{=}1)$.

\small
$$
\begin{aligned}
ATE \;&=\; E\!\left[Y_1 - Y_0\right] = E\!\left[\frac{D\,Y}{e(X)}\right] - E\!\left[\frac{(1-D)\,Y}{1 - e(X)}\right] \\
ATT \;&=\; E\!\left[Y_1 - Y_0 \middle| D = 1 \right] = \frac{1}{\pi}\,\left(E\!\left[DY\right] - E\!\left[\frac{e(X)(1-D)Y}{1 - e(X)}\right]\right) \\[1em]
\widehat{ATE}_{IPW} \;&=\; \frac{1}{n}\sum_{i=1}^{n}\left[\frac{D_i}{\hat e_i} - \frac{1 - D_i}{1 - \hat e_i}\right] Y_i \\
\widehat{ATT}_{IPW} \;&=\; \frac{1}{n_T}\sum_{i=1}^{n}\left[D_i - (1 - D_i)\,\frac{\hat e_i}{1 - \hat e_i}\right] Y_i
\end{aligned}
$$
\normalsize

- Both are **Horvitz--Thompson (1952)** forms.

## Why *that* weight for the ATE?

We prove $E\!\left[\frac{D\,Y}{e(X)}\right] = E[Y^1]$. The $E[Y^0]$ half is symmetric.

\scriptsize
$$
\begin{aligned}
E\!\left[ \frac{D \cdot Y}{e(X)} \right]
  &= E\!\left[ \frac{D \cdot Y^1}{e(X)} \right]
     && \text{switching equation (consistency): } D \cdot Y = D \cdot Y^1 \\
  &= E_X\!\left[ E\!\left( \frac{D \cdot Y^1}{e(X)} \,\Big|\, X \right) \right]
     && \text{law of total expectation} \\
  &= E_X\!\left[ \frac{1}{e(X)} \cdot E\big(D \cdot Y^1 \mid X\big) \right]
     && e(X) \text{ is fixed given } X \\
  &= E_X\!\left[ \frac{1}{e(X)} \cdot E(D \mid X) \cdot E\big(Y^1 \mid X\big) \right]
     && \text{unconfoundedness (exchangeability): } D \indep Y^1 \mid X \\
  &= E_X\!\left[ \frac{1}{e(X)} \cdot e(X) \cdot E\big(Y^1 \mid X\big) \right]
     && \text{by definition, } E(D \mid X) = e(X) \\
  &= E_X\big[ E(Y^1 \mid X) \big] \;=\; E[Y^1]
     && \text{cancel; law of total expectation}
\end{aligned}
$$

## Why *that* weight for the ATT?

\scriptsize
$E[Y^1 \mid D{=}1] = E[Y \mid D{=}1] = \frac{E[DY]}{E[D]}$.
$$
\begin{aligned}
E\big[Y^0 \mid D{=}1\big]
  &= E_X\big[\, E(Y^0 \mid X, D{=}1) \,\big|\, D{=}1 \,\big]
     && \text{law of total expectation, over } f(x \mid D{=}1) \\
  &= E_X\big[\, E(Y^0 \mid X, D{=}0) \,\big|\, D{=}1 \,\big]
     && \text{unconfoundedness --- swap the arm} \\
  &= E_X\!\left[\, E(Y^0 \mid X, D{=}0)\cdot\frac{e(X)}{\pi} \,\right]
     && \text{Bayes: } f(x \mid D{=}1) = e(x)f(x)/\pi \\
  &= E_X\!\left[\, \frac{E\big[(1-D)\,Y \mid X\big]}{1 - e(X)}\cdot\frac{e(X)}{\pi} \,\right]
     && E[(1{-}D)Y \mid X] = \big(1{-}e(X)\big)E[Y^0 \mid X, D{=}0] \\
  &= E\!\left[\, \frac{(1 - D)\,Y\,e(X)}{\big(1 - e(X)\big)\,\pi} \,\right]
     && \text{law of total expectation}
\end{aligned}
$$
\normalsize

Line 3: it converts an average over the *treated* covariate distribution into one over the *whole* population.

## Ten banks, split by size

Lecture 1's capital injections, now with a covariate: bank **size**.

\begin{center}
\scriptsize
\begin{tabular}{crcrrrr}
\toprule
Bank & Size & $D$ & $Y^0$ & $Y^1$ & $\tau$ & observed $Y$ \\
\midrule
$1$  & Small & $1$ & $10$ & $12$ & $2$ & $\mathbf{12}$ \\
$2$  & Small & $0$ & $8$  & $10$ & $2$ & $\mathbf{8}$  \\
$3$  & Small & $0$ & $9$  & $11$ & $2$ & $\mathbf{9}$  \\
$4$  & Small & $0$ & $11$ & $13$ & $2$ & $\mathbf{11}$ \\
$5$  & Small & $0$ & $12$ & $14$ & $2$ & $\mathbf{12}$ \\
\midrule
$6$  & Large & $1$ & $18$ & $24$ & $6$ & $\mathbf{24}$ \\
$7$  & Large & $1$ & $19$ & $25$ & $6$ & $\mathbf{25}$ \\
$8$  & Large & $1$ & $21$ & $27$ & $6$ & $\mathbf{27}$ \\
$9$  & Large & $1$ & $22$ & $28$ & $6$ & $\mathbf{28}$ \\
$10$ & Large & $0$ & $20$ & $26$ & $6$ & $\mathbf{20}$ \\
\bottomrule
\end{tabular}
\end{center}

- Small banks are **rarely** rescued, $1$ of $5$; large banks **usually** are, $4$ of $5$. So $e(\text{Small}) = 0.2$ and $e(\text{Large}) = 0.8$.
- Small banks gain $2$, large banks gain $6$. Truth: $ATE = 4$, $ATT = 5.2$.
- Only the **bold** column is observed. Comparing rescued with not gives $\mathbf{11.2}$ --- nearly three times the $ATE$.

## The ATE rebuilds the population

\begin{center}
\small
\begin{tabular}{lrrrrrr}
\toprule
Cell & $N$ & $e$ & injected & $\times\,1/e$ & controls & $\times\,1/(1-e)$ \\
\midrule
Small & $5$ & $0.2$ & $1$ & $\times 5 = \mathbf{5}$    & $4$ & $\times 1.25 = \mathbf{5}$ \\
Large & $5$ & $0.8$ & $4$ & $\times 1.25 = \mathbf{5}$ & $1$ & $\times 5 = \mathbf{5}$ \\
\midrule
Total & $10$ & & $5$ & $\mathbf{10}$ & $5$ & $\mathbf{10}$ \\
\bottomrule
\end{tabular}
\end{center}

- Bank $1$ is the **only** small bank rescued, so it must stand in for all five. Its weight is $1/0.2 = 5$.
- Every weighted entry comes back to $5$: each arm is blown back up to the **full population of ten**. That is the *pseudo-population*, and in it $D \indep X$ by construction.
- Before weighting, small banks are $20\%$ of the injected sample and $80\%$ of the controls. After weighting, $50\%$ of both --- the imbalance that caused the bias is gone.

## The ATT reweights controls only

The $ATT$ wants the **treated** distribution. The four large and one small rescued banks are already a sample from it, so they keep weight $1$. Only the controls move.

\begin{center}
\small
\begin{tabular}{lrrrr}
\toprule
Cell & controls & $\times\ e/(1-e)$ & $=$ & injected \\
\midrule
Small & $4$ & $0.25$ & $\mathbf{1}$ & $1$ \\
Large & $1$ & $4.00$ & $\mathbf{4}$ & $4$ \\
\bottomrule
\end{tabular}
\end{center}

- Multiplying the controls by $e/(1-e)$ makes their count match the injected count **exactly** --- last column.
- Bank $10$ is the lone large control, and it must cover **four** rescued banks, so it carries weight $4$. Push $e$ toward $1$ and there is nobody left to carry it: **this is where the $ATT$ fails**, and why it needs only $e(X) < 1$ while the $ATE$ needs $0 < e(X) < 1$.

## Does it work?

$$
\begin{aligned}
\widehat{E[Y^1]} &= \tfrac{1}{10}\!\left[\tfrac{12}{0.2} + \tfrac{24+25+27+28}{0.8}\right] = \tfrac{60 + 130}{10} = 19 \\
\widehat{E[Y^0]} &= \tfrac{1}{10}\!\left[\tfrac{8+9+11+12}{0.8} + \tfrac{20}{0.2}\right] = \tfrac{50 + 100}{10} = 15 \\[0.2em]
\widehat{ATT}\; &= \underbrace{\tfrac{12 + 24+25+27+28}{5}}_{\text{injected mean}\;=\;23.2} \;-\; \underbrace{\tfrac{(8+9+11+12)(0.25) + 20(4)}{4(0.25) + 1(4)}}_{\text{controls}\,\times\,\text{odds}\;=\;18.0}
\end{aligned}
$$

- Truth: $ATE = 4$, $ATT = 5.2$. The naive $SDO$ is $23.2 - 12 = \mathbf{11.2}$.
- IPW returns $19 - 15 = \mathbf{4}$ and $23.2 - 18.0 = \mathbf{5.2}$. Both exactly right.
- $ATT > ATE$ because large banks gain more, and large banks are where the injections went. **Selection on gains --- the two estimands correctly disagree.**


## Two ways to divide

Both estimators use the same weights $1/\hat e_i$. They differ **only in the denominator**.

\vspace{-0.2em}

$$
\underbrace{\widehat{E[Y^1]}_{\text{HT}} = \frac{1}{N}\sum_i \frac{D_i Y_i}{\hat e_i}}_{\text{divide by } N}
\qquad\qquad
\underbrace{\widehat{E[Y^1]}_{\text{Hájek}} = \frac{1}{\sum_i D_i/\hat e_i}\sum_i \frac{D_i Y_i}{\hat e_i}}_{\text{divide by the weights you got}}
$$


- In the **population** they agree: $E[D/e(X)] = 1$, so the weights sum to $N$.
- In a **sample** they do not. $\sum_i D_i/\hat e_i$ is only an *estimate* of $N$, and it misses.
- Hájek's weights sum to one by construction, so it is always a genuine **weighted average of outcomes you actually observed**. HT carries no such guarantee.
- The default in Stata's `teffects ipw` and R's `WeightIt`.

## Example 1

Four rescued banks. **Every one has lending growth $Y = 10$**, so the answer must be $10$ for $\widehat{E[Y^1]}$. Here $N = 8$.

\begin{center}
\footnotesize
\begin{tabular}{ccrc}
\toprule
Bank & $\hat e$ & weight $1/\hat e$ & $Y$ \\
\midrule
A & $0.50$ & $2$ & $10$ \\
B & $0.50$ & $2$ & $10$ \\
C & $0.50$ & $2$ & $10$ \\
D & $0.25$ & $4$ & $10$ \\
\midrule
sum & & $\mathbf{10}$ & \\
\bottomrule
\end{tabular}
\end{center}

\vspace{-0.3em}
\small
$$
\text{HT} = \frac{20+20+20+40}{8} = \frac{100}{8} = \mathbf{12.5},
\qquad
\text{Hájek} = \frac{100}{10} = \mathbf{10} .
$$
\normalsize

- The weights sum to $10$, not $8$. HT divides by $8$ anyway, returning $10 \times \tfrac{10}{8}$ --- a number that appears nowhere in the data.


## Example 2

\begin{center}
\footnotesize
\begin{tabular}{ccrrc}
\toprule
Bank & $\hat e$ & weight & share of weight & $Y$ \\
\midrule
A & $0.50$ & $2$  & $3.6\%$ & $10$ \\
B & $0.50$ & $2$  & $3.6\%$ & $10$ \\
C & $0.50$ & $2$  & $3.6\%$ & $10$ \\
D & $\mathbf{0.02}$ & $\mathbf{50}$ & $\mathbf{89.3\%}$ & $24$ \\
\midrule
sum & & $56$ & $100\%$ & \\
\bottomrule
\end{tabular}
\end{center}

\vspace{-0.3em}
$$
\text{HT} = \frac{1260}{8} = \mathbf{157.5},
\qquad
\text{Hájek} = \frac{1260}{56} = \mathbf{22.5} .
$$

- D carries $\mathbf{89\%}$ of the weight. 
- **Normalization is necessary, not sufficient.** Check overlap separately: trim, or retreat to the $ATT$ when the trouble is at $\hat e \approx 0$.

## Propensity score matching

- Match each treated unit to controls with a **similar score**, then difference. 
- **Caliper** matching: refuse any match farther than some tolerance (say $0.001$ on the score), *discarding* the treated unit if none qualifies. **Radius** matching: average all controls inside the tolerance.
- Two problems with both:
  - The tolerance is **chosen by the researcher** (Huntington-Klein et al. 2021).
  - Dropping treated units **changes the estimand** --- the $ATT$ summarised the *original* treated sample. 
<!-- - The score is estimated, so its sampling error enters the standard errors (Abadie--Imbens 2016). -->
- King & Nielsen, 2019: Two units can share an identical score $e(X_i) = e(X_j)$ despite having completely opposite covariate profiles $X_i \neq X_j$. Matching purely on the scalar score can actually increase local covariate imbalance compared to multivariate distance matching (e.g., Coarsened Exact Matching).

<!--
## King and Nielsen (2019): which experiment are you imitating?

*Why Propensity Scores Should Not Be Used for Matching*, **Political Analysis** 27, 435--454.

- Every matching method **recreates some experiment**. Which one decides how good it can get.
  - **Completely randomized**: covariates balanced *on average*. Draw again and the treated group is $52\%$ large firms, then $48\%$ --- balanced in expectation, never in the sample you hold.
  - **Fully blocked**: randomize *within* strata, so exactly $50\%$ large firms in both arms, every draw. Strictly better on imbalance, model dependence, variance and bias.
- **PSM targets the weaker one.** The balancing property promises equal covariate *distributions* within a score stratum --- complete randomization by construction, and no more.
- Mahalanobis and coarsened exact matching target the **blocked** experiment: they compare on $X$ itself. Cells A and B share $e = 0.2$ and are opposites on $X$, so PSM cannot tell an A--A pair from an A--B pair --- but **blocking would never have made that pair.**

## The PSM paradox, and what to do instead

- Matching prunes units to improve balance. For MDM and CEM, pruning **keeps helping**: you are always dropping the worst-matched pair on $X$.
- For PSM, once the remaining sample already looks completely randomized, the pairs you prune next are **arbitrary with respect to $X$** --- the score cannot distinguish them. You are now pruning at random.
- Randomly deleting units from a randomized experiment does not improve balance. It shrinks $n$ and makes realized imbalance *noisier*, so measured imbalance, model dependence and bias start going **up**.
- King and Nielsen: *"the more balanced the data, or the more balanced it becomes by \[trimming\] some of the observations through matching, the more likely propensity score matching will degrade inferences."*
- The trap is the standard workflow --- prune, check balance, adjust the caliper, prune again. That loop can walk you into the rising region while the diagnostics look like progress.

## Their recommendation, and its scope

- **Replace PSM with a method that blocks**: Mahalanobis distance matching, or coarsened exact matching. Match on $X$, not on a scalar summary of $X$.
- **Diagnose with the pruning curve.** Plot imbalance against the number of units pruned. Falling is fine; the moment it turns up, stop --- you are in the paradox.
- **The critique is narrow, and they say so.** It targets nearest-neighbour matching *on the score*. It leaves untouched:
  - inverse probability **weighting**,
  - **stratification** on the score (Rosenbaum--Rubin's original proposal),
  - **regression adjustment**, and the score as an ingredient in doubly robust estimators.
- So the propensity score survives this paper intact. What does not survive is one particular thing people do with it.
- Recommended viewing: Gary King, "16. Matching Methods" (83 min) --- `https://youtu.be/tvMyjDi4dyg`

## A surprise: estimate the score even when you know it

- Hahn (1998); Hirano, Imbens and Ridder (2003): weighting by the **estimated** $\hat e$ is *more efficient* than weighting by the **true** $e$ --- even when the truth is handed to you.
- Students find this backwards. The intuition is one you already accept.
- Run an experiment with a fair coin, $e = 0.5$. By chance the treated group comes out $55\%$ large firms and the control group $45\%$. The true score says "balanced" --- and it is right, *about the coin*. The estimated score reads $\hat e > 0.5$ for large firms and weights that chance imbalance away.
- **You do not need to correct the coin. You need to correct the sample you actually drew.** Only $\hat e$ knows what you drew.
- Same reason you put covariates in an RCT regression even though randomization already delivered identification.
- Scope: this is about *weighting*. For **matching** on $\hat e$, estimation moves the variance in either direction and must be accounted for (Abadie--Imbens 2016).
-->

## What the score does and does not buy

- It buys **dimension reduction**, not identification. If $X$ omits a confounder, $e(X)$ omits it too.
- Standard regression silently hides lack of common support through linear extrapolation, whereas plotting $\hat{e}(X)$ immediately exposes severe positivity violations.
- A high pseudo-$R^2$ in a treatment model means covariates perfectly predict treatment assignment ($e(X) \to 0$ or $1$). This breaks the positivity assumption and causes extreme IPW variance. Conversely, a low pseudo-$R^2$ indicates that treatment assignment is nearly independent of $X$ (resembling a randomized control trial).
- IPW shifts the modeling assumption from $E[Y \mid D, X]$ (outcome equation) to $P(D = 1 \mid X)$ (treatment assignment equation).

<!--

## You need not choose: doubly robust estimation

The choice just posed --- model the **outcome** or model the **treatment** --- is not forced. Fit *both*, and only one needs to be right.

\small
$$
\widehat{ATE}_{DR} = \frac{1}{n}\sum_i \Big[ \underbrace{\hat\mu^1(X_i) - \hat\mu^0(X_i)}_{\text{regression imputation}}
+ \underbrace{\frac{D_i\big(Y_i - \hat\mu^1(X_i)\big)}{\hat e(X_i)} - \frac{(1-D_i)\big(Y_i - \hat\mu^0(X_i)\big)}{1-\hat e(X_i)}}_{\text{IPW applied to the \emph{residuals}}} \Big]
$$
\normalsize

- Read it as regression imputation with an IPW **repair term** bolted on. The repair term is ordinary IPW --- but run on the residuals $Y - \hat\mu$, not on $Y$.
- If $\hat\mu^d$ is right, the residuals are mean-zero and the repair term **vanishes** --- whatever $\hat e$ is.
- If $\hat e$ is right, the repair term is a valid IPW estimate of exactly the error the first term made, and cancels it --- whatever $\hat\mu^d$ is.
- Consistent if **either** model is correct: *doubly robust*. Two chances instead of one.
- You have met the shape before: Lecture 3a, §8, Bias Correction, is the matching version of the same trick --- a crude estimator, plus a regression that prices out its leftover error.

## Doubly robust: why the repair term cancels

Take one cell with true score $e$, true means $y^1, y^0$, and any guessed outcome model $\hat\mu^1, \hat\mu^0$. Average the bracket over the units in that cell:

$$
\underbrace{\hat\mu^1 - \hat\mu^0}_{\text{guess}}
\;+\; \underbrace{\frac{e\,(y^1 - \hat\mu^1)}{e}}_{\text{treated share} \,\times\, \text{residual}}
\;-\; \underbrace{\frac{(1-e)(y^0 - \hat\mu^0)}{1-e}}_{\text{control share}\,\times\,\text{residual}}
\;=\; y^1 - y^0
$$

- The $e$'s cancel, the $\hat\mu$'s cancel, and out drops the true effect. **Whatever you guessed.**
- That is the whole of double robustness in one line: a *correct score* makes the share of treated units in the cell exactly undo the $1/\hat e$, so the residual is added back at full weight and the guess is erased.
- Symmetrically, if the guess is right then both residuals are zero and the second and third terms disappear, leaving $\hat\mu^1 - \hat\mu^0$ --- correct regardless of $\hat e$.
- **Only one of the two cancellations has to fire.**

## Doubly robust: the three cells, one model at a time

Truth: $ATE = 2$, naive $SDO = 7$. Now deliberately break one model at a time.

\begin{center}
\small
\begin{tabular}{llccc}
\toprule
Outcome model $\hat\mu$ & Score $\hat e$ & Regression only & IPW only & \textbf{Doubly robust} \\
\midrule
wrong: flat $10 / 16$ & correct $0.2/0.2/0.6$ & $6.0$ & $2.0$ & $\mathbf{2.0}$ \\
correct cell means    & wrong: $0.36$ for all & $2.0$ & $7.0$ & $\mathbf{2.0}$ \\
wrong: flat $10 / 16$ & wrong: $0.36$ for all & $6.0$ & $7.0$ & $7.0$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- Row 1, cell A: $16 - 10 + \tfrac{0.2(14-16)}{0.2} - \tfrac{0.8(12-10)}{0.8} = 6 - 2 - 2 = 2$. Cells B and C also give exactly $2$.
- Row 2: the wrong score is a logit **with no covariates**, so $\hat e = 360/1000$ for everyone. Constant weights are no weights, and IPW hands back the naive $7$. The correct $\hat\mu$ rescues it anyway.
- Row 3 is the honest one. **Two chances is not a guarantee** --- get both wrong and you get neither. Here it collapses back to $7$.

## Doubly robust in practice

- **What it buys**: you no longer have to win the argument about which model is right. You have to win only one of the two.
- **What it does not buy**: identification. Both models are built from the same $X$. If $X$ omits a confounder, both are wrong in the same direction and row 3 is where you live.
- The score still needs **overlap**. Doubly robust estimators divide by $\hat e$ and $1-\hat e$ just as IPW does, so extreme scores hurt them the same way --- trim first.
- Cheap version you can run today: estimate $\hat e$, then run **weighted least squares** of $Y$ on $D$ and $X$ with IPW weights. Regression plus weighting is doubly robust for the $ATE$ when the outcome model is linear.
- In software: Stata `teffects aipw (y x) (d x)`; R `AIPW` or `DoubleML`. Report it beside your regression and your IPW estimate --- three numbers that agree is a real robustness check, and three that disagree tells you which model was load-bearing.

-->


<!-- ARCHIVE: simulation versions of the two hand-worked examples above.

## Checking the balancing property

Two covariates, $e(X) = \Lambda(1.0 X_1 + 0.8 X_2)$, $n = 20{,}000$. Treated-minus-control gap in each covariate:

```{r}
#| echo: true
set.seed(975)
n  <- 20000; X1 <- rnorm(n); X2 <- rnorm(n)
D  <- rbinom(n, 1, plogis(1.0 * X1 + 0.8 * X2))
eh <- predict(glm(D ~ X1 + X2, family = binomial), type = "response")
dec <- cut(eh, quantile(eh, 0:10 / 10), include.lowest = TRUE)
gap <- function(V) c(overall   = mean(V[D == 1]) - mean(V[D == 0]),
                     max_within = max(abs(tapply(seq_len(n), dec,
                       function(i) mean(V[i][D[i] == 1]) - mean(V[i][D[i] == 0])))))
round(rbind(X1 = gap(X1), X2 = gap(X2)), 3)
```

\vspace{0.2em}

- Overall gaps of $0.77$ and $0.60$ collapse to at most $0.14$ and $0.15$ **within deciles of the score**.
- We never conditioned on $X_1$ or $X_2$ --- only on one number --- and both balanced.



## Comparing the estimators

```{r}
#| echo: true
set.seed(975)
n  <- 20000
X  <- rnorm(n)
ps <- plogis(1.2 * X)                  # true propensity score
D  <- rbinom(n, 1, ps)
Y  <- 2 * D + 3 * X + rnorm(n)         # TRUE effect = 2

eh <- predict(glm(D ~ X, family = binomial), type = "response")
round(c(naive = mean(Y[D == 1]) - mean(Y[D == 0]),
        ols   = unname(coef(lm(Y ~ D + X))[2]),
        ipw   = mean(D * Y / eh) - mean((1 - D) * Y / (1 - eh))), 3)
```

\vspace{0.2em}

- Naive comparison: $4.85$, more than double the truth. Both adjusted estimators land near $2$.
- They agree **because the model is correctly specified**. That is the assumption doing the work, not the method.

-->

# 2. Regression

## Crossing to the other side of the map

- Everything so far has been an **interpolation** method: subclassification, matching, the propensity score. Each one refuses to compare a treated unit that has no counterpart. Common support is the binding constraint.
- Regression sits on the other side. It does not need common support at all, and so **the curse of dimensionality simply does not arise**. 
- It buys that with a **functional form**. Where the data are silent, the fitted model speaks anyway: OLS imputes the missing counterfactual by **extrapolating** the outcome model into regions with no comparable units.
- That is the trade. Matching needs overlap; regression needs a correctly specified outcome model. 
- Both still need unconfoundedness. Nothing in this section relaxes it.

## Matching vs. regression

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
set.seed(975)
n0 <- 600; n1 <- 200; tau <- 2
f  <- function(x) 10 * (1 - exp(-x / 2.5))       # true E[Y0 | X], saturating
x0 <- runif(n0, 0, 5); x1 <- runif(n1, 3, 8)     # controls small, treated large
X  <- c(x0, x1); D <- c(rep(0, n0), rep(1, n1))
Y  <- f(X) + tau * D + rnorm(n0 + n1, 0, 0.4)    # constant effect of 2
ct <- which(D == 0); tr <- which(D == 1)
mate <- rep(NA_integer_, length(tr)); eff <- rep(NA_real_, length(tr))
for (k in seq_along(tr)) { i <- tr[k]; inn <- ct[abs(X[ct] - X[i]) <= 0.08]
  if (length(inn)) { mate[k] <- inn[which.min(abs(X[inn] - X[i]))]
                     eff[k]  <- Y[i] - mean(Y[inn]) } }
ok <- !is.na(mate); bb <- coef(lm(Y ~ D + X))    # Y = a + b D + c X + e

RED <- "#B2182B"; BLU <- "#2166AC"; yl <- c(-1, 14.6); xl <- c(0, 9.3)
par(mfrow = c(1, 2), mar = c(4, 4, 2.4, 1), cex = 0.8)

plot(NULL, xlim = xl, ylim = yl, xlab = "X", ylab = "Y",
     main = "Matching: interpolation")
rect(3, yl[1], 5, yl[2], col = "#0000000D", border = NA)
text(4, 13.9, "common support", cex = 0.7, col = "grey30")
points(X[ct], Y[ct], pch = 16, cex = 0.4, col = "#00000050")
points(X[tr][!ok], Y[tr][!ok], pch = 1, cex = 0.5, col = "#B2182B50")
points(X[tr][ok], Y[tr][ok], pch = 16, cex = 0.45, col = "#B2182B99")
sel <- which(ok)[seq(1, sum(ok), length.out = 11)]
segments(X[tr][sel], Y[tr][sel], X[mate[sel]], Y[mate[sel]], col = "grey15", lwd = 1.2)
arrows(7.4, 6.8, 6.5, 10.3, length = 0.06, col = "grey25")
text(7.5, 5.8, "no counterpart:\ndropped", cex = 0.7, col = "grey25")
legend("bottomright", c("control", "treated, matched", "treated, dropped"),
       pch = c(16, 16, 1), col = c("#00000088", RED, "#B2182B77"),
       bty = "n", cex = 0.7, pt.cex = 0.8)

plot(NULL, xlim = xl, ylim = yl, xlab = "firm size", ylab = "outcome",
     main = "Regression: extrapolation")
points(X[ct], Y[ct], pch = 16, cex = 0.4, col = "#00000024")
points(X[tr], Y[tr], pch = 16, cex = 0.4, col = "#B2182B2A")
abline(v = 5, lty = 3, col = "grey45")
text(4.88, 14.2, "controls end here", cex = 0.7, col = "grey30", adj = 1)
xs <- seq(0, 8.3, 0.05)
lines(xs, f(xs), col = BLU, lwd = 1.9); lines(xs, f(xs) + tau, col = BLU, lwd = 1.9)
lines(c(0, 5),   bb[1] + bb[3] * c(0, 5),   col = RED, lwd = 2.1)
lines(c(5, 8.3), bb[1] + bb[3] * c(5, 8.3), col = RED, lwd = 2.1, lty = 2)
lines(c(3, 8.3), bb[1] + bb[2] + bb[3] * c(3, 8.3), col = RED, lwd = 2.1)
xa <- 8.35
arrows(xa, f(xa), xa, f(xa) + tau, length = 0.035, code = 3, col = BLU)
text(xa + 0.12, f(xa) + tau/2, "2", cex = 0.72, col = BLU, adj = 0, font = 2)
yc <- bb[1] + bb[3] * xa
arrows(xa, yc, xa, yc + bb[2], length = 0.035, code = 3, col = RED)
text(xa + 0.12, yc + bb[2]/2, "0.94", cex = 0.72, col = RED, adj = 0, font = 2)
legend("bottomright", c("truth", "OLS",
                        "extrapolated: no controls"),
       lty = c(1, 1, 2), lwd = c(1.9, 2.1, 2.1), col = c(BLU, RED, RED),
       bty = "n", cex = 0.68)
```


<!--
## Gauss, Ceres, and the origin of least squares

- Giuseppe Piazzi discovered Ceres --- the first known asteroid --- on 1 January 1801, observed it for 42 days, and then lost it behind the Sun.
- Carl Friedrich Gauss, aged 24, modelled the observed positions as a function of orbital parameters with an error term for observational inaccuracy, and minimized the **sum of squared differences** between observed and predicted positions. By the end of 1801 astronomers recovered Ceres near his predicted position.
- So least squares was invented to solve a **prediction** problem, not a causal one. That origin still shapes what OLS is good at.
- Angrist and Pischke (2009) emphasize the same property in modern language: OLS minimizes mean squared prediction error among all linear functions, making it the **best linear approximation** to the conditional expectation function.
- Keep that distinction in view. In causal inference we are not predicting outcomes --- we are predicting **missing potential outcomes**, which is a different target.
-->

## The bivariate case is easy; the multivariate case looks hard

- With one regressor,
$$
Y_i = \beta_0 + \beta_1 X_i + u_i,
\qquad
\hat\beta_1 = \frac{\mathrm{Cov}(Y_i, X_i)}{\mathrm{Var}(X_i)} .
$$
- The coefficient is the slope of the best-fit line, a **covariance scaled by a variance**. 
- Now add covariates:
$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_k X_{ki} + \varepsilon_i .
$$
<!-- - How can $\hat\beta_1$ still be a scaled covariance? If the $X$'s are correlated with each other and with $Y$, then adding more of them *changes* $\hat\beta_1$ --- so it cannot be the same simple object. -->
- In a multivariate regression the coefficients are **not** simple scaled covariances. But the Frisch--Waugh--Lovell theorem (Frisch and Waugh 1933; Lovell 1963) says they are scaled covariances of something else.

## Frisch--Waugh--Lovell in three steps

**Step 1. Auxiliary regression**:
$$
X_{1i} = \gamma_0 + \gamma_2 X_{2i} + \cdots + \gamma_K X_{Ki} + f_i .
$$
$\hat X_{1i}$ is the part of $X_{1i}$ the other covariates explain.

**Step 2. Residualization**:
$$
\tilde X_{1i} = X_{1i} - \hat X_{1i} .
$$
The variation in $X_{1i}$ that is *uncorrelated* with everything else --- **partialling out**.

**Step 3. Regress the outcome on the residual.**
$$
Y_i = \alpha + \beta_1 \tilde X_{1i} + \varepsilon_i,
\qquad
\hat\beta_1 = \frac{\mathrm{Cov}(Y_i, \tilde X_{1i})}{\mathrm{Var}(\tilde X_{1i})} .
$$

- $\hat\beta_1$ is **identical** to the coefficient on $X_{1i}$ in the full regression. 

## FWL is an identity, not an approximation

Simulation: $4{,}000$ workers, earnings on a treatment indicator $d$, age, and high-school GPA, with a true effect of $\$1{,}500$.

\begin{center}
\footnotesize
\begin{tabular}{lcc}
\toprule
 & (1) Multivariate OLS & (2) FWL \\
\midrule
$d$   & $1503.201$ & $1503.201$ \\
      & $(8.249)$  & $(8.287)$ \\
age   & $1.173$    & --- \\
      & $(2.023)$  & \\
gpa   & $9.081$    & --- \\
      & $(4.615)$  & \\
$N$   & $4000$     & $4000$ \\
\bottomrule
\end{tabular}
\end{center}

- The estimate sits $0.39$ above the true $1500$ --- sampling noise.
- The standard errors do *not* agree, and that is not sampling noise. 

## Residualize both sides

- Another version residualizes the outcome too:
$$
\begin{align}
\tilde Y_i \;=\; Y_i \;-\; \hat Y_i &\quad\text{from}\quad Y_i = \lambda_0 + \lambda_2 X_{2i} + \cdots + \lambda_K X_{Ki} + v_i,\\
&\quad\text{then}\quad \tilde Y_i = \beta_1 \tilde X_{1i} + \varepsilon_i.
\end{align}
$$
- Both versions return **exactly the same $\hat\beta_1$**.
- But the double-residual version buys one more thing: **its residuals are the residuals of the full regression**. 
<!-- - That is not a technicality. The residuals are what the standard error is built from. -->

## Three regressions, one coefficient

\begin{center}
\footnotesize
\begin{tabular}{lcc}
\toprule
 & $\hat\beta_d$ & s.e. \\
\midrule
\multicolumn{3}{l}{\emph{original:} the controls explain $0.1\%$ of earnings} \\
full OLS                          & $1503.201$ & $8.249$ \\
single, $Y \sim \tilde D$          & $1503.201$ & $8.287$ \\
double, $\tilde Y \sim \tilde D$   & $1503.201$ & $8.247$ \\
\midrule
\multicolumn{3}{l}{\emph{expanded:} the controls explain $93\%$} \\
full OLS                          & $1503.201$ & $8.249$ \\
single, $Y \sim \tilde D$          & $1503.201$ & $\mathbf{91.493}$ \\
double, $\tilde Y \sim \tilde D$   & $1503.201$ & $8.247$ \\
\bottomrule
\end{tabular}
\end{center}

- Residualize only $D$: the residuals still carry everything age and GPA explain.

<!--

## Cunningham's `fwl.do`, extended

```stata
capture program drop fwl
program define fwl
    args bage bgpa

    clear
    set seed 1500
    set obs 4000
    gen age = rnormal(25,2)
    gen gpa = runiform(1,4)
    sum gpa
    replace gpa = gpa - r(mean)
    gen y0 = 15000 + `bage'*age + `bgpa'*gpa + rnormal(1000,250)
    gen y1 = y0 + 1500
    gen d = 0
    replace d = 1 in 1/1500
    gen earnings = d*y1 + (1-d)*y0     // true ATE = 1500

    reg earnings d age gpa             // the full regression
    reg d age gpa                      // step 1: auxiliary regression
    predict double dhat
    gen double dtilde = d - dhat       // step 2: residualize the regressor
    reg earnings dtilde                // step 3, as written
    reg earnings age gpa               // residualize the outcome too
    predict double yhat
    gen double ytilde = earnings - yhat
    reg ytilde dtilde                  // the double-residual version
end

fwl 0.5 1.5      // top panel:    controls explain 0.1% of earnings
fwl 500 3000     // bottom panel: controls explain 93%
```
-->

<!--
## Where the degrees-of-freedom factor comes from

- The homoskedastic OLS standard error is
$$
\mathrm{se}(\hat\beta_1) \;=\; \sqrt{\frac{s^2}{\sum_i \tilde X_{1i}^{2}}},
\qquad s^2 = \frac{SSR}{\mathrm{df}} .
$$
- The full regression and the double-residual regression share **both** pieces of that fraction's numerator: the same residualized regressor, so the same $\sum_i \tilde X_{1i}^2$; and, by the theorem, the same residuals, so the same $SSR$.
- Only the **divisor** differs. The full model spends $p$ parameters, so $\mathrm{df} = n - p$. The residualized regression sees an intercept and one slope, so $\mathrm{df} = n - 2$ --- it does not know the controls were already fitted.
$$
\frac{\mathrm{se}_{\text{full}}}{\mathrm{se}_{\text{double}}}
\;=\; \sqrt{\frac{SSR/(n-p)}{SSR/(n-2)}} \;=\; \sqrt{\frac{n-2}{n-p}}
$$
- Here $n = 4000$ and $p = 4$: $\sqrt{3998/3996} = 1.0002502$, which is the ratio $8.2495/8.2474$ in **both** panels above, to eight decimals.
- So the double-residual standard error is always **too small**, and correctably so. The single-residual one has a different $SSR$ --- no rescaling fixes it.
-->

## Why FWL is worth knowing

- **It tells you what variation identifies your coefficient.** $\hat\beta_1$ uses only the part of $X_1$ that the controls cannot explain. <!-- Add a control that absorbs most of the variation in $X_1$, and you are estimating from whatever scraps remain.-->
- **It makes the estimate visualizable.** Residualize *both* sides against the controls and scatter $\tilde Y$ against $\tilde X_1$: a multivariate coefficient becomes a two-dimensional picture whose slope is $\hat\beta_1$.\footnote{Filoso (2013) gives the worked proof and a Stata implementation.} 
- **It is the engine behind a great deal else.** Fixed effects, double machine learning, and the weighting results later in this section are all FWL in disguise.
- Angrist and Pischke (2009) call this **regression anatomy**, and it is the single most useful algebraic fact about OLS.

## Exogeneity, and the canonical OLS model

- Write the **canonical OLS model**:
$$
Y_i = \alpha + \delta D_i + \gamma X_i + \varepsilon_i .
$$
- For $\hat\delta$ to be causal, $D_i$ must be *exogenous* $\Rightarrow$ Imbens and Rubin (2015) put it plainly: exogeneity "combines unconfoundedness with functional form and constant treatment effect assumptions that are quite strong, and arguably unnecessary."
- Three assumptions, not one. Unconfoundedness you have already met and argued for. The other two arrived silently, in the act of writing down the equation.


## Taking conditional expectations

- Take expectations of the canonical model separately for the treated and the control group:
$$
\begin{aligned}
E[Y_i \mid D_i = 1, X_i] &= \alpha + \delta\, E[D_i \mid D_i = 1, X_i] + \gamma_{1} E[X_i \mid D_i = 1, X_i] \\
E[Y_i \mid D_i = 0, X_i] &= \alpha + \delta\, E[D_i \mid D_i = 0, X_i] + \gamma_{1} E[X_i \mid D_i = 0, X_i]
\end{aligned}
$$
- Now replace $Y$ with $Y^1$ or $Y^0$ as appropriate --- and do the same for $X$. 
- Why would $X$ have potential outcomes? **If $X^1 \neq X^0$, then $D$ causes $X$, and $X$ is an outcome, not a covariate.** That is the bad-control problem.
- And allow the covariate to matter differently in the two worlds, $\gamma_{11}$ against $\gamma_{01}$:
$$
\begin{aligned}
E[Y^1_i \mid D_i = 1, X_i] &= \alpha + \delta\, E[D_i \mid D_i = 1, X^1_i] + \gamma_{11} E[X^1_i \mid D_i = 1, X^1_i] \\
E[Y^0_i \mid D_i = 0, X_i] &= \alpha + \delta\, E[D_i \mid D_i = 0, X^0_i] + \gamma_{01} E[X^0_i \mid D_i = 0, X^0_i]
\end{aligned}
$$

## What $\gamma_{11} \neq \gamma_{01}$ actually means

- Let $Y^1$ be earnings with a college degree, $Y^0$ earnings without one, and $X$ be age.
- Then $\gamma_{11}$ is the age--earnings profile for **college graduates** and $\gamma_{01}$ is the age--earnings profile for the same people **had they not gone**.
- $\gamma_{11} \neq \gamma_{01}$ **is** heterogeneous treatment effects --- the return to treatment varying with $X$.
- In finance the same statement is unremarkable: the effect of an index inclusion, a rating change, or a governance shock varies with firm size. If it does, $\gamma_{11} \neq \gamma_{01}$.

## Heterogeneous treatment effect bias

- OLS on the canonical model is a difference in conditional means:
$$
\hat\delta_{OLS} = E[Y^1_i \mid D_i = 1, X^1_i] - E[Y^0_i \mid D_i = 0, X^0_i].
$$
- Substitute the two expressions from the previous slide. 
$$
\hat\delta_{OLS} \;=\; \delta \;+\; \underbrace{\Big(\gamma_{11} E[X^1_i \mid D_i = 1, X^1_i] \;-\; \gamma_{01} E[X^0_i \mid D_i = 0, X^0_i]\Big)}_{\text{heterogeneous treatment effect bias}}
$$
- This is what the canonical OLS model identifies. Not $\delta$. $\delta$ plus a term that has nothing to do with omitted variables.
- Notice what is *not* in the bias term: any unobserved confounder. We assumed unconfoundedness holds perfectly and there is nothing left to omit.

## The two assumptions you signed up for

The bias term vanishes if and only if **both** of the following hold.

1. **Covariates are not outcomes.** $X^1 = X^0 = X$: the covariates do not respond to treatment.
2. **Treatment effects are homogeneous in $X$.** $\gamma_{11} = \gamma_{01} = \gamma_1$: the return to treatment does not vary with the covariates.

- Under A1 and A2, $\hat\delta_{OLS} = \delta$. Fail either one and the canonical model is biased.
- **A1 has no repair.** Do not control for a collider or a mediator $\Rightarrow$ Lecture 2!
- **A2 does have a repair**, and it is a change to the specification, not to the data. That is the next slide.
- Note how strong A2 is. Neither matching nor the propensity score ever required it: they average heterogeneous effects, they do not assume them away.

## Regression adjustment: Oaxaca's answer

- The problem was never OLS. The problem was that **one specification** --- additive and separable.
<!-- - The fix comes from Ronald Oaxaca, who finished his PhD at Princeton in 1971 under Orley Ashenfelter. His dissertation introduced a decomposition of wage differentials, published as Oaxaca (1973), developed independently and contemporaneously with Blinder (1973), and generalizing a technique of the sociologist Evelyn Kitagawa (1955). -->
- The **Kitagawa--Oaxaca--Blinder decomposition** splits a mean gap into a part attributable to differences in *endowments* (education, experience) and a part attributable to differences in the *returns* to those endowments.
<!-- - Becker (1957) had given economics a theory of labour-market discrimination; Oaxaca gave it an empirical framework. -->
- Słoczyński (2015, 2020) showed the decomposition is also a **treatment effects estimator** --- and that it identifies both the $ATE$ and the $ATT$. This is what "regression adjustment" means.

## Deriving the specification


- Write both potential outcomes, letting $X$ act differently on each:
$$
\begin{aligned}
Y^0_i &= \alpha + \gamma_{01} X_i + \varepsilon_i \\
Y^1_i &= \alpha + \gamma_{01} X_i + \delta D_i + \gamma_{11} X_i \cdot D_i + \varepsilon_i
\end{aligned}
$$
- Take conditional expectations, then form $\delta = E[Y^1_i \mid D_i, X_i] - E[Y^0_i \mid D_i, X_i]$:
\small
$$
\begin{aligned}
&= \big(\alpha + \gamma_{01} E[X_i \mid D_i{=}0, X_i] + \delta + \gamma_{11} E[X_i \cdot D_i \mid D_i{=}1, X_i]\big) - \big(\alpha + \gamma_{01} E[X_i \mid D_i{=}0, X_i]\big) \\[0.2em]
&= \delta + \gamma_{11} E[X_i \cdot D_i \mid D_i = 1]
\end{aligned}
$$
\normalsize
- The $\gamma_{01}$ terms cancel because the *same* coefficient appears in both potential outcomes. The interaction is what remains --- and it is exactly the term the canonical model omitted.


## The regression adjustment estimator

To identify the $ATE$ and the $ATT$ under unconfoundedness **with heterogeneous effects**, run
$$
Y_i = \alpha + \delta D_i + \gamma_{01} X_i + \gamma_{11}\, D_i \cdot X_i + \varepsilon_i .
$$

- Interact the treatment with **every** covariate. Then
$$
\widehat{ATE} = \hat\delta, \qquad \widehat{ATT} = \hat\delta + \hat\gamma_{11}\, E[X_i \mid D_i = 1],
$$
where $E[X_i \mid D_i = 1]$ is the covariate mean **in the treated group only**.\footnote{The $ATE$ result requires $X$ to be demeaned in the full sample. If you interact with terms that are not mean-zero --- squares, cross-products --- add them back: $\widehat{ATE} = \hat\delta + \sum_k \hat\gamma_{1k}\bar X_k$.}

- The difference between the two estimands is now visible and mechanical: the same coefficients, evaluated at a different covariate distribution.
<!-- - See Imbens and Wooldridge (2009), Wooldridge (2010), and Słoczyński (2020). -->

## Does it work?

Simulation: $5{,}000$ workers, $5{,}000$ with a college degree; two exogenous covariates, age and GPA, with **different means in the treated and control groups**; both recentred, plus their squares and their interaction. Then
$$
\begin{aligned}
Y^0 &= 15000 + 10.25\,\text{age} - 10.5\,\text{age}^2 + 1000\,\text{gpa} - 10.5\,\text{gpa}^2 + 500\,(\text{age}\cdot\text{gpa}) + e \\
Y^1 &= Y^0 + 2500 + 100\,\text{age} + 1100\,\text{gpa}
\end{aligned}
$$

- Because age and GPA are recentred, $E[Y^1 - Y^0] = \$2{,}500$ exactly. The $ATT$ is **not** $\$2{,}500$: the treated group has a different age and GPA composition, and the effect varies with both.
- The canonical model here controls for age, age$^2$, GPA, GPA$^2$ and their interaction. Every covariate needed for unconfoundedness is in it. Nothing is omitted.
<!-- - Cunningham reports the canonical model centred at $\$2{,}380$ against a true $ATE$ of $\$2{,}500$, and regression adjustment centred exactly on both the $ATE$ and the $ATT$ of $\$1{,}952$. -->

## When is the canonical model actually wrong?

Simulation, varying the **share treated**:

\begin{center}
\small
\begin{tabular}{lccccc}
\toprule
Share treated & $0.20$ & $0.35$ & $0.50$ & $0.65$ & $0.80$ \\
\midrule
True $ATE$          & $2500$ & $2500$ & $2500$ & $2500$ & $2500$ \\
True $ATT$          & $1813$ & $1942$ & $2070$ & $2199$ & $2328$ \\
Canonical OLS       & $2097$ & $2310$ & $2500$ & $2689$ & $2902$ \\
Regression adj.\ ($ATE$) & $2500$ & $2500$ & $2499$ & $2499$ & $2499$ \\
Regression adj.\ ($ATT$) & $1812$ & $1941$ & $2069$ & $2198$ & $2327$ \\
\bottomrule
\end{tabular}
\end{center}

- Regression adjustment recovers **both** estimands at every split. 
- At exactly $50/50$ it lands on the $ATE$ --- **by coincidence of the design, not by validity**. 
<!-- - The pattern is Słoczyński (2022): the canonical estimand is a weighted average of $ATT$ and $ATU$ in which the **smaller group gets the larger weight**. In finance, treated shares of $5\%$ are routine. -->

## What regression adjustment does not buy

- Regression adjustment removes the **constant-effects** assumption. It does not remove the **functional form** assumption, and cannot.
- Inside the model is a model of the missing potential outcome. In the simulation the true $Y^0$ contained age, age$^2$, GPA, GPA$^2$ and age$\,\cdot\,$GPA --- so the treatment had to be interacted with **all five**.
- Drop just one interaction and it breaks. With the treatment effect also varying in age$\,\cdot\,$GPA, the fully interacted model returns $2576$ against a true $ATE$ of $2575$; omitting the single $D \cdot (\text{age}\cdot\text{gpa})$ term returns $2368$.
- **The functional-form requirement replaces the common-support requirement. It is not free.** You have swapped a checkable condition for an uncheckable one.
- This is the honest summary of the trade made at the start of the section.

## Three things wrong with the canonical model

1. **It assumes constant treatment effects.** Neither matching nor the propensity score ever asked for this. Regression adjustment removes it, at the cost of interacting the treatment with everything.
2. **Its weights are wrong.** Słoczyński (2022) shows the coefficient is $w_1 \cdot ATT + w_0 \cdot ATU$, with each group's weight **inversely** proportional to its size; **it is generally none of the three**. 
3. **It can only ever deliver one number.** There is one treatment indicator, so there is one coefficient. Regression adjustment delivers the $ATE$ and the $ATT$ from the same fit.


# 3. What Matching Cannot Do

## Matching creates no new variation

- Matching **does not introduce exogenous variation**. It reorganizes and reweights the data you already have, using observables you already measured.
- Consequence: if your OLS regression was biased by endogeneity, **the matched estimate is biased too**.
- Contrast this with the designs still to come:
  - **IV** imports variation from outside the model.
  - **RDD** exploits assignment that is locally as good as random.
  - **DiD** differences out fixed unobservables.
- Matching belongs to a different category. It is a *conditioning strategy*, like regression --- not a research design.

## What it cannot fix

Matching is powerless against:

- **Unobserved confounders.** You cannot match on what you cannot measure.
- **Reverse causality.** Matching on $X$ says nothing about whether $Y$ caused $D$.
- **Measurement error** in treatment or outcome.
- **Bad controls.** Matching on a post-treatment variable is exactly as damaging as controlling for it --- Lecture 2 applies unchanged, and the collider $F$ of Lecture 3a, §2, Confounders, Covariates, and Colliders, is the warning.

\vspace{0.4em}

\begin{center}
\emph{You cannot match on what you cannot measure.}
\end{center}

## So what is it good for?

Given all that, matching still earns its place:

1. **A non-parametric benchmark.** It estimates the effect without assuming a functional form. If matching and OLS disagree sharply, your specification is doing more work than you thought.
2. **Common-support discipline.** Matching refuses to compare units with no counterpart. Regression silently extrapolates into regions with no data.
3. **Diagnostic value.** The propensity score distribution reveals overlap problems that a regression table hides completely.
4. **Robustness when the functional form is wrong.** Matching can outperform a misspecified regression in finite samples.

\vspace{0.3em}

- Use it as a *complement* to regression and a check on it --- not as an identification strategy.


<!--
# 4. Sensitivity Analysis

## The assumption you cannot test, revisited

- Two things this course has said are both true and both final: unconfoundedness is **untestable**, and no amount of matching creates variation that fixes an unobserved confounder.
- Neither is the end of the conversation. **Sensitivity analysis** changes the question:

\begin{center}
\emph{Not} ``is my design valid?'' --- unanswerable. \quad But: \\[0.25em]
\textbf{how large would a hidden confounder have to be to overturn my conclusion?}
\end{center}

- The output is not a $p$-value. It is a **threshold**, plus an argument about whether a confounder that large is plausible in your setting. The argument is yours; the threshold is arithmetic.
- Two families, matching the two routes of §1, Propensity Scores, and §2, Regression:
  - **Design-based**, for matched pairs: Rosenbaum bounds.
  - **Regression-based**, for the coefficient: Oster's $\delta$, and the Cinelli--Hazlett robustness value.
- Referees in finance now ask for one of these by name. Budget a table for it.

## Rosenbaum bounds: one parameter, $\Gamma$

- Matching made two units identical in $X$. Unconfoundedness then says they had **equal** probability of treatment. Suppose they did not.
- Let $\Gamma \geq 1$ bound how unequal. For any two units $i, j$ with $X_i = X_j$,
$$
\frac{1}{\Gamma} \;\leq\; \frac{\pi_i (1 - \pi_j)}{\pi_j (1 - \pi_i)} \;\leq\; \Gamma .
$$
- $\Gamma = 1$ **is** unconfoundedness. $\Gamma = 2$ says a hidden variable could leave one unit of the pair twice as likely, in odds, to be treated.
- For each $\Gamma$, compute the **worst case** over every hidden assignment consistent with that $\Gamma$, and report the resulting bound on the $p$-value of the randomization test.
- Then read off $\Gamma^{*}$: the point where the upper bound crosses your threshold.
- Note what $\Gamma^{*}$ is *not*. It is a statement about how much hidden bias your **design** can absorb. It never says the design is fine.

## Reading a Rosenbaum bound

$200$ matched pairs; one-sided Wilcoxon signed-rank test on the within-pair differences.

\begin{center}
\small
\begin{tabular}{lcccccccc}
\toprule
$\Gamma$ & $1.00$ & $1.10$ & $1.20$ & $1.30$ & $\mathbf{1.40}$ & $1.50$ & $1.75$ & $2.00$ \\
\midrule
upper bound on $p$ & $0.0001$ & $0.0008$ & $0.0040$ & $0.0149$ & $\mathbf{0.0415}$ & $0.0922$ & $0.3354$ & $0.6386$ \\
\bottomrule
\end{tabular}
\end{center}

- At $\Gamma = 1$ the result looks bulletproof: $p = 0.0001$. It is not. $\Gamma^{*} = 1.42$.
- A hidden variable that shifts the odds of treatment by $42\%$ between two otherwise identical units is enough to make this finding insignificant.
- This is the standard and slightly deflating experience. Strong $p$-values routinely coexist with small $\Gamma^{*}$, because the two answer different questions.
- Which leaves the question the table cannot answer: **is $1.42$ big?**

## Calibrating $\Gamma$: what does $1.42$ mean?

- $\Gamma$ is one number doing the work of two. Rosenbaum and Silber (2009) split it back apart: let $\Lambda$ be the odds ratio linking the hidden variable to **treatment**, and $\Delta$ the odds ratio linking it to the **outcome**. Then
$$
\Gamma \;=\; \frac{\Lambda\Delta + 1}{\Lambda + \Delta}.
$$
- Every $(\Lambda, \Delta)$ on that curve produces the same $\Gamma$. At $\Gamma^{*} = 1.42$:

\begin{center}
\small
\begin{tabular}{lccc}
\toprule
$\Lambda$ (odds of treatment) & $2.43$ & $3.00$ & $5.00$ \\
$\Delta$ (odds of a higher outcome) & $2.43$ & $2.07$ & $1.71$ \\
\bottomrule
\end{tabular}
\end{center}

- Now the question is one a *finance* audience can argue about: is there a plausible omitted characteristic that roughly **doubles** the odds of treatment and **doubles** the odds of a better outcome? Often the honest answer is yes.
- **Benchmark it.** Compute $\Lambda$ and $\Delta$ for your strongest *measured* confounder. If the killer confounder must be twice as strong as anything you measured, that is an argument. If half as strong will do, so is that.
- The bound is worst-case, so $\Gamma^{*}$ understates the true robustness.

## Oster's $\delta$: coefficient stability, done properly

- The informal version is in every seminar: *``the coefficient barely moved when I added controls, so unobservables cannot matter much.''*
- That intuition is not wrong, it is **unfinished**. It ignores how much the controls explained. A coefficient that is stable while $R^2$ barely moves has demonstrated nothing.
- Altonji, Elder and Taber (2005) turned it into an argument; Oster (2019) turned it into an estimator. Two inputs:
  - $\delta$ --- selection on unobservables **relative to** selection on observables. $\delta = 1$ means they are equally severe.
  - $R_{\max}$ --- the $R^2$ of the infeasible regression that included the unobservables.
- With $(\mathring\beta, \mathring R)$ from the short regression and $(\tilde\beta, \tilde R)$ from the controlled one,
$$
\beta^{*} \;\approx\; \tilde\beta \;-\; \delta\,\big(\mathring\beta - \tilde\beta\big)\,\frac{R_{\max} - \tilde R}{\tilde R - \mathring R}.
$$
- Two ways to report it: fix $\delta = 1$ and give $\beta^{*}$; or set $\beta^{*} = 0$ and solve for the $\delta$ that would be needed. Oster's calibration from randomized trials: $\delta = 1$, $R_{\max} = \min(1.3\,\tilde R,\, 1)$.

## Oster in practice: a warning

Simulated data, one unobserved confounder, **true effect $= 1.000$**:

\begin{center}
\small
\begin{tabular}{lcc}
\toprule
 & $\beta$ & $R^2$ \\
\midrule
short: $Y \sim D$ & $2.086$ & $0.452$ \\
controlled: $Y \sim D + X$ & $1.743$ & $0.484$ \\
infeasible: $Y \sim D + X + U$ & $1.014$ & $0.562$ \\
\bottomrule
\end{tabular}
\end{center}

- Oster's default rule sets $R_{\max} = 1.3 \times 0.484 = 0.629$, and returns $\beta^{*} = \mathbf{0.201}$ --- a correction that overshoots the truth by a mile.
- Feed it the *true* $R_{\max} = 0.562$ and it returns $\beta^{*} = \mathbf{0.918}$, almost exactly right. The method works. The **guess** was what failed.
- Meanwhile the $\delta$ test passes: the $\delta$ that would drive $\beta^{*}$ to zero is $1.13 > 1$.
- **The lesson.** Report $\delta$, which is a bounded comparison you can defend. Do not report $\beta^{*}$ as if it were a point estimate --- it is a linear function of a number you guessed. If you show it, show it across a range of $R_{\max}$.

## Cinelli and Hazlett (2020): the robustness value

- Same problem, third parameterization --- and the one that travels best, because it uses **partial $R^2$** rather than odds ratios or selection ratios.
- $RV_q$ is the minimum share of the residual variation that a confounder must explain in **both** $D$ and $Y$ in order to move the estimate by $100q\%$.
- In the regression above: $t = 47.1$ and $RV_{q=1} = \mathbf{0.48}$. Any confounder explaining less than $48\%$ of what is left in both $D$ and $Y$ cannot drive the effect to zero.
- Is that reassuring? Check it against the confounder we planted: $U$ explains $15\%$ of the residual variation in $Y$ and $34\%$ in $D$. Below the bar --- and indeed it does not kill the effect, it moves it from $1.743$ to $1.014$.
- The feature that sells it: **benchmark bounds.** Express $RV$ as a multiple of an observed covariate's partial $R^2$ --- ``the confounder would have to be $3\times$ as important as firm size'' is a sentence a referee can evaluate.
- Software: `sensemakr` (R and Stata), `psacalc` for Oster, `rbounds` and `mhbounds` for Rosenbaum.
-->


# 4. The NSW Demonstration

## The 1970s empirical crisis

- The methods in this lecture were not adopted because they were elegant. They were adopted after a crisis of confidence in observational work.
- Job-training programs in the 1970s were evaluated observationally, and the estimates were all over the place --- depending, apparently, on the analyst.
- The **National Supported Work (NSW) Demonstration** was different: it *randomized* participation. So the experimental effect was known.
- That created something rare and valuable: a **benchmark**. You could now ask whether observational methods, applied to the same treated units, recover the experimental answer.
- Everything in this lecture gets graded against it.

## What LaLonde did

- LaLonde (1986) ran the decisive test. He took the NSW experiment, **threw away the experimental control group**, and replaced it with comparison groups built from survey data (CPS and PSID).
- He then applied the standard observational toolkit of the day and compared the results to the experimental benchmark of roughly **\$1,800** in annual earnings gains.
- The observational estimates ranged from a fraction of the benchmark to **the wrong sign** --- and which one you got depended heavily on how the comparison group was built.
- The paper was devastating and hugely influential. It is a major reason the profession moved toward IV, RDD, and DiD --- the other river.

## Dehejia and Wahba: not so fast

- **Dehejia and Wahba (1999, 2002)** revisited LaLonde's data with propensity score methods.
- Their finding: the observational estimates *did* recover the experimental benchmark reasonably well --- **once common support was imposed**.
- The key diagnosis: much of LaLonde's failure was an **overlap** problem, not only a confounding problem. The CPS and PSID comparison groups contained enormous numbers of people utterly unlike the NSW participants.
- Discard the non-comparable units and the methods perform far better.
- This is why the distinction of Lecture 3a, §3, Identification Assumptions, matters so much. Two different assumptions were failing, and only one of them is checkable.

## Smith and Todd: not so fast either

- **Smith and Todd (2005)** re-examined the Dehejia--Wahba result and found it fragile.
- The success was **sensitive to the subsample** analysed and to the **propensity score specification** chosen. Reasonable alternative choices undid it.
- Their charge is uncomfortable and worth stating plainly: the specification that worked may have been found *because* the experimental answer was already known.
- You cannot run that search in real research. In a genuine application there is no benchmark to check against --- that is the entire point of doing the study.

## The honest summary

- Selection on observables **can** work --- when overlap is genuine and the confounders are truly measured.
- Neither condition is verifiable in the setting where you need them:
  - Unconfoundedness is **untestable**, always.
  - Common support is **checkable**.
- So the practical lesson is asymmetric: do the check you can do, and argue explicitly for the assumption you cannot.
- **A useful exercise**: the NSW data are freely available. Apply every method in this lecture and compare them to the \$1,800 benchmark. Cunningham's chapter walks through it, and it is the fastest way to develop instincts about when these methods hold up.

# Wrapping Up

## Summary [Part 1]: the score and what it does

- **The propensity score** reduces $X$ to one number and defeats the curse of dimensionality for *conditioning*. It does nothing for **identification** --- unconfoundedness is still assumed, never delivered.
- **Judge the score by balance achieved, not by fit.** A high pseudo-$R^2$ with bad balance is a bad score; the reverse is fine. The whole stage is blind to $Y$, which is what makes the specification search legitimate.
- **Balance means the distribution, not the mean.** Standardized differences *and* variance ratios. Our `Volatility` covariate had $\text{SMD} = -0.010$ and a variance ratio of $2.27$.
- **IPW** reweights the sample into a pseudo-experiment: $1/\hat e$ for treated and $1/(1-\hat e)$ for controls builds the population; $\hat e/(1-\hat e)$ on controls alone rebuilds the *treated* distribution. Watch the tails, and normalize.
- **Propensity score matching** is the one use with a serious case against it: King and Nielsen (2019) show it targets complete randomization when you could have had blocking, so pruning can *increase* imbalance.

## Summary [Part 2]: the verdict

- **Regression is a conditioning strategy too**, not a rival family. FWL shows it uses only the **residual variation in $D$** --- and Słoczyński's weights show that the coefficient this produces is generally neither the $ATE$ nor the $ATT$.
- **Matching creates no new exogenous variation.** If OLS was endogenous, matching is endogenous. It cannot fix an unobserved confounder, reverse causality, or a bad control.
- **NSW**: these methods can work when overlap is real and the confounders are measured --- and you can only ever verify one of those two.
- Next lecture: designs that *create* variation instead of assuming it away.

## A checklist for your own paper

1. State the **estimand** before the estimator. $ATE$, $ATT$, or $ATU$ --- and say why the policy question needs that one.
2. Defend the **assignment mechanism** in institutional terms. Name the confounders and say how you measured them.
3. Justify every covariate as **pre-treatment**. Any post-treatment control needs an argument, not a footnote.
4. **Plot the overlap.** Report what you trimmed and how the estimand changed.
5. Report **balance**, not fit: standardized differences and variance ratios, before and after.
6. Report **matching and regression**. If they disagree, diagnose whether it is weights or extrapolation.
7. Run a **sensitivity analysis**, and calibrate the threshold against an observed covariate.
8. Run a **placebo** on a pre-treatment period.
9. Say plainly what would break the design --- and what unobservable would have to exist.

## References

\tiny

- Altonji, J., T. Elder and C. Taber (2005). Selection on Observed and Unobserved Variables: Assessing the Effectiveness of Catholic Schools. *JPE* 113, 151--184.
- Ben-Michael, E., A. Feller and J. Rothstein (2021). The Augmented Synthetic Control Method. *JASA* 116, 1789--1803.
- Blinder, A. (1973). Wage Discrimination: Reduced Form and Structural Estimates. *Journal of Human Resources* 8, 436--455.
- Cinelli, C. and C. Hazlett (2020). Making Sense of Sensitivity: Extending Omitted Variable Bias. *JRSS Series B* 82, 39--67.
- Cunningham, S. *Causal Inference: The Remix*, Ch. 5 (§5.9 for regression).
- Dehejia, R. and S. Wahba (1999). Causal Effects in Nonexperimental Studies. *JASA* 94, 1053--1062.
- Ding, P. (2024). *A First Course in Causal Inference*. Chapman and Hall/CRC.
- Filoso, V. (2013). Regression Anatomy, Revealed. *Stata Journal* 13(1), 92--106.
- Frisch, R. and F. Waugh (1933). Partial Time Regressions as Compared with Individual Trends. *Econometrica* 1, 387--401.
- Hájek, J. (1971). Comment on Basu. In *Foundations of Statistical Inference*, Holt, Rinehart and Winston, 236.
- Hirano, K. and G. Imbens (2001). Estimation of Causal Effects Using Propensity Score Weighting. *Health Services and Outcomes Research Methodology* 2, 259--278.
- Imbens, G. (2004). Nonparametric Estimation of Average Treatment Effects under Exogeneity. *Review of Economics and Statistics* 86, 4--29.
- Imbens, G. and D. Rubin (2015). *Causal Inference for Statistics, Social, and Biomedical Sciences*. Cambridge University Press.
- Imbens, G. and J. Wooldridge (2009). Recent Developments in the Econometrics of Program Evaluation. *JEL* 47(1), 5--86.
- King, G. and R. Nielsen (2019). Why Propensity Scores Should Not Be Used for Matching. *Political Analysis* 27, 435--454.
- Kitagawa, E. (1955). Components of a Difference Between Two Rates. *JASA* 50, 1168--1194.
- LaLonde, R. (1986). Evaluating the Econometric Evaluations of Training Programs with Experimental Data. *AER* 76, 604--620.
- Lovell, M. (1963). Seasonal Adjustment of Economic Time Series and Multiple Regression Analysis. *JASA* 58, 993--1010.
- Oaxaca, R. (1973). Male--Female Wage Differentials in Urban Labor Markets. *International Economic Review* 14, 693--709.
- Oster, E. (2019). Unobservable Selection and Coefficient Stability: Theory and Evidence. *JBES* 37, 187--204.
- Rosenbaum, P. (2002). *Observational Studies*, 2nd ed. Springer.
- Rosenbaum, P. and D. Rubin (1983). The Central Role of the Propensity Score in Observational Studies for Causal Effects. *Biometrika* 70, 41--55.
- Rosenbaum, P. and J. Silber (2009). Amplification of Sensitivity Analysis in Matched Observational Studies. *JASA* 104, 1398--1405.
- Słoczyński, T. (2022). Interpreting OLS Estimands When Treatment Effects Are Heterogeneous: Smaller Groups Get Larger Weights. *Review of Economics and Statistics* 104(3), 501--509.
- Smith, J. and P. Todd (2005). Does Matching Overcome LaLonde's Critique of Nonexperimental Estimators? *Journal of Econometrics* 125, 305--353.

