---
title: "Unconfoundedness II: Propensity Scores and Regression"
subtitle: "BUSS975: Causal Inference in Financial Research"
author: "Professor Ji-Woong Chung"
institute: "Korea University"
format:
  beamer:
    aspectratio: 169
    fontsize: 11pt
    slide-level: 2
    toc: false
    pdf-engine: pdflatex
    pdf-engine-opts:
      - "-interaction=nonstopmode"
    keep-tex: true
    include-in-header:
      text: |
        \setbeamertemplate{navigation symbols}{}
        \setbeamertemplate{footline}[frame number]
        \setbeamertemplate{itemize items}[circle]
        \setbeamertemplate{itemize subitem}[triangle]
        \setbeamercolor{frametitle}{fg=black}
        \setbeamerfont{frametitle}{series=\bfseries}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{booktabs}
        \usepackage{tikz}
        \usetikzlibrary{arrows.meta,positioning,calc}
        \tikzset{
          dnode/.style={circle,draw,thick,minimum size=9mm,inner sep=1pt,font=\small},
          unode/.style={circle,draw,thick,dashed,minimum size=9mm,inner sep=1pt,font=\small},
          bnode/.style={rectangle,draw,thick,fill=gray!25,minimum size=9mm,inner sep=3pt,font=\small},
          de/.style={-{Stealth[length=2.2mm]},thick},
          dde/.style={-{Stealth[length=2.2mm]},thick,dashed}
        }
        % shrink code input (fancyvrb) and code output (verbatim)
        \usepackage{fancyvrb}
        \fvset{fontsize=\scriptsize}
        \makeatletter
        \def\verbatim@font{\scriptsize\ttfamily}
        \makeatother
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

## Outline

1. Propensity scores
2. Regression
3. What matching cannot do
4. Sensitivity analysis
5. The NSW demonstration

\vspace{0.6em}

Lecture 3a matched units cell by cell and paid for it twice --- in bias and in variance. This lecture replaces the whole vector $X$ with a single number, and asks what that buys and what it costs.

# 1. Propensity Scores

## Defining the propensity score

- **Propensity score**: $e(X) = \Pr(D = 1 \mid X)$ 
  - "the conditional probability of assignment to a particular treatment given a vector of observed covariates"

- **Balancing property** (Rosenbaum--Rubin 1983): $X \perp D \mid \Pr(D = 1 \mid X)$
  - $E[X \mid D=1, \Pr(D=1 \mid X)] = E[X \mid D=0, \Pr(D=1 \mid X)]$
  - Covariate distributions are similar between treatment and control conditional the propensity score.

- **Propensity score theorem** (Rosenbaum--Rubin 1983): if $(Y^1,Y^0) \indep D \mid X$, then
$$
(Y^1, Y^0) \;\indep\; D \;\mid\; e(X)
$$
  - Conditioning on a **single number** does what conditioning on all of $X$ does. Ten covariates collapse to one scalar --- the curse of dimensionality defeated, for *conditioning*.




<!--
## Why the score balances

By definition, $\Pr(D{=}1 \mid X,\, e(X)) = e(X)$. Now compute the same probability knowing *only* the score:

$$
\Pr(D{=}1 \mid e(X)) \;=\; E\big[\,\underbrace{\Pr(D{=}1 \mid X)}_{=\;e(X)} \;\big|\; e(X)\,\big]
\;=\; E\big[e(X) \mid e(X)\big] \;=\; e(X)
$$

- The two are equal: knowing $X$ *on top of* $e(X)$ does not change the chance of treatment. That is $D \indep X \mid e(X)$, and **no assumption was used**.
- Add unconfoundedness --- $\Pr(D{=}1 \mid Y^0,Y^1,X) = e(X)$ --- and the identical argument conditioning on $(Y^0,Y^1,e(X))$ again returns $e(X)$. The potential outcomes carry no extra information either.
- **In one sentence**: $e(X)$ is a *sufficient statistic for the assignment mechanism*. Once you know how biased the coin is, knowing which $X$ made it that biased says nothing more about how it landed.
- What is **not** claimed: units with the same score are not similar. Only their *distributions* match.
-->

<!--
## Three cells, one number

\begin{center}
\small
\begin{tabular}{llrrrr}
\toprule
Cell & $X$ = (size, leverage) & $N$ & $e(X)$ & treated & control \\
\midrule
A & Large, Low  & $200$ & $0.2$ & $40$  & $160$ \\
B & Small, High & $400$ & $0.2$ & $80$  & $320$ \\
C & Large, High & $400$ & $0.6$ & $240$ & $160$ \\
\midrule
  & Total       & $1{,}000$ & & $360$ & $640$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- Read the score as a frequency: *what fraction of firms like this are treated?* In cell A, $40$ of $200$, so $e = 0.2$.
- Cells **A and B share a score** but are *opposites* on $X$. The stratum $e = 0.2$ pools $600$ firms: $120$ treated, $480$ control.
- Of those $120$ treated, $40$ come from A --- $33.3\%$. Of the $480$ controls, $160$ come from A --- $33.3\%$. **Identical**, and not an approximation: both cells drew from the same $0.2$ coin, so pooling preserves their $1{:}2$ ratio on *both* sides.

## The balance table

\begin{center}
\small
\begin{tabular}{lrrr}
\toprule
 & Treated & Control & Gap (pp) \\
\midrule
\emph{Raw sample} ($n = 1{,}000$) & & & \\
\quad \% Large             & $77.8$ & $50.0$ & $27.8$ \\
\quad \% High leverage     & $88.9$ & $75.0$ & $13.9$ \\
\midrule
\emph{Within} $e = 0.2$ (cells A and B) & & & \\
\quad \% Large             & $33.3$ & $33.3$ & $\mathbf{0.0}$ \\
\quad \% High leverage     & $66.7$ & $66.7$ & $\mathbf{0.0}$ \\
\midrule
\emph{Within} $e = 0.6$ (cell C) & & & \\
\quad \% Large             & $100.0$ & $100.0$ & $\mathbf{0.0}$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- The raw sample is badly imbalanced on **both** covariates. Conditioning on one number --- never on size, never on leverage --- balances **both**, exactly.
- This is the whole appeal of the score, and it is also the check you run in your own data: split $\hat e$ into deciles and put this table in the paper.
-->

## Checking balance: two numbers, not a $t$-test

The balancing property is testable. Report it covariate by covariate, before and after.

$$
\text{SMD} \;=\; \frac{\bar X_1 - \bar X_0}{\sqrt{\big(s_1^2 + s_0^2\big)/2}},
\qquad\qquad
\text{VR} \;=\; \frac{s_1^2}{s_0^2}
$$

- **Standardized mean difference.** Scale-free, so leverage and log assets are comparable. $|\text{SMD}| < 0.1$ is good; Rubin (2001) treats $0.25$ as the outer limit past which regression adjustment is untrustworthy.
- Keep the denominator **fixed at its raw-sample value** when recomputing after weighting, or the yardstick moves with the thing you are measuring.
- **Variance ratio.** Means can match while spreads do not. Rubin: stay inside $[0.5,\,2]$, ideally $[0.8,\,1.25]$.
- **Do not use $t$-tests.** Balance is a property of *this sample*, not a population hypothesis --- and $p$ confounds imbalance with sample size, so pruning to a matched subsample raises it mechanically even as balance worsens (Imai, King and Stuart 2008).

## A balance plot: means are not enough

```{r}
#| echo: false
#| fig-width: 9.6
#| fig-height: 3.25
set.seed(975)
n <- 5000
size <- rnorm(n); lev <- rnorm(n); roa <- rnorm(n)
age <- rnorm(n); vol <- rnorm(n); rnd <- rnorm(n)
X  <- cbind(Size=size, Leverage=lev, ROA=roa, Age=age, Volatility=vol, `R&D`=rnd)
D  <- rbinom(n, 1, plogis(0.9*size + 0.7*lev - 0.6*roa + 0.25*age + 0.7*vol^2 - 0.7))
eh <- predict(glm(D ~ size+lev+roa+age+vol, family = binomial), type = "response")
w  <- ifelse(D == 1, 1/eh, 1/(1-eh))
wm <- function(x,u) sum(u*x)/sum(u)
wv <- function(x,u){ m<-wm(x,u); sum(u)*sum(u*(x-m)^2)/(sum(u)^2-sum(u^2)) }
den <- apply(X,2,function(x) sqrt((var(x[D==1])+var(x[D==0]))/2))
smd <- function(u) sapply(1:ncol(X), function(k){ x<-X[,k]
         (wm(x[D==1],u[D==1])-wm(x[D==0],u[D==0]))/den[k] })
vr  <- function(u) sapply(1:ncol(X), function(k){ x<-X[,k]
         wv(x[D==1],u[D==1])/wv(x[D==0],u[D==0]) })
one <- rep(1,n); k <- ncol(X); y <- k:1; RED <- "#B2182B"
par(mfrow=c(1,2), mar=c(4,7,2.2,1), cex=0.8)
plot(NULL, xlim=c(0,0.78), ylim=c(0.5,k+0.5), yaxt="n", ylab="",
     xlab="| standardized difference |", main="Means")
axis(2, at=y, labels=colnames(X), las=1); abline(h=y, col="grey92")
abline(v=0.1, lty=2, col=RED)
points(abs(smd(one)), y, pch=1,  cex=1.3, col="grey35")
points(abs(smd(w)),   y, pch=19, cex=1.3, col=RED)
legend("right", c("raw","IPW weighted"), pch=c(1,19),
       col=c("grey35",RED), bty="n", cex=0.95)
plot(NULL, xlim=c(0.6,3.0), ylim=c(0.5,k+0.5), yaxt="n", ylab="",
     xlab="variance ratio  (treated / control)", main="Spreads")
axis(2, at=y, labels=colnames(X), las=1); abline(h=y, col="grey92")
abline(v=1, col="grey40"); abline(v=c(0.8,1.25), lty=2, col=RED)
points(vr(one), y, pch=1,  cex=1.3, col="grey35")
points(vr(w),   y, pch=19, cex=1.3, col=RED)
```

\vspace{-0.3em}
\footnotesize
Selection here is linear in `Size`, `Leverage`, `ROA`, `Age` --- and **quadratic in `Volatility`**. The score was fitted with linear terms only.

## Reading the balance plot

- **Left panel, the success.** Four covariates start badly imbalanced --- `Size` at $0.65$, `Leverage` and `ROA` near $0.49$. After weighting, every one is under $0.02$. This is the plot referees want, and it is the one most papers stop at.
- **Left panel, the trap.** `Volatility` looks *perfect* from the start: $\text{SMD} = -0.010$, and a $t$-test returns $p = 0.72$. On the evidence of means, nothing to see.
- **Right panel.** Its variance ratio is $\mathbf{2.27}$ --- treated firms have more than twice the spread. Extreme volatility drives selection in *both* tails, so the means cancel and the distributions do not.
- **Weighting makes it worse**, $2.27 \to \mathbf{2.83}$, because the fitted score never contained $\texttt{vol}^2$. Add that one term and the ratio falls to $\mathbf{1.03}$.
- The lesson of this section: judge the score by **balance achieved, not by fit** --- and balance means the whole distribution. Check squares and interactions, not just levels.

## Estimating the score

- You know the **true** score only when you ran the experiment yourself. Otherwise, it must be estimated: $\Pr(D_i {=} 1 \mid X_i) = F(\beta X_i)$, fitted by **logit or probit**.
- Judge the specification by **balance achieved**, not by fit. A score with a high pseudo-$R^2$ but poor balance is a bad score; the reverse is fine.
- Fit, check balance, add interactions and polynomials where balance fails, refit. A legitimate specification search, because it never looks at $Y$ --- the whole stage is **blind to the outcome**, which protects against fishing. Say so in the paper.
- **The score creates no support that $X$ lacks.** If a covariate cell is all-treated or all-control, collapsing $X$ to a scalar hides that; it does not fix it.

## Three things to do with the score

Once every unit has a $\hat e(X_i)$, there are three standard uses. 

1. **Diagnose.** Plot $\hat e$ separately for treated and control units. This is the common-support check, and it costs nothing.
2. **Weight.** Reweight the sample to build a pseudo-experiment --- inverse probability weighting.
3. **Match.** Pair treated to control units with similar scores, then difference --- propensity score matching.

<!--
\vspace{0.3em}

- A fourth, older use: **stratify** on the score. That is Lecture 3a, §4, Subclassification, with strata defined by deciles of $\hat e$ --- Rosenbaum and Rubin's original proposal, and still a reasonable one.
- Uses 1 and 4 are the safest. Use 2 is standard but fragile in the tails. Use 3 has a serious critique against it.
- The section closes with a fifth use that combines 2 with a regression: **doubly robust** estimation.
-->

## Visualizing overlap

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.0
set.seed(975)
n <- 4000
par(mfrow = c(1, 2), mar = c(4, 4, 2.2, 1), cex = 0.8)
for (a in c(0.8, 3.5)) {
  X  <- rnorm(n); D <- rbinom(n, 1, plogis(a * X))
  eh <- predict(glm(D ~ X, family = binomial), type = "response")
  br <- seq(0, 1, 0.05)
  h1 <- hist(eh[D == 1], breaks = br, plot = FALSE)
  h0 <- hist(eh[D == 0], breaks = br, plot = FALSE)
  plot(NULL, xlim = c(0, 1), ylim = c(0, max(h1$density, h0$density) * 1.28),
       xlab = "estimated propensity score", ylab = "density",
       main = if (a < 1) "Good overlap" else "Poor overlap")
  rect(br[-length(br)], 0, br[-1], h0$density, col = "#00000022", border = "grey40")
  rect(br[-length(br)], 0, br[-1], h1$density, col = "#B2182B55", border = "#B2182B")
  legend("top", c("control", "treated"), fill = c("#00000022", "#B2182B55"),
         bty = "n", horiz = TRUE, cex = 0.9)
}
```

\vspace{-0.4em}

\footnotesize
Right panel: a treated mass at $\hat e \approx 1$ with no controls, and a control mass at $\hat e \approx 0$ with no treated. Those units have **no counterfactual**. Any estimate covering them is extrapolation, not comparison.

## This plot belongs in every such paper

- Common support is verifiable, unlike unconfoundedness.
- What to do when it looks like the right panel:
  - **Trim** observations with extreme scores, or impose a caliper. Drop units with $\hat e \notin [0.1,\,0.9]$. A rule of thumb, but a defensible and widely accepted one (Crump, Hotz, Imbens and Mitnik, 2009).
  - **Report the change in estimand.** After trimming you describe a subpopulation, not the original one. Say how many observations went and how the treated sample shifted.
  - Consider retreating to the $ATT$, which only needs overlap in one direction.
- What *not* to do: report the estimate and stay silent. The regression will happily extrapolate into a region with no data and give you a standard error that reflects none of that.

## Inverse probability weighting

Rather than matching, **reweight**. Write $\hat e_i$ for $\hat e(X_i)$ and $\pi = \Pr(D{=}1)$.

\small
$$
\begin{aligned}
ATE \;&=\; E\!\left[Y_1 - Y_0\right] = E\!\left[\frac{D\,Y}{e(X)}\right] - E\!\left[\frac{(1-D)\,Y}{1 - e(X)}\right] \\
ATT \;&=\; E\!\left[Y_1 - Y_0 \middle| D = 1 \right] = \frac{1}{\pi}\,\left(E\!\left[DY\right] - E\!\left[\frac{e(X)(1-D)Y}{1 - e(X)}\right]\right) \\[1em]
\widehat{ATE}_{IPW} \;&=\; \frac{1}{n}\sum_{i=1}^{n}\left[\frac{D_i}{\hat e_i} - \frac{1 - D_i}{1 - \hat e_i}\right] Y_i \\
\widehat{ATT}_{IPW} \;&=\; \frac{1}{n_T}\sum_{i=1}^{n}\left[D_i - (1 - D_i)\,\frac{\hat e_i}{1 - \hat e_i}\right] Y_i
\end{aligned}
$$
\normalsize

- Both are **Horvitz--Thompson (1952)** forms.

## Why *that* weight for the ATE?

We prove $E\!\left[\frac{D\,Y}{e(X)}\right] = E[Y^1]$. The $E[Y^0]$ half is symmetric.

\scriptsize
$$
\begin{aligned}
E\!\left[ \frac{D \cdot Y}{e(X)} \right]
  &= E\!\left[ \frac{D \cdot Y^1}{e(X)} \right]
     && \text{switching equation (consistency): } D \cdot Y = D \cdot Y^1 \\
  &= E_X\!\left[ E\!\left( \frac{D \cdot Y^1}{e(X)} \,\Big|\, X \right) \right]
     && \text{law of total expectation} \\
  &= E_X\!\left[ \frac{1}{e(X)} \cdot E\big(D \cdot Y^1 \mid X\big) \right]
     && e(X) \text{ is fixed given } X \\
  &= E_X\!\left[ \frac{1}{e(X)} \cdot E(D \mid X) \cdot E\big(Y^1 \mid X\big) \right]
     && \text{unconfoundedness (exchangeability): } D \indep Y^1 \mid X \\
  &= E_X\!\left[ \frac{1}{e(X)} \cdot e(X) \cdot E\big(Y^1 \mid X\big) \right]
     && \text{by definition, } E(D \mid X) = e(X) \\
  &= E_X\big[ E(Y^1 \mid X) \big] \;=\; E[Y^1]
     && \text{cancel; law of total expectation}
\end{aligned}
$$

## Why *that* weight for the ATT?

\scriptsize
$E[Y^1 \mid D{=}1] = E[Y \mid D{=}1] = \frac{E[DY]}{E[D]}$.
$$
\begin{aligned}
E\big[Y^0 \mid D{=}1\big]
  &= E_X\big[\, E(Y^0 \mid X, D{=}1) \,\big|\, D{=}1 \,\big]
     && \text{law of total expectation, over } f(x \mid D{=}1) \\
  &= E_X\big[\, E(Y^0 \mid X, D{=}0) \,\big|\, D{=}1 \,\big]
     && \text{unconfoundedness --- swap the arm} \\
  &= E_X\!\left[\, E(Y^0 \mid X, D{=}0)\cdot\frac{e(X)}{\pi} \,\right]
     && \text{Bayes: } f(x \mid D{=}1) = e(x)f(x)/\pi \\
  &= E_X\!\left[\, \frac{E\big[(1-D)\,Y \mid X\big]}{1 - e(X)}\cdot\frac{e(X)}{\pi} \,\right]
     && E[(1{-}D)Y \mid X] = \big(1{-}e(X)\big)E[Y^0 \mid X, D{=}0] \\
  &= E\!\left[\, \frac{(1 - D)\,Y\,e(X)}{\big(1 - e(X)\big)\,\pi} \,\right]
     && \text{law of total expectation}
\end{aligned}
$$
\normalsize

Line 3: it converts an average over the *treated* covariate distribution into one over the *whole* population.

## The intuition: the ATE rebuilds the population

Two kinds of firm. Small firms are **rarely** treated, large firms **usually** are.

\begin{center}
\small
\begin{tabular}{lrrrrrr}
\toprule
Cell & $N$ & $e$ & treated & $\times\,1/e$ & controls & $\times\,1/(1-e)$ \\
\midrule
Small & $500$ & $0.2$ & $100$ & $\times 5 = \mathbf{500}$    & $400$ & $\times 1.25 = \mathbf{500}$ \\
Large & $500$ & $0.8$ & $400$ & $\times 1.25 = \mathbf{500}$ & $100$ & $\times 5 = \mathbf{500}$ \\
\midrule
Total & $1{,}000$ & & $500$ & $\mathbf{1{,}000}$ & $500$ & $\mathbf{1{,}000}$ \\
\bottomrule
\end{tabular}
\end{center}

- Only one small firm in five was treated, so that one must stand in for all five. Its weight is $1/0.2 = 5$.
- Every weighted entry comes back to $500$: each arm is blown back up to the **full population**. That is the *pseudo-population*, and in it $D \indep X$ by construction.
- Before weighting, small firms are $20\%$ of the treated sample and $80\%$ of the control sample. After weighting, $50\%$ of both --- the imbalance that caused the bias is gone.


## ... and the ATT reweights controls only

The $ATT$ does not want the population. It wants the **treated** --- and the treated are already a sample from that distribution, so they keep weight $1$. Only the controls move.

\begin{center}
\small
\begin{tabular}{lrrrr}
\toprule
Cell & controls & $\times\ e/(1-e)$ & $=$ & treated \\
\midrule
Small & $400$ & $0.25$ & $\mathbf{100}$ & $100$ \\
Large & $100$ & $4.00$ & $\mathbf{400}$ & $400$ \\
\bottomrule
\end{tabular}
\end{center}

- **Why the odds?** In a cell with score $e$ the sample splits $e : (1-e)$ between the arms. Multiplying the controls by $e/(1-e)$ makes their count match the treated count *exactly* --- last column.
- The $ATE$ weight asks "how many units like me are in the population?" The $ATT$ weight asks "how many *treated* units like me are there?"
- Among large firms, $100$ controls must cover $400$ treated, so each carries weight $4$. Push $e$ toward $1$ and that weight explodes: **this is where the $ATT$ fails**, and why it needs overlap in one direction only, $e(X) < 1$, while the $ATE$ needs $0 < e(X) < 1$.

## Does it work? 

\begin{center}
\small
\begin{tabular}{lrrrrrr}
\toprule
Cell & $e$ & $E[Y^0]$ & $E[Y^1]$ & effect & treated & controls \\
\midrule
Small & $0.2$ & $10$ & $12$ & $2$ & $100$ & $400$ \\
Large & $0.8$ & $20$ & $26$ & $6$ & $400$ & $100$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{-0.2em}
\small
$$
\begin{aligned}
\widehat{E[Y^1]} &= \tfrac{1}{1000}\!\left[\tfrac{100(12)}{0.2} + \tfrac{400(26)}{0.8}\right] = \tfrac{6000 + 13000}{1000} = 19 \\
\widehat{E[Y^0]} &= \tfrac{1}{1000}\!\left[\tfrac{400(10)}{0.8} + \tfrac{100(20)}{0.2}\right] = \tfrac{5000 + 10000}{1000} = 15 \\[0.2em]
\widehat{ATT}\; &= \underbrace{\tfrac{100(12) + 400(26)}{500}}_{\text{treated mean}\;=\;23.2} \;-\; \underbrace{\tfrac{400(10)(0.25) + 100(20)(4)}{500}}_{\text{controls}\,\times\,\text{odds}\;=\;18.0}
\end{aligned}
$$
\normalsize

- Truth: $ATE = 4$, $ATT = 5.2$. The naive $SDO$ is $23.2 - 12 = \mathbf{11.2}$.
- IPW returns $19 - 15 = \mathbf{4}$ and $23.2 - 18.0 = \mathbf{5.2}$. Both exactly right.
- $ATT > ATE$ because the effect is bigger for large firms, and large firms are where treatment concentrated. **Selection on gains --- the two estimands correctly disagree.**

## Why normalize the weights?

IPW is **sensitive to extreme propensity scores**: one observation with $\hat e$ near $0$ or $1$ gets an enormous weight and can dominate everything else.

Four treated firms with scores $\hat e = (0.5,\,0.5,\,0.5,\,0.25)$, and **every one of them has $Y = 10$**. With $n = 8$:

\vspace{-0.4em}
\small
$$
\widehat{E[Y^1]} = \frac{1}{8}\left[\frac{10}{0.5} + \frac{10}{0.5} + \frac{10}{0.5} + \frac{10}{0.25}\right] = \frac{20 + 20 + 20 + 40}{8} = \frac{100}{8} = 12.5
$$
\normalsize

- Every number in the data is $10$. Any sensible estimator must return $10$. This one returns $12.5$.
- The reason: the weights are $2, 2, 2, 4$ and they **sum to $10$, not to $8$**. Dividing by $n$ assumes they sum to $n$ --- true in population, where $E[D/e(X)] = 1$, but never exactly true in a sample with $\hat e$ estimated.
- Fix: **divide by the weights you actually got, not by $n$.** Here $100/10 = 10$.

## The normalized (Hájek) estimator

Hirano and Imbens (2001) normalize the weights **by their own sum within each group**, so the weights sum to one in each arm:

\vspace{-0.3em}
\small
$$
\widehat{ATE}_{\text{norm}} =
\underbrace{\left[\sum_{i=1}^N \frac{Y_i D_i}{\hat e_i}\right] \bigg/ \left[\sum_{i=1}^N \frac{D_i}{\hat e_i}\right]}_{\widehat{E[Y^1]}}
\;-\;
\underbrace{\left[\sum_{i=1}^N \frac{Y_i (1 - D_i)}{1 - \hat e_i}\right] \bigg/ \left[\sum_{i=1}^N \frac{1 - D_i}{1 - \hat e_i}\right]}_{\widehat{E[Y^0]}}
$$
\normalsize

- Each term is now a genuine **weighted average of observed outcomes**, so it must fall between the smallest and largest $Y$ in that arm. The unnormalized version carries no such guarantee --- which is exactly how it produced $12.5$.
- Same repair for the $ATT$: divide the reweighted control sum by $\sum_i (1-D_i)\frac{\hat e_i}{1-\hat e_i}$ rather than by $n_T$.
- Default in Stata's `teffects ipw` and R's `WeightIt`.


## Propensity score matching

- Match each treated unit to controls with a **similar score**, then difference. 
- **Caliper** matching: refuse any match farther than some tolerance (say $0.001$ on the score), *discarding* the treated unit if none qualifies. **Radius** matching: average all controls inside the tolerance.
- Two problems with both:
  - The tolerance is **chosen by the researcher** (Huntington-Klein et al. 2021).
  - Dropping treated units **changes the estimand** --- the $ATT$ summarised the *original* treated sample. 
- The score is estimated, so its sampling error enters the standard errors (Abadie--Imbens 2016).
- King & Nielsen, 2019: Two units can share an identical score $e(X_i) = e(X_j)$ despite having completely opposite covariate profiles $X_i \neq X_j$. Matching purely on the scalar score can actually increase local covariate imbalance compared to multivariate distance matching (e.g., Coarsened Exact Matching).

<!--
## King and Nielsen (2019): which experiment are you imitating?

*Why Propensity Scores Should Not Be Used for Matching*, **Political Analysis** 27, 435--454.

- Every matching method **recreates some experiment**. Which one decides how good it can get.
  - **Completely randomized**: covariates balanced *on average*. Draw again and the treated group is $52\%$ large firms, then $48\%$ --- balanced in expectation, never in the sample you hold.
  - **Fully blocked**: randomize *within* strata, so exactly $50\%$ large firms in both arms, every draw. Strictly better on imbalance, model dependence, variance and bias.
- **PSM targets the weaker one.** The balancing property promises equal covariate *distributions* within a score stratum --- complete randomization by construction, and no more.
- Mahalanobis and coarsened exact matching target the **blocked** experiment: they compare on $X$ itself. Cells A and B share $e = 0.2$ and are opposites on $X$, so PSM cannot tell an A--A pair from an A--B pair --- but **blocking would never have made that pair.**

## The PSM paradox, and what to do instead

- Matching prunes units to improve balance. For MDM and CEM, pruning **keeps helping**: you are always dropping the worst-matched pair on $X$.
- For PSM, once the remaining sample already looks completely randomized, the pairs you prune next are **arbitrary with respect to $X$** --- the score cannot distinguish them. You are now pruning at random.
- Randomly deleting units from a randomized experiment does not improve balance. It shrinks $n$ and makes realized imbalance *noisier*, so measured imbalance, model dependence and bias start going **up**.
- King and Nielsen: *"the more balanced the data, or the more balanced it becomes by \[trimming\] some of the observations through matching, the more likely propensity score matching will degrade inferences."*
- The trap is the standard workflow --- prune, check balance, adjust the caliper, prune again. That loop can walk you into the rising region while the diagnostics look like progress.

## Their recommendation, and its scope

- **Replace PSM with a method that blocks**: Mahalanobis distance matching, or coarsened exact matching. Match on $X$, not on a scalar summary of $X$.
- **Diagnose with the pruning curve.** Plot imbalance against the number of units pruned. Falling is fine; the moment it turns up, stop --- you are in the paradox.
- **The critique is narrow, and they say so.** It targets nearest-neighbour matching *on the score*. It leaves untouched:
  - inverse probability **weighting**,
  - **stratification** on the score (Rosenbaum--Rubin's original proposal),
  - **regression adjustment**, and the score as an ingredient in doubly robust estimators.
- So the propensity score survives this paper intact. What does not survive is one particular thing people do with it.
- Recommended viewing: Gary King, "16. Matching Methods" (83 min) --- `https://youtu.be/tvMyjDi4dyg`

## A surprise: estimate the score even when you know it

- Hahn (1998); Hirano, Imbens and Ridder (2003): weighting by the **estimated** $\hat e$ is *more efficient* than weighting by the **true** $e$ --- even when the truth is handed to you.
- Students find this backwards. The intuition is one you already accept.
- Run an experiment with a fair coin, $e = 0.5$. By chance the treated group comes out $55\%$ large firms and the control group $45\%$. The true score says "balanced" --- and it is right, *about the coin*. The estimated score reads $\hat e > 0.5$ for large firms and weights that chance imbalance away.
- **You do not need to correct the coin. You need to correct the sample you actually drew.** Only $\hat e$ knows what you drew.
- Same reason you put covariates in an RCT regression even though randomization already delivered identification.
- Scope: this is about *weighting*. For **matching** on $\hat e$, estimation moves the variance in either direction and must be accounted for (Abadie--Imbens 2016).
-->

## What the score does and does not buy

- It buys **dimension reduction**, not identification. If $X$ omits a confounder, $e(X)$ omits it too.
- Standard regression silently hides lack of common support through linear extrapolation, whereas plotting $\hat{e}(X)$ immediately exposes severe positivity violations.
- a high pseudo-$R^2$ in a treatment model means covariates perfectly predict treatment assignment ($e(X) \to 0$ or $1$). This breaks the positivity assumption and causes extreme IPW variance. Conversely, a low pseudo-$R^2$ indicates that treatment assignment is nearly independent of $X$ (resembling a randomized control trial).
- IPW shifts the modeling assumption from $E[Y \mid D, X]$ (outcome equation) to $P(D = 1 \mid X)$ (treatment assignment equation).

<!--

## You need not choose: doubly robust estimation

The choice just posed --- model the **outcome** or model the **treatment** --- is not forced. Fit *both*, and only one needs to be right.

\small
$$
\widehat{ATE}_{DR} = \frac{1}{n}\sum_i \Big[ \underbrace{\hat\mu^1(X_i) - \hat\mu^0(X_i)}_{\text{regression imputation}}
+ \underbrace{\frac{D_i\big(Y_i - \hat\mu^1(X_i)\big)}{\hat e(X_i)} - \frac{(1-D_i)\big(Y_i - \hat\mu^0(X_i)\big)}{1-\hat e(X_i)}}_{\text{IPW applied to the \emph{residuals}}} \Big]
$$
\normalsize

- Read it as regression imputation with an IPW **repair term** bolted on. The repair term is ordinary IPW --- but run on the residuals $Y - \hat\mu$, not on $Y$.
- If $\hat\mu^d$ is right, the residuals are mean-zero and the repair term **vanishes** --- whatever $\hat e$ is.
- If $\hat e$ is right, the repair term is a valid IPW estimate of exactly the error the first term made, and cancels it --- whatever $\hat\mu^d$ is.
- Consistent if **either** model is correct: *doubly robust*. Two chances instead of one.
- You have met the shape before: Lecture 3a, §8, Bias Correction, is the matching version of the same trick --- a crude estimator, plus a regression that prices out its leftover error.

## Doubly robust: why the repair term cancels

Take one cell with true score $e$, true means $y^1, y^0$, and any guessed outcome model $\hat\mu^1, \hat\mu^0$. Average the bracket over the units in that cell:

$$
\underbrace{\hat\mu^1 - \hat\mu^0}_{\text{guess}}
\;+\; \underbrace{\frac{e\,(y^1 - \hat\mu^1)}{e}}_{\text{treated share} \,\times\, \text{residual}}
\;-\; \underbrace{\frac{(1-e)(y^0 - \hat\mu^0)}{1-e}}_{\text{control share}\,\times\,\text{residual}}
\;=\; y^1 - y^0
$$

- The $e$'s cancel, the $\hat\mu$'s cancel, and out drops the true effect. **Whatever you guessed.**
- That is the whole of double robustness in one line: a *correct score* makes the share of treated units in the cell exactly undo the $1/\hat e$, so the residual is added back at full weight and the guess is erased.
- Symmetrically, if the guess is right then both residuals are zero and the second and third terms disappear, leaving $\hat\mu^1 - \hat\mu^0$ --- correct regardless of $\hat e$.
- **Only one of the two cancellations has to fire.**

## Doubly robust: the three cells, one model at a time

Truth: $ATE = 2$, naive $SDO = 7$. Now deliberately break one model at a time.

\begin{center}
\small
\begin{tabular}{llccc}
\toprule
Outcome model $\hat\mu$ & Score $\hat e$ & Regression only & IPW only & \textbf{Doubly robust} \\
\midrule
wrong: flat $10 / 16$ & correct $0.2/0.2/0.6$ & $6.0$ & $2.0$ & $\mathbf{2.0}$ \\
correct cell means    & wrong: $0.36$ for all & $2.0$ & $7.0$ & $\mathbf{2.0}$ \\
wrong: flat $10 / 16$ & wrong: $0.36$ for all & $6.0$ & $7.0$ & $7.0$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

- Row 1, cell A: $16 - 10 + \tfrac{0.2(14-16)}{0.2} - \tfrac{0.8(12-10)}{0.8} = 6 - 2 - 2 = 2$. Cells B and C also give exactly $2$.
- Row 2: the wrong score is a logit **with no covariates**, so $\hat e = 360/1000$ for everyone. Constant weights are no weights, and IPW hands back the naive $7$. The correct $\hat\mu$ rescues it anyway.
- Row 3 is the honest one. **Two chances is not a guarantee** --- get both wrong and you get neither. Here it collapses back to $7$.

## Doubly robust in practice

- **What it buys**: you no longer have to win the argument about which model is right. You have to win only one of the two.
- **What it does not buy**: identification. Both models are built from the same $X$. If $X$ omits a confounder, both are wrong in the same direction and row 3 is where you live.
- The score still needs **overlap**. Doubly robust estimators divide by $\hat e$ and $1-\hat e$ just as IPW does, so extreme scores hurt them the same way --- trim first.
- Cheap version you can run today: estimate $\hat e$, then run **weighted least squares** of $Y$ on $D$ and $X$ with IPW weights. Regression plus weighting is doubly robust for the $ATE$ when the outcome model is linear.
- In software: Stata `teffects aipw (y x) (d x)`; R `AIPW` or `DoubleML`. Report it beside your regression and your IPW estimate --- three numbers that agree is a real robustness check, and three that disagree tells you which model was load-bearing.

-->


<!-- ARCHIVE: simulation versions of the two hand-worked examples above.

## Checking the balancing property

Two covariates, $e(X) = \Lambda(1.0 X_1 + 0.8 X_2)$, $n = 20{,}000$. Treated-minus-control gap in each covariate:

```{r}
#| echo: true
set.seed(975)
n  <- 20000; X1 <- rnorm(n); X2 <- rnorm(n)
D  <- rbinom(n, 1, plogis(1.0 * X1 + 0.8 * X2))
eh <- predict(glm(D ~ X1 + X2, family = binomial), type = "response")
dec <- cut(eh, quantile(eh, 0:10 / 10), include.lowest = TRUE)
gap <- function(V) c(overall   = mean(V[D == 1]) - mean(V[D == 0]),
                     max_within = max(abs(tapply(seq_len(n), dec,
                       function(i) mean(V[i][D[i] == 1]) - mean(V[i][D[i] == 0])))))
round(rbind(X1 = gap(X1), X2 = gap(X2)), 3)
```

\vspace{0.2em}

- Overall gaps of $0.77$ and $0.60$ collapse to at most $0.14$ and $0.15$ **within deciles of the score**.
- We never conditioned on $X_1$ or $X_2$ --- only on one number --- and both balanced.



## Comparing the estimators

```{r}
#| echo: true
set.seed(975)
n  <- 20000
X  <- rnorm(n)
ps <- plogis(1.2 * X)                  # true propensity score
D  <- rbinom(n, 1, ps)
Y  <- 2 * D + 3 * X + rnorm(n)         # TRUE effect = 2

eh <- predict(glm(D ~ X, family = binomial), type = "response")
round(c(naive = mean(Y[D == 1]) - mean(Y[D == 0]),
        ols   = unname(coef(lm(Y ~ D + X))[2]),
        ipw   = mean(D * Y / eh) - mean((1 - D) * Y / (1 - eh))), 3)
```

\vspace{0.2em}

- Naive comparison: $4.85$, more than double the truth. Both adjusted estimators land near $2$.
- They agree **because the model is correctly specified**. That is the assumption doing the work, not the method.

-->

# 2. Regression

## Regression as a conditioning strategy

- Regression belongs in this chapter. It is not a rival to matching --- it is another way of executing the same substitution from Lecture 3a, §3, Identification Assumptions.
- Under unconfoundedness and **constant** treatment effects, $Y = \alpha + \beta D + X'\gamma + \varepsilon$ recovers $\beta$, because unconfoundedness delivers exactly the exogeneity condition OLS needs:
$$
(Y^1,Y^0)\indep D \mid X \;\;\Longrightarrow\;\; E[\varepsilon \mid D, X] = 0
$$
- Recall Lecture 1: the error term contains $Y^0$. Selection bias was $\mathrm{cov}(D, Y^0) \neq 0$. Conditioning on $X$ is the attempt to make that covariance zero *within strata*.
- What regression adds that matching lacks: it **interpolates and extrapolates** with functional form, so empty cells do not stop it.
- What it hides: it does that silently.

## Frisch--Waugh--Lovell

The coefficient on $D$ in a multiple regression equals the coefficient from a **bivariate** regression of residualized $Y$ on residualized $D$:

```{r}
#| echo: true
set.seed(975)
n <- 20000; X <- rnorm(n); D <- rbinom(n, 1, plogis(1.2 * X))
Y <- 2 * D + 3 * X + rnorm(n)
c(direct = unname(coef(lm(Y ~ D + X))[2]),
  fwl    = unname(coef(lm(resid(lm(Y ~ X)) ~ resid(lm(D ~ X))))[2]))
```

\vspace{0.2em}

- Identical to machine precision. "Controlling for $X$" *means* "using only the variation in $D$ that $X$ cannot explain."
- This is the cleanest statement of what a control variable does --- and it tells you where the identifying variation comes from.
- It also warns you: if $X$ explains nearly all of $D$, almost no variation remains, and the estimate rests on very few effective observations.

## What FWL implies about weights

- FWL says OLS uses the residual variation in $D$. Where is that variation largest? Where treatment is **least predictable**.
- Formally, with $X$ dummies, OLS weights each stratum by $N_j\, p_j(1-p_j)$ --- the **conditional variance of treatment** within it.
- So OLS gives most weight to strata where treatment is closest to a coin flip, and least where it is nearly deterministic.
- That is a defensible **efficiency** motive: those strata carry the most information per observation.
- It also means **OLS is not estimating the $ATT$** --- or the $ATE$ --- when effects are heterogeneous. It estimates a variance-weighted average with no policy interpretation.

## Regression adjustment (imputation)

- A more flexible use of regression that mirrors matching directly:
  1. Fit $\hat\mu^1(x)$ on the **treated** and $\hat\mu^0(x)$ on the **controls**, separately.
  2. For *every* unit, impute both potential outcomes and average the difference:
$$
\widehat{ATE} = \frac{1}{n}\sum_i \big( \hat\mu^1(X_i) - \hat\mu^0(X_i) \big)
$$
- Separate models allow the covariates to matter differently under treatment and control --- exactly what a single $\beta$ on $D$ forbids.
- This is the same imputation logic as matching, with a model in place of a neighbour.
- Combining it with weighting gives **doubly robust** estimators: consistent if *either* the outcome model or the propensity model is right. A genuinely useful hedge, and increasingly the default in applied work.

## Matching versus regression

- Both rely on the **same** identifying assumption: unconfoundedness. Neither creates exogenous variation.
- The difference is in *how they weight strata* when the effect is heterogeneous.
  - **Matching ($ATT$)** weights stratum $j$ by the number of **treated** units in it.
  - **OLS with $X$ dummies** weights stratum $j$ by $N_j\, p_j (1-p_j)$.
- They also differ in what they do about missing cells: matching refuses; regression extrapolates.
- So when the two disagree, ask two questions in order: *is it the weights, or is it the extrapolation?* Both are diagnosable.

## The same example under both

Two strata. $p_{\text{small}} = 0.5$, $p_{\text{large}} = 0.1$; effects $+2$ and $+3$; 100 firms per stratum.

\begin{center}
\begin{tabular}{lccc}
\toprule
& Weight on small & Weight on large & Estimate \\
\midrule
Matching ($ATT$): treated counts & $50/60 = 0.833$ & $10/60 = 0.167$ & $\mathbf{2.167}$ \\
OLS: $N_j p_j(1-p_j)$            & $25/34 = 0.735$ & $9/34 = 0.265$  & $\mathbf{2.265}$ \\
$ATE$: population shares         & $0.5$           & $0.5$           & $\mathbf{2.500}$ \\
\bottomrule
\end{tabular}
\end{center}

```{r}
#| echo: true
X <- c(rep(0, 100), rep(1, 100))
D <- c(rep(c(1, 0), each = 50), rep(c(1, 0), times = c(10, 90)))
Y <- ifelse(X == 0, 0, 5) + ifelse(X == 0, 2, 3) * D
round(unname(coef(lm(Y ~ D + factor(X)))[2]), 3)
```

## Reading the comparison

- The regression returns $2.265$ --- exactly the variance-weighted average, not the $ATT$ and not the $ATE$.
- The gap is small here. Angrist and Pischke argue it usually is, in practice.
- But "usually small" is not "zero", and the difference is entirely about **which firms your estimate describes**:
  - Matching describes the treated.
  - OLS describes wherever treatment was most unpredictable.
- **When it matters most**: when treatment probability varies a lot across strata *and* effects are heterogeneous. Both are typical in finance.
- This is the same variance-weighting issue that resurfaces, far more destructively, in staggered difference-in-differences.

# 3. What Matching Cannot Do

## Matching creates no new variation

- Matching **does not introduce exogenous variation**. It reorganizes and reweights the data you already have, using observables you already measured.
- Consequence: if your OLS regression was biased by endogeneity, **the matched estimate is biased too**.
- Contrast this with the designs still to come:
  - **IV** imports variation from outside the model.
  - **RDD** exploits assignment that is locally as good as random.
  - **DiD** differences out fixed unobservables.
- Matching belongs to a different category. It is a *conditioning strategy*, like regression --- not a research design.

## What it cannot fix

Matching is powerless against:

- **Unobserved confounders.** You cannot match on what you cannot measure.
- **Reverse causality.** Matching on $X$ says nothing about whether $Y$ caused $D$.
- **Measurement error** in treatment or outcome.
- **Bad controls.** Matching on a post-treatment variable is exactly as damaging as controlling for it --- Lecture 2 applies unchanged, and the collider $F$ of Lecture 3a, §2, Confounders, Covariates, and Colliders, is the warning.

\vspace{0.4em}

\begin{center}
\emph{You cannot match on what you cannot measure.}
\end{center}

## So what is it good for?

Given all that, matching still earns its place:

1. **A non-parametric benchmark.** It estimates the effect without assuming a functional form. If matching and OLS disagree sharply, your specification is doing more work than you thought.
2. **Common-support discipline.** Matching refuses to compare units with no counterpart. Regression silently extrapolates into regions with no data.
3. **Diagnostic value.** The propensity score distribution reveals overlap problems that a regression table hides completely.
4. **Robustness when the functional form is wrong.** Matching can outperform a misspecified regression in finite samples.

\vspace{0.3em}

- Use it as a *complement* to regression and a check on it --- not as an identification strategy.


# 4. Sensitivity Analysis

## The assumption you cannot test, revisited

- Two things this course has said are both true and both final: unconfoundedness is **untestable**, and no amount of matching creates variation that fixes an unobserved confounder.
- Neither is the end of the conversation. **Sensitivity analysis** changes the question:

\begin{center}
\emph{Not} ``is my design valid?'' --- unanswerable. \quad But: \\[0.25em]
\textbf{how large would a hidden confounder have to be to overturn my conclusion?}
\end{center}

- The output is not a $p$-value. It is a **threshold**, plus an argument about whether a confounder that large is plausible in your setting. The argument is yours; the threshold is arithmetic.
- Two families, matching the two routes of §1, Propensity Scores, and §2, Regression:
  - **Design-based**, for matched pairs: Rosenbaum bounds.
  - **Regression-based**, for the coefficient: Oster's $\delta$, and the Cinelli--Hazlett robustness value.
- Referees in finance now ask for one of these by name. Budget a table for it.

## Rosenbaum bounds: one parameter, $\Gamma$

- Matching made two units identical in $X$. Unconfoundedness then says they had **equal** probability of treatment. Suppose they did not.
- Let $\Gamma \geq 1$ bound how unequal. For any two units $i, j$ with $X_i = X_j$,
$$
\frac{1}{\Gamma} \;\leq\; \frac{\pi_i (1 - \pi_j)}{\pi_j (1 - \pi_i)} \;\leq\; \Gamma .
$$
- $\Gamma = 1$ **is** unconfoundedness. $\Gamma = 2$ says a hidden variable could leave one unit of the pair twice as likely, in odds, to be treated.
- For each $\Gamma$, compute the **worst case** over every hidden assignment consistent with that $\Gamma$, and report the resulting bound on the $p$-value of the randomization test.
- Then read off $\Gamma^{*}$: the point where the upper bound crosses your threshold.
- Note what $\Gamma^{*}$ is *not*. It is a statement about how much hidden bias your **design** can absorb. It never says the design is fine.

## Reading a Rosenbaum bound

$200$ matched pairs; one-sided Wilcoxon signed-rank test on the within-pair differences.

\begin{center}
\small
\begin{tabular}{lcccccccc}
\toprule
$\Gamma$ & $1.00$ & $1.10$ & $1.20$ & $1.30$ & $\mathbf{1.40}$ & $1.50$ & $1.75$ & $2.00$ \\
\midrule
upper bound on $p$ & $0.0001$ & $0.0008$ & $0.0040$ & $0.0149$ & $\mathbf{0.0415}$ & $0.0922$ & $0.3354$ & $0.6386$ \\
\bottomrule
\end{tabular}
\end{center}

- At $\Gamma = 1$ the result looks bulletproof: $p = 0.0001$. It is not. $\Gamma^{*} = 1.42$.
- A hidden variable that shifts the odds of treatment by $42\%$ between two otherwise identical units is enough to make this finding insignificant.
- This is the standard and slightly deflating experience. Strong $p$-values routinely coexist with small $\Gamma^{*}$, because the two answer different questions.
- Which leaves the question the table cannot answer: **is $1.42$ big?**

## Calibrating $\Gamma$: what does $1.42$ mean?

- $\Gamma$ is one number doing the work of two. Rosenbaum and Silber (2009) split it back apart: let $\Lambda$ be the odds ratio linking the hidden variable to **treatment**, and $\Delta$ the odds ratio linking it to the **outcome**. Then
$$
\Gamma \;=\; \frac{\Lambda\Delta + 1}{\Lambda + \Delta}.
$$
- Every $(\Lambda, \Delta)$ on that curve produces the same $\Gamma$. At $\Gamma^{*} = 1.42$:

\begin{center}
\small
\begin{tabular}{lccc}
\toprule
$\Lambda$ (odds of treatment) & $2.43$ & $3.00$ & $5.00$ \\
$\Delta$ (odds of a higher outcome) & $2.43$ & $2.07$ & $1.71$ \\
\bottomrule
\end{tabular}
\end{center}

- Now the question is one a *finance* audience can argue about: is there a plausible omitted characteristic that roughly **doubles** the odds of treatment and **doubles** the odds of a better outcome? Often the honest answer is yes.
- **Benchmark it.** Compute $\Lambda$ and $\Delta$ for your strongest *measured* confounder. If the killer confounder must be twice as strong as anything you measured, that is an argument. If half as strong will do, so is that.
- The bound is worst-case, so $\Gamma^{*}$ understates the true robustness.

## Oster's $\delta$: coefficient stability, done properly

- The informal version is in every seminar: *``the coefficient barely moved when I added controls, so unobservables cannot matter much.''*
- That intuition is not wrong, it is **unfinished**. It ignores how much the controls explained. A coefficient that is stable while $R^2$ barely moves has demonstrated nothing.
- Altonji, Elder and Taber (2005) turned it into an argument; Oster (2019) turned it into an estimator. Two inputs:
  - $\delta$ --- selection on unobservables **relative to** selection on observables. $\delta = 1$ means they are equally severe.
  - $R_{\max}$ --- the $R^2$ of the infeasible regression that included the unobservables.
- With $(\mathring\beta, \mathring R)$ from the short regression and $(\tilde\beta, \tilde R)$ from the controlled one,
$$
\beta^{*} \;\approx\; \tilde\beta \;-\; \delta\,\big(\mathring\beta - \tilde\beta\big)\,\frac{R_{\max} - \tilde R}{\tilde R - \mathring R}.
$$
- Two ways to report it: fix $\delta = 1$ and give $\beta^{*}$; or set $\beta^{*} = 0$ and solve for the $\delta$ that would be needed. Oster's calibration from randomized trials: $\delta = 1$, $R_{\max} = \min(1.3\,\tilde R,\, 1)$.

## Oster in practice: a warning

Simulated data, one unobserved confounder, **true effect $= 1.000$**:

\begin{center}
\small
\begin{tabular}{lcc}
\toprule
 & $\beta$ & $R^2$ \\
\midrule
short: $Y \sim D$ & $2.086$ & $0.452$ \\
controlled: $Y \sim D + X$ & $1.743$ & $0.484$ \\
infeasible: $Y \sim D + X + U$ & $1.014$ & $0.562$ \\
\bottomrule
\end{tabular}
\end{center}

- Oster's default rule sets $R_{\max} = 1.3 \times 0.484 = 0.629$, and returns $\beta^{*} = \mathbf{0.201}$ --- a correction that overshoots the truth by a mile.
- Feed it the *true* $R_{\max} = 0.562$ and it returns $\beta^{*} = \mathbf{0.918}$, almost exactly right. The method works. The **guess** was what failed.
- Meanwhile the $\delta$ test passes: the $\delta$ that would drive $\beta^{*}$ to zero is $1.13 > 1$.
- **The lesson.** Report $\delta$, which is a bounded comparison you can defend. Do not report $\beta^{*}$ as if it were a point estimate --- it is a linear function of a number you guessed. If you show it, show it across a range of $R_{\max}$.

## Cinelli and Hazlett (2020): the robustness value

- Same problem, third parameterization --- and the one that travels best, because it uses **partial $R^2$** rather than odds ratios or selection ratios.
- $RV_q$ is the minimum share of the residual variation that a confounder must explain in **both** $D$ and $Y$ in order to move the estimate by $100q\%$.
- In the regression above: $t = 47.1$ and $RV_{q=1} = \mathbf{0.48}$. Any confounder explaining less than $48\%$ of what is left in both $D$ and $Y$ cannot drive the effect to zero.
- Is that reassuring? Check it against the confounder we planted: $U$ explains $15\%$ of the residual variation in $Y$ and $34\%$ in $D$. Below the bar --- and indeed it does not kill the effect, it moves it from $1.743$ to $1.014$.
- The feature that sells it: **benchmark bounds.** Express $RV$ as a multiple of an observed covariate's partial $R^2$ --- ``the confounder would have to be $3\times$ as important as firm size'' is a sentence a referee can evaluate.
- Software: `sensemakr` (R and Stata), `psacalc` for Oster, `rbounds` and `mhbounds` for Rosenbaum.

# 5. The NSW Demonstration

## The 1970s empirical crisis

- The methods in this lecture were not adopted because they were elegant. They were adopted after a crisis of confidence in observational work.
- Job-training programs in the 1970s were evaluated observationally, and the estimates were all over the place --- depending, apparently, on the analyst.
- The **National Supported Work (NSW) Demonstration** was different: it *randomized* participation. So the experimental effect was known.
- That created something rare and valuable: a **benchmark**. You could now ask whether observational methods, applied to the same treated units, recover the experimental answer.
- Everything in this lecture gets graded against it.

## What LaLonde did

- LaLonde (1986) ran the decisive test. He took the NSW experiment, **threw away the experimental control group**, and replaced it with comparison groups built from survey data (CPS and PSID).
- He then applied the standard observational toolkit of the day and compared the results to the experimental benchmark of roughly **\$1,800** in annual earnings gains.
- The observational estimates ranged from a fraction of the benchmark to **the wrong sign** --- and which one you got depended heavily on how the comparison group was built.
- The paper was devastating and hugely influential. It is a major reason the profession moved toward IV, RDD, and DiD --- the other river.

## Dehejia and Wahba: not so fast

- **Dehejia and Wahba (1999, 2002)** revisited LaLonde's data with propensity score methods.
- Their finding: the observational estimates *did* recover the experimental benchmark reasonably well --- **once common support was imposed**.
- The key diagnosis: much of LaLonde's failure was an **overlap** problem, not only a confounding problem. The CPS and PSID comparison groups contained enormous numbers of people utterly unlike the NSW participants.
- Discard the non-comparable units and the methods perform far better.
- This is why the distinction of Lecture 3a, §3, Identification Assumptions, matters so much. Two different assumptions were failing, and only one of them is checkable.

## Smith and Todd: not so fast either

- **Smith and Todd (2005)** re-examined the Dehejia--Wahba result and found it fragile.
- The success was **sensitive to the subsample** analysed and to the **propensity score specification** chosen. Reasonable alternative choices undid it.
- Their charge is uncomfortable and worth stating plainly: the specification that worked may have been found *because* the experimental answer was already known.
- You cannot run that search in real research. In a genuine application there is no benchmark to check against --- that is the entire point of doing the study.

## The honest summary

- Selection on observables **can** work --- when overlap is genuine and the confounders are truly measured.
- Neither condition is verifiable in the setting where you need them:
  - Unconfoundedness is **untestable**, always.
  - Common support is **checkable**.
- So the practical lesson is asymmetric: do the check you can do, and argue explicitly for the assumption you cannot.
- **A useful exercise**: the NSW data are freely available. Apply every method in this lecture and compare them to the \$1,800 benchmark. Cunningham's chapter walks through it, and it is the fastest way to develop instincts about when these methods hold up.

# Wrapping Up

## Summary [Part 1]: the score and what it does

- **The propensity score** reduces $X$ to one number and defeats the curse of dimensionality for *conditioning*. It does nothing for **identification** --- unconfoundedness is still assumed, never delivered.
- **Judge the score by balance achieved, not by fit.** A high pseudo-$R^2$ with bad balance is a bad score; the reverse is fine. The whole stage is blind to $Y$, which is what makes the specification search legitimate.
- **Balance means the distribution, not the mean.** Standardized differences *and* variance ratios. Our `Volatility` covariate had $\text{SMD} = -0.010$ and a variance ratio of $2.27$.
- **IPW** reweights the sample into a pseudo-experiment: $1/\hat e$ for treated and $1/(1-\hat e)$ for controls builds the population; $\hat e/(1-\hat e)$ on controls alone rebuilds the *treated* distribution. Watch the tails, and normalize.
- **Propensity score matching** is the one use with a serious case against it: King and Nielsen (2019) show it targets complete randomization when you could have had blocking, so pruning can *increase* imbalance.

## Summary [Part 2]: the verdict

- **Regression is a conditioning strategy too**, not a rival family. FWL shows it uses the residual variation in $D$, which is why it weights strata by $N_j p_j(1-p_j)$: $2.167$ versus matching's $2.265$ in our example.
- **Matching creates no new exogenous variation.** If OLS was endogenous, matching is endogenous. It cannot fix an unobserved confounder, reverse causality, or a bad control.
- **Sensitivity analysis is the honest response** to an untestable assumption. Report $\Gamma^{*}$, or $\delta$, or a robustness value --- and calibrate it against a covariate you actually measured.
- **NSW**: these methods can work when overlap is real and the confounders are measured --- and you can only ever verify one of those two.
- Next lecture: designs that *create* variation instead of assuming it away.

## A checklist for your own paper

1. State the **estimand** before the estimator. $ATE$, $ATT$, or $ATU$ --- and say why the policy question needs that one.
2. Defend the **assignment mechanism** in institutional terms. Name the confounders and say how you measured them.
3. Justify every covariate as **pre-treatment**. Any post-treatment control needs an argument, not a footnote.
4. **Plot the overlap.** Report what you trimmed and how the estimand changed.
5. Report **balance**, not fit: standardized differences and variance ratios, before and after.
6. Report **matching and regression**. If they disagree, diagnose whether it is weights or extrapolation.
7. Run a **sensitivity analysis**, and calibrate the threshold against an observed covariate.
8. Run a **placebo** on a pre-treatment period.
9. Say plainly what would break the design --- and what unobservable would have to exist.

## References

\scriptsize

- Cunningham, S. *Causal Inference: The Remix*, Ch. 5.
- Rosenbaum, P. and D. Rubin (1983). The Central Role of the Propensity Score in Observational Studies for Causal Effects. *Biometrika* 70, 41--55.
- Hirano, K. and G. Imbens (2001). Estimation of Causal Effects Using Propensity Score Weighting. *Health Services and Outcomes Research Methodology* 2, 259--278.
- Imbens, G. (2004). Nonparametric Estimation of Average Treatment Effects under Exogeneity. *Review of Economics and Statistics* 86, 4--29.
- King, G. and R. Nielsen (2019). Why Propensity Scores Should Not Be Used for Matching. *Political Analysis* 27, 435--454.
- Ben-Michael, E., A. Feller and J. Rothstein (2021). The Augmented Synthetic Control Method. *JASA* 116, 1789--1803.
- LaLonde, R. (1986). Evaluating the Econometric Evaluations of Training Programs with Experimental Data. *American Economic Review* 76, 604--620.
- Dehejia, R. and S. Wahba (1999). Causal Effects in Nonexperimental Studies. *JASA* 94, 1053--1062.
- Smith, J. and P. Todd (2005). Does Matching Overcome LaLonde's Critique of Nonexperimental Estimators? *Journal of Econometrics* 125, 305--353.

## References: sensitivity analysis

\scriptsize

- Rosenbaum, P. (2002). *Observational Studies*, 2nd ed. Springer.
- Rosenbaum, P. and J. Silber (2009). Amplification of Sensitivity Analysis in Matched Observational Studies. *JASA* 104, 1398--1405.
- Altonji, J., T. Elder and C. Taber (2005). Selection on Observed and Unobserved Variables: Assessing the Effectiveness of Catholic Schools. *Journal of Political Economy* 113, 151--184.
- Oster, E. (2019). Unobservable Selection and Coefficient Stability: Theory and Evidence. *Journal of Business \& Economic Statistics* 37, 187--204.
- Cinelli, C. and C. Hazlett (2020). Making Sense of Sensitivity: Extending Omitted Variable Bias. *Journal of the Royal Statistical Society Series B* 82, 39--67.
